### 2010-06-30 trunk/cs848-db/Paper.tex
first commit

%\documentclass[conference]{IEEEtran}
\documentclass[12pt,a4paper]{article}
\usepackage{fullpage}

\usepackage[pdftex]{graphicx}

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{A Survey of Physical Design Techniques of Information Systems }

% author names and affiliations
% use a multiple column layout for up to three different
% affiliations
%\author{\IEEEauthorblockN{Karim Ali, Sarah Nadi}
%\IEEEauthorblockA{School of Computer Science\\
%University of Waterloo\\
%Email: {{karim, snadi}}@uwaterloo.ca}
%}

\author{Karim Ali and Sarah Nadi\\
\{karim, snadi\}@cs.uwaterloo.ca \\
David R. Cheriton School of Computer Science\\
University of Waterloo\\
}

% make the title area
\maketitle

\begin{abstract}
%\boldmath
The abstract goes here.
\end{abstract}

\section{Introduction}

Database users write queries and updates using a "user-friendly" language such as SQL without having to worry about how the underlying data is stored. Physical database design is concerned with the actual storage of data on disk. This means how the files are organized, and how they are accessed. Data is stored in the form of records in files which reside on a physical disk. In this sense, we try to find the most efficient way possible to store and access the data such that queries are answered and executed efficiently. This is reflected in designing proper indexing techniques, partitioning data for more efficient access, designing materialized views, data clustering, data compression, striping, mirroring and denormalization~\cite{lightstone2007physical}.

This survey presents the different physical design patterns that have been used in different types of information systems. The systems discussed in this survey are disk based relational databases, main memory relational databases, data warehouses, and XML databases. The aim of this survey is not to explain the details of every data structure used in physical design, but rather to compare how the different data structures have been used in these systems. This includes explaining how alterations or additions are made to data structures to fit the specific needs of every system. For a complete reference on the details of the data structures used in physical database design, please see (CITE HERE).

The rest of the survey is organized as follows. Section~\ref{SEC-ELEMENTS} first explains different elements of physical database design. For example, this includes explaining what indexes are and briefly mentioning what the different types of indexes used in information systems are. After explaining the different elements of physical database design, Section~\ref{SEC-DIFFSYS} discusses how each of the four systems uses these elements to improve its efficiency. Sections~\ref{SEC-DDRDB},~\ref{SEC-MMDB},~\ref{SEC-WAREHOUSES},~\ref{SEC-XML} discuss Disk Based Databases, Main Memory Databases, Data Warehouses, and XML Databases respectively. Throughout these sections, comparisons are made as to how the different elements have been modified to fit the needs of each system. Section~\ref{SEC-AUTO} then presents physical database design can be automated. Section~\ref{SEC-OPEN} mentions some of the open problems of physical database design, and suggests possible solutions. Finally Section~\ref{SEC-CONCL} summarizes and concludes this summary.

\section{Elements of Physical Database Design}
\label{SEC-ELEMENTS}

\subsection{Indexes}

The first physical database design decision to be considered is the choice of indexes to be implemented. The concept of indices has been around for a long time. Just like and index at the end of the book is provided to help find certain content faster, indices in an information system help find content in tables faster. The main idea of an index is to have each record in a table have a unique identifier (primary key) and then organize these identifiers in a certain way that allows for fast access of a specific record. Of course, indexing can be done on any column in the database and not necessarily the primary key. There are many type of indexes that have been used in information systems.

B-trees (short for Balanced Trees) are the most used database index structure.  The B-tree is essentially similar to the traditional Binary Search Tree, but instead of having one value per node, a B-tee can have many values per node~
cite{comer1979ubiquitous} as shown in Figure (PUT A FIGURE FOR A B TREE HERE). B-trees are suitable for relational databases since the cost of retrieval in a B-tree is at most proportional to: $log_{d}\frac{n+1}{2}$. The cost of insertion and deletion is at most proportional to $log_{d}n$ due to the possibility of progressing back up the tree to balance it after an insertion or a deletion. Although B-trees do well in retrieval, deletion and insertion, they do not perform well in sequential search.

Different variations of B-trees have been used. For example, $B^{+}-trees$ are now the main method of indexing in Disk Based Relational Databases (DRDB). These will be discussed in Section~\ref{SEC-DDRDB}. The T Tree has been proposed for main memory databases, and will be discussed in Section~\ref{SEC-MMDB}. Cache sensitive indexes were then proposed. A variation of T Trees, Cache Sensitive T-Trees (CST-Trees) was introduced, and is also discussed in Section~\ref{SEC-MMDB}. On the same note, Cache Conscious $B^{+}-trees$ and Prefetching $J^{+}-Trees$ are also discussed in Section~\ref{SEC-MMDB}.

Of course, there are other indexing techniques used such as hash tables, and bitmap indices. Bitmap indices are used in Data Warehouses, and will be discussed in details in Section~\ref{SEC-WAREHOUSES}.

\subsection{Materialized Views}

In a large database system, there are usually a few complex queries that require the joining of many large tables. If these queries are frequently run, then the database performance is likely to suffer, and users have to wait for a long time to get the results back. If we know these queries beforehand, it makes sense to simply store the results of the queries on disk instead of recomputing them each time. This is exactly what materialized views aim to do. By precalculating the results of a complex query, and storing them in a table on disk, the new table with the results is very likely to be much smaller than the original tables decreasing I/O access costs and thus increasing performance~\cite{lightstone2007physical}. Once a materialized view exists, the user can either explicitly query the materialized view, or enter the original query, and have the query optimizer rewrite the query to use the materialized views instead of the original tables. Query rewriting using materialized views is a very popular database research topic (CITE SOME REFERENCES), but is beyond the scope of this survey.

Despite the fact that materialized views provide a significant improvement in performance, we cannot simply create a materialized view for each common query. To begin with, materialized views consume disk storage which might be limited. The second, and main, problem with materialized views is their maintenance. Ensuring that the views have the most up to date data is tricky, and may outset the benefit of using materialized views if it is not properly designed. Finally, having several materialized views may increase the cost of searching for the appropriate view to use during the query optimization stage. Data warehouses heavily rely on materialized views since they have many queries with several joins. This will be discussed in details in Section~\ref{SEC-WAREHOUSES}.

Many query optimization techniques have relied on rewriting queries using views (E.g.~\cite{levy1995answering, gupta1995aggregate, goldstein2001optimizing, abiteboul1998complexity}).

\subsection{Partitioning}

Partitioning is an important aspect of physical design as it reduces table scanning time. There are two main categories of partitioning: horizontal partitioning and vertical partitioning. Horizontal partitioning divides the table into sets of table rows where each row or record still has all the attributes of the table. For example, dividing the data by date where all data dating less than year 2000 lies in one partition while all data dating more lies in another. On the other hand, vertical partitioning reduces the width of the dataset by storing some attributes in one table, and some in another.

The main advantage of any type of partitioning is to reduce the amount of time it takes to scan a table which in turn improves performance. For example, if a table has 1000 records, and a query is only using the first 20 records then having these 20 records in a partition containing 50 records will save time as opposed to examining all the 1000 records.

\subsection{Clustering}

\subsection{Data Compression}

\subsection{Other Techniques}

Striping, mirroring, denormalization.

\section{Physical Design of Different Information Systems}
\label{SEC-DIFFSYS}

Given the different physical design methods explained in the previous section, we now look at how these methods apply to different types of information systems. Mainly, how they apply to Disk-based relational database systems, main memory relational database systems, data warehouses, and XML databases.

\subsection{Disk Based Relational Database System (DRDB)}
\label{SEC-DDRDB}

Relational Databases were first introduced by Codd~\cite{codd1970relational}
In relational databases, users specify the actions they want to execute without having to worry about how these operations will actually be performed. This is where the role of physical design comes in. Physical design specifies the access paths used to get data from the tables. This includes which indices to use, and in which order to access the tables~\cite{finkelstein1988physical} and so on.

There are many variations of B-trees. In relational databases, the most popular index is the $B^{+}-tree$. The $B^{+}-tree$ is a variant of the B  tree which is becoming the main indexing method supported by many databases such as DB2, Oracle, and SQL Server~cite{lightstone2007physical}. The main difference between the $B^{+}-trees$ and the B-tree is that only leaf nodes contain data pointers in a $B^{+}-tree$. (COMPARE COST HERE). In Prefix $B^{+}-trees$, instead of using the actual key value of a record, a prefix from the key value is used to decrease the storage size needed which will in turn allow more records to be stored per node thus decreasing the height of the tree.

The B+tree is the main indexing method used in current relational databases~\cite{lightstone2007physical}. B+trees have been successful in relational databases since data is ultimately recorded in files. Since B+trees have a high fanout, they allow less I/O operations to access a specific data record which is stored in a specific file.

\subsection{Main Memory Relational Database System (MMDB)}
\label{SEC-MMDB}

A Main Memory Database (MMDB) is one where the data resides in the main memory of the system rather than on a disk~\cite{garcia1992main}. It should be noted that this is different from caching. Disk based database systems use the main memory to cache query results. However, MMDB's primary copy of the data resides in main memory. MMDB have many implementation challenges that have been addressed throughout the research community.

Garcia-Molina and Salem~\cite{garcia1992main} mention some of the challenges involved. One is concurrency control. In disk based databases, locks are tracked through a hash table, but the actual objects on disk do not contain lock information. On the other hand, in MMDB, lock status is part of the object itself since it is cheap to keep a number of bits for that. The next challenge is access methods. In disk-based databases, B-Trees are used to index the data for faster access. B-Trees have a short bushy nature to try to decrease the height of the tree to decrease the number of I/O accesses. Since I/O access is not a problem in MMDB, longer tree structures are used since its cheap to access main memory. Another point is that data values do not need to be stored in the index itself since they will be stored in main memory anyways so there are no performance gains from storing them in the index itself.

\subsubsection{Type of Indexes Used}

Two considerations are taken when designing index structures for MMDB. The first is that the data resides in main memory and not on disk, and so many of the considerations taken for I/O operations is no longer there. The second is having index structures cache conscious. Lots of research work has shown that cache performance is very critical in MMDBs~\cite{rao1999cache, boncz1999database}. This is because the big difference between processor speed and main memory access speed. Thus, a cache miss is still relatively expensive in main memory databases. Accordingly, we will see that many of the index data structures proposed for main memory databases focused on being cache conscious or cache sensitive.

\subsubsection*{$B^{+}-Trees$}
The $B^{+}-Tree$, explained in the previous section, is originally designed for disk based database systems. However, it has also been found useful for main memory databases (REF) as it is cache conscious.

\subsubsection*{T Trees}
Lehmen and Carey~\cite{lehman1986study} study the different features needed for indices of MMDB. When designing data structures for MMDB, we are mainly concerned with the efficient use of CPU cycles and memory. Accordingly, the main goal of an index structure for MMDB is to minimize the overall computation time while using the minimum memory possible. Since all the database data is in main memory, index structures do not need to store actual values like those in disk-oriented databases. Instead, a pointer to the data instead of the actual data can be stored. Lehmen and Carey introduce a new index structure, the T Tree, which combines the features from AVL Trees and B Trees that are suited to main memory. Figure~\ref{fig:ttree} shows the proposed structure of the T Tree. They run some experiments on data residing in main memory to compare the performance of the T Tree to existing structures. All structures were modified to contain pointers to data values rather than the data values themselves. The structures included were  AVL Trees, simple arrays, B Trees, Chained Bucket Hashing, Extendible Hashing, Linear Hashing and Modified Linear Hashing. Their experiments showed that for unordered data, Modified Linear Hashing gave the best performance, and for ordered data, T Trees gave the best performance for a mix of searches, inserts and deletes.

\begin{figure}[!t]
\centering
\includegraphics[width=9cm]{Ttree.png}
\caption{T Tree by Lehmen and Carey~\cite{lehman1986study}}
\label{fig:ttree}
\end{figure}

\subsubsection*{Cache Sensitive Search Trees (CSS-Trees)}
\cite{rao1999cache}.

\subsubsection*{Cache Sensitive T-Trees (CST-Trees)}

T Trees were the main index structure used for main memory databases for some time after their proposal. However, it was discovered that B+-trees outperform T-Trees on modern processors because of the growth of cache miss latency relative to processor speed~\cite{rao1999cache,lee2007cst}. This is because the height of the tree is high which makes the total number of memory accesses from the root to the leaf node higher. The other reason is that the node size is not aligned with the cache line size. The other problem with T-Trees is that there is a big waste of space in the cache because unnecessary data is brought into the cache. Additionally, record pointers stored in the tree take up a lot of space.

To resolve these issues, Lee et al.~
cite{lee2007cst} propose the Cache Sensitive T-Trees (CST-Trees). This is achieved in the following ways. First, a binary search tree whose node values are the maximum key of each node in the T Tre is constructed. That way, for a given value, the node containing this value could be quickly located through searching the binary tree. Second, the need to store pointers is eliminated by storing the tree in an array and locating the necessary nodes through index calculation. Finally, node sizes are aligned with cache line size such that there are no cache misses when accessing data in each binary search tree in the array.

\subsubsection*{Cache Sensitive $B^{+}-trees$ ($CSB^{+}-trees)$}

Rao and Ross~\cite{rao1999cache} show that $B^{+}-trees$ are more cache conscious than binary search trees and T-Trees which means they are more suited to main memory databases. However, traditional B+-Trees still had poor cache performance. Therefore, they propose Cache Sensitive B+-Trees ($CSB^{+}-Trees$)~\cite{rao2000making} as an index structure for main memory. To do so, they eliminated most of the child pointers and had more keys in each node to improve locality and reduce tree height~\cite{luan2009prefetching}. Since the number of cache misses in search operations is proportional to the height of the tree, $CSB^{+}-trees$ have fewer cache misses, and thus better performance.

\subsubsection*{$Prefetching B^{+}-Trees (pB^{+}-trees)$}

Chen et al.~\cite{chen2001improving} propose using prefetching to improve the performance of $B^{+}-trees$. Since current computer systems can prefetch different data simultaneously, the author take advantage of this. The node size of the $B^{+}-tree$ is made wider than the cache line size. This way, several cache misses can be overlapped at the same time thus improving performance.

\subsubsection*{$J^{+}-trees$ and $pJ^{+}-trees$}

\subsection{Data Warehouses}
\label{SEC-WAREHOUSES}

\subsection{XML Databases}
\label{SEC-XML}

XML (eXtensible Markup Language) is increasingly being used as a data exchange format. Accordingly, the ability to store and manage XML documents is needed. Salminen and Tompa~\cite{salminen2001requirements} define an XML database as ``a collection of XML documents
and their parts, maintained by a system having capabilities to
manage and control the collection itself and the information
represented by that collection.'' Salminena and Tompa also highlight some of the requirements of an XML database system. These requirements include dealing with namespaces, Internet resources, querying parts of an XML document, and transformations of XML documents.

There are two ways to store XML documents in a database. The first is to have the database internally translate the XML document into relational tables (XML-enabled databases). The second is to have data structures that can persistently store XML documents (native XML databases). eXist~\cite{meier2009exist} and TIMBER~\cite{jagadish2002timber} are two of the famous research efforts to design a native XML database.

\subsubsection{Index Structures}

Despite the special nature of XML documents, some of the index structures used in relational databases can still be utilized in XML databases. However, some adjustments may need to be made to fit the nature of the XML documents. For XML-enabled databases, XML documents are usually represented as relational tables, and then indexed similar to other tables. Pal et al.~\cite{pal2004indexing} describe how this is done in Microsoft SQL Server 2005. To start with, a new data type called `XML' was introduced. This data type could contain values of complete XML documents, or just fragments of XML data. First, the different nodes in the XML data are labelled with the ORDPATH~\cite{o2004ordpaths} mechanism. The XML data is then shredded into a relational table with five main columns:ORDPATH, TAG, NODE\_TYPE, VALUE, and PATH\_ID. This information is then stored in a $B^{+}-tree$ as the primary index of the XML type. Additionally, a secondary index can be created on any of the columns of the primary index to speed things up.

For native XML databases such as eXist~\cite{meier2009exist}, $B^{+}-trees$ were still used to index the XML documents. However, in contrast to XML-enabled databases, the XML data is not first stripped into a relational table. eXist uses a number schema to assign unique identifiers to each node in the XML document. These unique number identifiers allow the determination of any node's parent, sibling or possible child nodes. Four index files are created for XML data where all the indexes are based on $B^{+}-trees$. One index manages the collection hierarchy, one index collects nodes in a paged file and associates unique node identifiers to the actual nodes, one index indexes elements and attributes, and the last index keeps track of word occurrences and is used by the fulltext search extensions.

\subsubsection{Materialized Views}

Since XPath query processing is expensive, materialized views come in handy in XML databases. Balmin et al.~\cite{balmin2004framework} propose a framework for using XPath materialized views for XML query processing. (TALK ABOUT IT MORE)

\section{Real Databases (incorporate in previous section)}

Oracle Database uses B tree indexes and their subtypes such as Index-organized tables, reverse key indexes, descending indexes, B-tree cluster indexes. It also uses Bitmap and bitmap join indexes, function based indexes and domain based indexes.

MySQL mostly uses B trees, except for spatial data where R trees are used and hash indexes are used for memory tables.

\section{Automating Physical Database Design}
\label{SEC-AUTO}

\section{Open Problems in Physical Database Design}
\label{SEC-OPEN}

Indexing in the cloud~\cite{wang2010indexing}.

\section{Conclusions}
\label{SEC-CONCL}
% use section* for acknowledgement
\section*{Acknowledgment}

The authors would like to thank...

%\bibliographystyle{IEEEtran}

\bibliographystyle{plain}
\bibliography{references}

% that's all folks
\end{document}

### 2010-07-01 cs848-db/Paper.tex

--- 2010-06-30 trunk/cs848-db/Paper.tex
@@
 % can use linebreaks \\ within to get better formatting as desired
-\title{A Survey of Physical Design Techniques of Information Systems }
+\title{A Survey of Physical Design Techniques of Information Systems}
 
@@
 \centering
-\includegraphics[width=9cm]{Ttree.png}
+\includegraphics[width=9cm]{figs/Ttree.png}
 \caption{T Tree by Lehmen and Carey~\cite{lehman1986study}}
@@
 
-There are two ways to store XML documents in a database.
+There are two ways in which an XML database system can be designed.
 The first is to have the database internally translate the XML document into relational tables (XML-enabled databases).
 The second is to have data structures that can persistently store XML documents (native XML databases).
-eXist~\cite{meier2009exist} and TIMBER~\cite{jagadish2002timber} are two of the famous research efforts to design a native XML database.
+In this section, we will discuss how the database's physical design needs to be adapted in both cases.
+Other challenges such as expressing and executing queries and updates are faced in XML database design, but these are beyond the scope of this survey.
+
+%eXist~\cite{meier2009exist} and TIMBER~\cite{jagadish2002timber} are two of the famous research efforts to design a native XML database.
 
@@
 
-Despite the special nature of XML documents, some of the index structures used in relational databases can still be utilized in XML databases.
+Despite the special nature of XML documents, the same index structures used in relational databases can still be utilized in XML databases.
 However, some adjustments may need to be made to fit the nature of the XML documents.
@@
 
-For native XML databases such as eXist~\cite{meier2009exist}, $B^{+}-trees$ were still used to index the XML documents.
+For native XML databases such as eXist~\cite{meier2009exist} or TIMBER~\cite{jagadish2002timber}, $B^{+}-trees$ were still used to index the XML documents.
 However, in contrast to XML-enabled databases, the XML data is not first stripped into a relational table.
@@
 One index manages the collection hierarchy, one index collects nodes in a paged file and associates unique node identifiers to the actual nodes, one index indexes elements and attributes, and the last index keeps track of word occurrences and is used by the fulltext search extensions.
+
+TIMBER also uses a similar numbering schema.
+There are minor differences between the two numbering techniques, but the main idea is the same.
+The numbering is based on three values of a node: its start label, its end label, and its level (i.e.
+nested depth).
+On the other hand, Sedna~\cite{taranov2010sedna} uses a different reasoning to produce its numbering schema.
+Sedna assignes string identifiers to nodes such that they are lexically ordered according to the position of the node.
+Despite the differences in the ways the numbering is constructed, it seems that that no novel indexing structure is needed for XML databases.
+The problem is rather how to assign the nodes unique identifiers (the numbers or labels in the numbering schema) that can be used as identifiers in the index to reflect the XML document structure.
 

### 2010-07-03 cs848-db/Paper.tex
cleanup

@@
 
-\usepackage[pdftex]{graphicx}
+\usepackage{graphics}
+\usepackage{color}
+\usepackage[pdftex]{hyperref}
+\usepackage{graphicx}
 
@@
 \label{SEC-WAREHOUSES}
+\subsubsection*{Background}
+Data warehouses, in contrast, are targeted for decision support. Historical, summarized and consolidated data is more important than detailed, individual records.
+
+The workloads are query intensive with mostly ad hoc, complex queries that can access millions of records and perform a lot of scans, joins, and aggregates.
+Query throughput and response times are more important than transaction throughput.
+
+Operational databases are finely tuned to support known OLTP workloads.
+
+Understanding trends or making predictions requires historical data, whereas operational databases store only current data.
+
+Decision support usually requires consolidating data from many heterogeneous sources of varying quality, or use inconsistent representations, codes and formats.
+
+Supporting the multidimensional data models and operations typical of OLAP requires special data organization, access methods, and implementation methods, not generally provided by DBMSs targeted for OLTP.
+
+bases for decision support systems \cite{248616}
+OLAP vs OLTP
+ROLAP
+          o Data is stored in relational databases.
+          o Support extensions to SQL.
+          o Efficiently implement the multidimensional data model and operations.
+          o Ex: Microstrategy.
+MOLAP
+          o Directly store multidimensional data in special data structures.
+          o Implement the OLAP operations over these special data structures.
+          o Star Schema.
+          o Snowflake Schema.
+          o Fact Constellations.
+          o Ex: Essbase (Arbor), statistical DBMSes.
+star schema
+fact table
+
+Typical Operations
+    * Data Cleaning
+    * Load
+    * Refresh
+
+\subsubsection*{Challenges}
+- joins in data warehouses are usualy expensive since the fact table (the largest one) participates in every join \cite{628286}
+-
+
+\subsubsection*{Indexes}
+\paragraph{Tree-based Indexes}
+B-tree (B+ trees but commonly referred to as B-trees) \cite{253268}, R-tree \cite{602266}, K-D-B-tree \cite{582321}, BV-tree \cite{223796}
+
+\paragraph{Bitmapped Join Indexes}
+\cite{212001}
+
+\paragraph{Projection Indexes}
+Involves positional indexing where tuples are accessed based on their ordinal position \cite{628286}
+
+\paragraph{Bit-sliced Indexes}
+
+\subsubsection*{Partitioning}
+\paragraph{Vertical Partitioning}
+
+\paragraph{Horizontal Partitioning}
+
+\subsubsection*{Materlized Views}
+not very useful since most queries are ad-hoc \cite{653447}
+\cite{355309}
+
+\subsubsection*{Data Cubes}
 

### 2010-07-03 trunk/cs848-db/Paper.tex

@@
 Of course, there are other indexing techniques used such as hash tables, and bitmap indices.
-Bitmap indices are used in Data Warehouses, and will be discussed in details in Section~\ref{SEC-WAREHOUSES}.
+For example, Bitmap indices are commonly used in Data Warehouses, and will be discussed in details in Section~\ref{SEC-WAREHOUSES}.
 
@@
 \subsection{Clustering}
+When only records satisfying certain criteria are need from a table, it is often unnecessary to scan the whole table.
+For example, if only records that occrurred in January 2010 are needed, then querying all the table would be an extremely unnecessary overhead.
+This is where clustering comes in.
+Clustering relfects how the records are physically located together on disk.
+Records that are likely to be queried together will be stored physically together to avoid querying the whole table.
+For the previously mentioned example, the data can be clustered by month, such that all the records of January 2010 are physically in the same page, for example.
+In this case, only this page will need to be read.
+As people realized the potential of clustering, multidimensional clustering was introduced where data is clustered based on different criteria at the same time.
 

### 2010-07-05 cs848-db/Paper.tex
modified makefile to work with texmaker, just change the "Quick Build" user defined command to "make" and voila ;)
added some text to the data warehouses section, references

Note: when you compile the code, u will find in ur local repo a file named Paper.log (unversioned), use the context menu to ignore it :)

@@
 \subsubsection*{Background}
+
+OLAP vs OLTP
+
+OLTP \cite{sen2005comparison}
+
+bases for decision support systems \cite{248616}
 Data warehouses, in contrast, are targeted for decision support. Historical, summarized and consolidated data is more important than detailed, individual records.
@@
 
-bases for decision support systems \cite{248616}
-OLAP vs OLTP
-ROLAP
-          o Data is stored in relational databases.
-          o Support extensions to SQL.
-          o Efficiently implement the multidimensional data model and operations.
-          o Ex: Microstrategy.
-MOLAP
-          o Directly store multidimensional data in special data structures.
-          o Implement the OLAP operations over these special data structures.
-          o Star Schema.
-          o Snowflake Schema.
-          o Fact Constellations.
-          o Ex: Essbase (Arbor), statistical DBMSes.
 star schema
@@
 
-\subsubsection*{Challenges}
 - joins in data warehouses are usualy expensive since the fact table (the largest one) participates in every join \cite{628286}
--
 
@@
 \paragraph{Tree-based Indexes}
-B-tree (B+ trees but commonly referred to as B-trees) \cite{253268}, R-tree \cite{602266}, K-D-B-tree \cite{582321}, BV-tree \cite{223796}
+B-tree (B+ trees but commonly referred to as B-trees) \cite{253268}, R-tree \cite{602266, Cheung20011}, K-D-B-tree \cite{582321}, BV-tree \cite{223796}, UB-tree \cite{bayer1997universal}
 
@@
 \subsubsection*{Partitioning}
+\cite{thusoo2010data}
 \paragraph{Vertical Partitioning}
@@
 \subsubsection*{Data Cubes}
+\cite{Cheung20011}
+ROLAP
+          o Data is stored in relational databases.
+          o Support extensions to SQL.
+          o Efficiently implement the multidimensional data model and operations.
+          o Ex: Microstrategy.
+MOLAP
+          o Directly store multidimensional data in special data structures (arrays).
+          o Implement the OLAP operations over these special data structures.
+          o Star Schema.
+          o Snowflake Schema. \cite{kimball2009data}
+          o Fact Constellations.
+          o Ex: Essbase (Arbor), statistical DBMSes.
+
+hierachical data clusters \cite{karayannidis2008hierarchical}
 

### 2010-07-06 trunk/cs848-db/Paper.tex

@@
 \label{SEC-WAREHOUSES}
-\subsubsection*{Background}
+
+OLAP vs OLTP
+
+OLTP \cite{sen2005comparison}
+
+bases for decision support systems \cite{248616}
 Data warehouses, in contrast, are targeted for decision support. Historical, summarized and consolidated data is more important than detailed, individual records.
@@
 
-\subsubsection*{Challenges}
 - joins in data warehouses are usualy expensive since the fact table (the largest one) participates in every join \cite{628286}
--
 
-\subsubsection*{Indexes}
-\paragraph{Tree-based Indexes}
-B-tree (B+ trees but commonly referred to as B-trees) \cite{253268}, R-tree \cite{602266}, K-D-B-tree \cite{582321}, BV-tree \cite{223796}
+\subsubsection{Indexes}
+\subsubsection*{Tree-based Indexes}
+B-tree (B+ trees but commonly referred to as B-trees) \cite{253268}, R-tree \cite{602266, Cheung20011}, K-D-B-tree \cite{582321}, BV-tree \cite{223796}, UB-tree \cite{bayer1997universal}
 
-\paragraph{Bitmapped Join Indexes}
+\subsubsection*{Bitmapped Join Indexes}
 \cite{212001}
 
-\paragraph{Projection Indexes}
+\subsubsection*{Projection Indexes}
 Involves positional indexing where tuples are accessed based on their ordinal position \cite{628286}
 
-\paragraph{Bit-sliced Indexes}
+\subsection*{Bit-sliced Indexes}
 
-\subsubsection*{Partitioning}
-\paragraph{Vertical Partitioning}
+\subsubsection{Partitioning}
+\cite{thusoo2010data}
+\subsubsection*{Vertical Partitioning}
 
-\paragraph{Horizontal Partitioning}
+\subsubsection*{Horizontal Partitioning}
 
-\subsubsection*{Materlized Views}
+\subsubsection{Materlized Views}
 not very useful since most queries are ad-hoc \cite{653447}
@@
 
-\subsubsection*{Data Cubes}
+\subsubsection{Data Cubes}
+\cite{Cheung20011}
+ROLAP
+          o Data is stored in relational databases.
+          o Support extensions to SQL.
+          o Efficiently implement the multidimensional data model and operations.
+          o Ex: Microstrategy.
+MOLAP
+          o Directly store multidimensional data in special data structures (arrays).
+          o Implement the OLAP operations over these special data structures.
+          o Star Schema.
+          o Snowflake Schema. \cite{kimball2009data}
+          o Fact Constellations.
+          o Ex: Essbase (Arbor), statistical DBMSes.
+
+hierachical data clusters \cite{karayannidis2008hierarchical}
 
@@
 \label{SEC-CONCL}
+
 % use section* for acknowledgement

### 2010-07-07 cs848-db/Paper.tex

@@
 \label{SEC-WAREHOUSES}
-\subsubsection*{Background}
+A data warehouse is a subject-oriented, integrated, time-variant, and non-volatile collection of data and decision support technologies, aimed at enabling the knowledge worker (executive, manager, analyst) to make better and faster decisions \cite{560407, 248616}.
+More than half of IT executives named Data warehousing as the highest-priority post-millennium project for them \cite{sen2005comparison}.
+The value of data warehousing for an organization depends on the organization's need for reliable, consolidated, unique and integrated reporting and analysis of its data, at different levels of aggregation.
+
+Data warehousing has been shown useful in many industries: manufacturing (for order shipment and customer support), retail (for user profiling and inventory management), financial services (for claims analysis, risk analysis, credit card analysis, and fraud detection), transportation (for fleet management), telecommunications (for call analysis and fraud detection), utilities (for power usage analysis), and healthcare (for outcomes analysis) \cite{248616}.
+
+Data warehouses tend to be extremely large, in fact it is quite possible for a data warehouse to store tens of petabytes of data and load tens of terabytes of data everyday \cite{thusoo2010data}.
+Datta et al.
+\cite {628286} notes that the information in a data warehouse is usually multidimensional in nature, requiring the capability to view the data from a variety of perspectives.
+Aggregated and summarized data become more crucial than detailed records in such environment.
+Therefore, the workloads are query intensive with mostly ad hoc, complex queries, often requiring computationally expensive operations such as scans, joins, and aggregation.
+Performing such operations on large amounts of data, like in the case of data warehousing, complicates the situation further.
+Moreover, the results have to be delivered interactively to the business analyst using the system.
+
+The data warehouse supports on-line analytical processing (OLAP), the functional and performance requirements of which are quite different from those of the on-line transaction processing (OLTP) applications traditionally supported by the operational databases.
+Given the above challenges, the data warehouse is usually maintained separately from the organization’s operational databases \cite{248616}.
+Query throughput and response times are more important than transaction throughput.
 
@@
 
-bases for decision support systems \cite{248616}
 Data warehouses, in contrast, are targeted for decision support. Historical, summarized and consolidated data is more important than detailed, individual records.
@@
 
-\subsubsection*{Indexes}
+\subsubsection{Indexes}
 \paragraph{Tree-based Indexes}
@@
 
-\subsubsection*{Partitioning}
+\subsubsection{Partitioning}
 \cite{thusoo2010data}
@@
 
-\subsubsection*{Materlized Views}
+\subsubsection{Materlized Views}
 not very useful since most queries are ad-hoc \cite{653447}
@@
 
-\subsubsection*{Data Cubes}
+\subsubsection{Data Cubes}
 \cite{Cheung20011}

### 2010-07-07 trunk/cs848-db/Paper.tex

@@
 
-The first physical database design decision to be considered is the choice of indexes to be implemented.
-The concept of indices has been around for a long time.
+The first physical database design decision to be considered is the choice of indexes to be implemented. The concept of indices has been around for a long time.
 Just like and index at the end of the book is provided to help find certain content faster, indices in an information system help find content in tables faster.
-The main idea of an index is to have each record in a table have a unique identifier (primary key) and then organize these identifiers in a certain way that allows for fast access of a specific record.
-Of course, indexing can be done on any column in the database and not necessarily the primary key.
-There are many type of indexes that have been used in information systems.
-
-B-trees (short for Balanced Trees) are the most used database index structure.
-The B-tree is essentially similar to the traditional Binary Search Tree, but instead of having one value per node, a B-tee can have many values per node~
+The main idea of an index is to have each record in a table have a unique identifier (primary key) and then organize these identifiers in a certain way that
+allows for fast access of a specific record. Of course, indexing can be done on any column in the database and not necessarily the primary key. There are many
+type of indexes that have been used in information systems.
+B-trees (short for Balanced Trees) are the most used database index structure.  The B-tree is essentially similar to the traditional Binary Search Tree, but
+instead of having one value per node, a B-tee can have many values per node~
 cite{comer1979ubiquitous} as shown in Figure (PUT A FIGURE FOR A B TREE HERE).
@@
 
+\subsection{Clustering}
+Clustering in disk based systems is important because sequential I/O access is cheaper than random or dispersed I/O access. This is not the case with main
+memory, and so clustering is not needed in MMDB~\cite{garcia1992main, moldovan2008databases}. Therefore, components of one object may be spread across memory
+without impacting performance.
+
 \subsection{Data Warehouses}
 \label{SEC-WAREHOUSES}
+A data warehouse is a subject-oriented, integrated, time-variant, and non-volatile collection of data and decision support technologies, aimed at enabling the knowledge worker (executive, manager, analyst) to make better and faster decisions \cite{560407, 248616}.
+More than half of IT executives named Data warehousing as the highest-priority post-millennium project for them \cite{sen2005comparison}.
+The value of data warehousing for an organization depends on the organization's need for reliable, consolidated, unique and integrated reporting and analysis of its data, at different levels of aggregation.
+
+Data warehousing has been shown useful in many industries: manufacturing (for order shipment and customer support), retail (for user profiling and inventory management), financial services (for claims analysis, risk analysis, credit card analysis, and fraud detection), transportation (for fleet management), telecommunications (for call analysis and fraud detection), utilities (for power usage analysis), and healthcare (for outcomes analysis) \cite{248616}.
+
+Data warehouses tend to be extremely large, in fact it is quite possible for a data warehouse to store tens of petabytes of data and load tens of terabytes of data everyday \cite{thusoo2010data}.
+Datta et al.
+\cite {628286} notes that the information in a data warehouse is usually multidimensional in nature, requiring the capability to view the data from a variety of perspectives.
+Aggregated and summarized data become more crucial than detailed records in such environment.
+Therefore, the workloads are query intensive with mostly ad hoc, complex queries, often requiring computationally expensive operations such as scans, joins, and aggregation.
+Performing such operations on large amounts of data, like in the case of data warehousing, complicates the situation further.
+Moreover, the results have to be delivered interactively to the business analyst using the system.
+
+The data warehouse supports on-line analytical processing (OLAP), the functional and performance requirements of which are quite different from those of the on-line transaction processing (OLTP) applications traditionally supported by the operational databases.
+Given the above challenges, the data warehouse is usually maintained separately from the organization’s operational databases \cite{248616}.
+Query throughput and response times are more important than transaction throughput.
 
@@
 \subsubsection{Indexes}
-\subsubsection*{Tree-based Indexes}
+\paragraph{Tree-based Indexes}
 B-tree (B+ trees but commonly referred to as B-trees) \cite{253268}, R-tree \cite{602266, Cheung20011}, K-D-B-tree \cite{582321}, BV-tree \cite{223796}, UB-tree \cite{bayer1997universal}
 
-\subsubsection*{Bitmapped Join Indexes}
+\paragraph{Bitmapped Join Indexes}
 \cite{212001}
 
-\subsubsection*{Projection Indexes}
+\paragraph{Projection Indexes}
 Involves positional indexing where tuples are accessed based on their ordinal position \cite{628286}
 
-\subsection*{Bit-sliced Indexes}
+\paragraph{Bit-sliced Indexes}
 
@@
 \cite{thusoo2010data}
-\subsubsection*{Vertical Partitioning}
+\paragraph{Vertical Partitioning}
 
-\subsubsection*{Horizontal Partitioning}
+\paragraph{Horizontal Partitioning}
 

### 2010-07-08 trunk/cs848-db/Paper.tex

@@
 
-Partitioning is an important aspect of physical design as it reduces table scanning time.
-There are two main categories of partitioning: horizontal partitioning and vertical partitioning.
-Horizontal partitioning divides the table into sets of table rows where each row or record still has all the attributes of the table.
-For example, dividing the data by date where all data dating less than year 2000 lies in one partition while all data dating more lies in another.
-On the other hand, vertical partitioning reduces the width of the dataset by storing some attributes in one table, and some in another.
-
-The main advantage of any type of partitioning is to reduce the amount of time it takes to scan a table which in turn improves performance.
-For example, if a table has 1000 records, and a query is only using the first 20 records then having these 20 records in a partition containing 50 records will save time as opposed to examining all the 1000 records.
+Partitioning is an important aspect of physical design as it reduces table scanning time. There are two main categories of partitioning: horizontal partitioning
+and vertical partitioning. Horizontal partitioning divides the table into sets of table rows where each row or record still has all the attributes of the table.
+For example, dividing the data by date where all data dating less than year 2000 lies in one partition while all data dating more lies in another. On the other
+hand, vertical partitioning reduces the width of the dataset by storing some attributes in one table, and some in another.
+The main advantage of any type of partitioning is to reduce the amount of time it takes to scan a table which in turn improves performance. For example, if a
+table has 1000 records, and a query is only using the first 20 records then having these 20 records in a partition containing 50 records will save time as
+opposed to examining all the 1000 records. Further classification in partitioning is single vs group horizontal or vertical
+partitioning~\cite{gruenwald1990database}. Horizontal partitioning just divides the data into partitions where each group contains complete tuples. Group
+horizontal partitioning groups tuples that are more frequently used together. Single vertical partitioning divides a relation into groups each of which contains
+only one type of attribute. Group vertical partitioning divides the data vertically such that attributes that are commonly required together are physically
+stored together.
 
 \subsection{Clustering}
-When only records satisfying certain criteria are need from a table, it is often unnecessary to scan the whole table.
-For example, if only records that occrurred in January 2010 are needed, then querying all the table would be an extremely unnecessary overhead.
-This is where clustering comes in.
-Clustering relfects how the records are physically located together on disk.
-Records that are likely to be queried together will be stored physically together to avoid querying the whole table.
-For the previously mentioned example, the data can be clustered by month, such that all the records of January 2010 are physically in the same page, for example.
-In this case, only this page will need to be read.
-As people realized the potential of clustering, multidimensional clustering was introduced where data is clustered based on different criteria at the same time.
+When only records satisfying certain criteria are need from a table, it is often unnecessary to scan the whole table. For example, if only records that occurred
+in January 2010 are needed, then querying all the table would be an extremely unnecessary overhead. This is where clustering comes in. Clustering reflects how
+the records are physically located together on disk. Records that are likely to be queried together will be stored physically together to avoid querying the
+whole table. For the previously mentioned example, the data can be clustered by month, such that all the records of January 2010 are physically in the same
+page, for example. In this case, only this page will need to be read. As
+people realized the potential of clustering, multidimensional clustering was introduced where data is clustered based on different criteria at the same time.
 
@@
 
-Two considerations are taken when designing index structures for MMDB.
-The first is that the data resides in main memory and not on disk, and so many of the considerations taken for I/O operations is no longer there.
-The second is having index structures cache conscious.
-Lots of research work has shown that cache performance is very critical in MMDBs~\cite{rao1999cache, boncz1999database}.
-This is because the big difference between processor speed and main memory access speed.
-Thus, a cache miss is still relatively expensive in main memory databases.
-Accordingly, we will see that many of the index data structures proposed for main memory databases focused on being cache conscious or cache sensitive.
+Two considerations are taken when designing index structures for MMDB. The first is that the data resides in main memory and not on disk, and so many of the
+considerations taken for I/O operations is no longer there. The second is having index structures cache conscious. Lots of research work has shown that cache
+performance is very critical in MMDBs~\cite{rao1999cache, boncz1999database}. This is because the big difference between processor speed and main memory access
+speed. Thus, a cache miss is still relatively expensive in main memory databases. Accordingly, we will see that many of the index data structures proposed for
+main memory databases focused on being cache conscious or cache sensitive.
+
+cite~\cite{dewitt1984implementation}.
 
@@
 
+\subsection{Partitioning}
+
+Partitioning is used in disk based systems to divide the data into the physical pages in a way that will yield the best performance. This could be through
+vertical partitioning or horizontal partitioning, as previously explained. In main memory databases, we do not have the issue of expensive disk access which
+makes partitioning in memory unnecessary as it will not save any costs. However, partionining is needed in the secondary storage of the data on
+disk~\cite{gruenwald1990database}. Disk storage is still used as a backup of main memory databases. When a crash occurs, a reload of the database from disk or
+archive memory takes place. In order to avoid page faults caused by referencing data that is still being reloaded, the reloading process must be designed to be
+efficient in loading the important data such that the database users are not affected. Gruenwald and Eich~\cite{gruenwald1990database,gruenwald1990choosing}
+
 \subsection{Clustering}
@@
 
-- joins in data warehouses are usualy expensive since the fact table (the largest one) participates in every join \cite{628286}
+- joins in data warehouses are usually expensive since the fact table (the largest one) participates in every join \cite{628286}
 
@@
 
-\subsubsection{Materlized Views}
+\subsubsection{Materialized Views}
 not very useful since most queries are ad-hoc \cite{653447}
@@
 
-hierachical data clusters \cite{karayannidis2008hierarchical}
+hierarchical data clusters \cite{karayannidis2008hierarchical}
 

### 2010-07-08 trunk/cs848-db/Paper.tex

@@
 
-Database users write queries and updates using a "user-friendly" language such as SQL without having to worry about how the underlying data is stored.
-Physical database design is concerned with the actual storage of data on disk.
-This means how the files are organized, and how they are accessed.
-Data is stored in the form of records in files which reside on a physical disk.
-In this sense, we try to find the most efficient way possible to store and access the data such that queries are answered and executed efficiently.
-This is reflected in designing proper indexing techniques, partitioning data for more efficient access, designing materialized views, data clustering, data compression, striping, mirroring and denormalization~\cite{lightstone2007physical}.
+Database users write queries and updates using a "user-friendly" language such as SQL without having to worry about how the underlying data is stored. Physical
+database design is concerned with the actual storage of data on disk. This means how the files are organized, and how they are accessed. Data is stored in the
+form of records in files which reside on a physical disk. In this sense, we try to find the most efficient way possible to store and access the data such that
+queries are answered and executed efficiently. This is reflected in designing proper indexing techniques, partitioning data for more efficient access, designing
+materialized views, data clustering, data compression, striping, mirroring and denormalization~\cite{lightstone2007physical}.
 
-This survey presents the different physical design patterns that have been used in different types of information systems.
-The systems discussed in this survey are disk based relational databases, main memory relational databases, data warehouses, and XML databases.
-The aim of this survey is not to explain the details of every data structure used in physical design, but rather to compare how the different data structures have been used in these systems.
-This includes explaining how alterations or additions are made to data structures to fit the specific needs of every system.
-For a complete reference on the details of the data structures used in physical database design, please see (CITE HERE).
+This survey presents the different physical design patterns that have been used in different types of information systems. The systems discussed in this survey
+are disk based relational databases, main memory relational databases, data warehouses, and XML databases. The aim of this survey is not to explain the details
+of every data structure used in physical design, but rather to compare how the different data structures have been used in these systems. This includes
+explaining how alterations or additions are made to data structures to fit the specific needs of every system. For a complete reference on the details of the
+data structures used in physical database design, please see (CITE HERE).
 
-The rest of the survey is organized as follows.
-Section~\ref{SEC-ELEMENTS} first explains different elements of physical database design.
-For example, this includes explaining what indexes are and briefly mentioning what the different types of indexes used in information systems are.
-After explaining the different elements of physical database design, Section~\ref{SEC-DIFFSYS} discusses how each of the four systems uses these elements to improve its efficiency.
-Sections~\ref{SEC-DDRDB},~\ref{SEC-MMDB},~\ref{SEC-WAREHOUSES},~\ref{SEC-XML} discuss Disk Based Databases, Main Memory Databases, Data Warehouses, and XML Databases respectively.
-Throughout these sections, comparisons are made as to how the different elements have been modified to fit the needs of each system.
-Section~\ref{SEC-AUTO} then presents physical database design can be automated.
-Section~\ref{SEC-OPEN} mentions some of the open problems of physical database design, and suggests possible solutions.
-Finally Section~\ref{SEC-CONCL} summarizes and concludes this summary.
+The rest of the survey is organized as follows. Section~\ref{SEC-ELEMENTS} first explains different elements of physical database design. For example, this
+includes explaining what indexes are and briefly mentioning what the different types of indexes used in information systems are. After explaining the different
+elements of physical database design, Section~\ref{SEC-DIFFSYS} discusses how each of the four systems uses these elements to improve its efficiency.
+Sections~\ref{SEC-DDRDB},~\ref{SEC-MMDB},~\ref{SEC-WAREHOUSES},~\ref{SEC-XML} discuss Disk Based Databases, Main Memory Databases, Data Warehouses, and XML
+Databases respectively. Throughout these sections, comparisons are made as to how the different elements have been modified to fit the needs of each system.
+Section~\ref{SEC-AUTO} then presents physical database design can be automated. Section~\ref{SEC-OPEN} mentions some of the open problems of physical database
+design, and suggests possible solutions. Finally Section~\ref{SEC-CONCL} summarizes and concludes this summary.
 
@@
 type of indexes that have been used in information systems.
+
 B-trees (short for Balanced Trees) are the most used database index structure.  The B-tree is essentially similar to the traditional Binary Search Tree, but
-instead of having one value per node, a B-tee can have many values per node~
-cite{comer1979ubiquitous} as shown in Figure (PUT A FIGURE FOR A B TREE HERE).
-B-trees are suitable for relational databases since the cost of retrieval in a B-tree is at most proportional to: $log_{d}\frac{n+1}{2}$.
-The cost of insertion and deletion is at most proportional to $log_{d}n$ due to the possibility of progressing back up the tree to balance it after an insertion or a deletion.
+instead of having one value per node, a B-tee can have many values per node~\cite{comer1979ubiquitous} as shown in Figure (PUT A FIGURE FOR A B TREE HERE).
+B-trees are suitable for relational databases since the cost of retrieval in a B-tree is at most proportional to: $log_{d}\frac{n+1}{2}$. The cost of insertion
+and deletion is at most proportional to $log_{d}n$ due to the possibility of progressing back up the tree to balance it after an insertion or a deletion.
 Although B-trees do well in retrieval, deletion and insertion, they do not perform well in sequential search.
 
-Different variations of B-trees have been used.
-For example, $B^{+}-trees$ are now the main method of indexing in Disk Based Relational Databases (DRDB).
-These will be discussed in Section~\ref{SEC-DDRDB}.
-The T Tree has been proposed for main memory databases, and will be discussed in Section~\ref{SEC-MMDB}.
-Cache sensitive indexes were then proposed.
-A variation of T Trees, Cache Sensitive T-Trees (CST-Trees) was introduced, and is also discussed in Section~\ref{SEC-MMDB}.
-On the same note, Cache Conscious $B^{+}-trees$ and Prefetching $J^{+}-Trees$ are also discussed in Section~\ref{SEC-MMDB}.
+Different variations of B-trees have been used. For example, $B^{+}-trees$ are now the main method of indexing in Disk Based Relational Databases (DRDB). These
+will be discussed in Section~\ref{SEC-DDRDB}. The T Tree has been proposed for main memory databases, and will be discussed in Section~\ref{SEC-MMDB}. Cache
+sensitive indexes were then proposed. A variation of T Trees, Cache Sensitive T-Trees (CST-Trees) was introduced, and is also discussed in
+Section~\ref{SEC-MMDB}. On the same note, Cache Conscious $B^{+}-trees$ and Prefetching $J^{+}-Trees$ are also discussed in Section~\ref{SEC-MMDB}.
 
-Of course, there are other indexing techniques used such as hash tables, and bitmap indices.
-For example, Bitmap indices are commonly used in Data Warehouses, and will be discussed in details in Section~\ref{SEC-WAREHOUSES}.
+Of course, there are other indexing techniques used such as hash tables, and bitmap indices. For example, Bitmap indices are commonly used in Data Warehouses,
+and will be discussed in details in Section~\ref{SEC-WAREHOUSES}.
 
@@
 
-In a large database system, there are usually a few complex queries that require the joining of many large tables.
-If these queries are frequently run, then the database performance is likely to suffer, and users have to wait for a long time to get the results back.
-If we know these queries beforehand, it makes sense to simply store the results of the queries on disk instead of recomputing them each time.
-This is exactly what materialized views aim to do.
-By precalculating the results of a complex query, and storing them in a table on disk, the new table with the results is very likely to be much smaller than the original tables decreasing I/O access costs and thus increasing performance~\cite{lightstone2007physical}.
-Once a materialized view exists, the user can either explicitly query the materialized view, or enter the original query, and have the query optimizer rewrite the query to use the materialized views instead of the original tables.
-Query rewriting using materialized views is a very popular database research topic (CITE SOME REFERENCES), but is beyond the scope of this survey.
+In a large database system, there are usually a few complex queries that require the joining of many large tables. If these queries are frequently run, then the
+database performance is likely to suffer, and users have to wait for a long time to get the results back. If we know these queries beforehand, it makes sense to
+simply store the results of the queries on disk instead of recomputing them each time. This is exactly what materialized views aim to do. By precalculating the
+results of a complex query, and storing them in a table on disk, the new table with the results is very likely to be much smaller than the original tables
+decreasing I/O access costs and thus increasing performance~\cite{lightstone2007physical, chaudhuri1998overview}. Once a materialized view exists, the user can
+either explicitly query the materialized view, or enter the original query, and have the query optimizer rewrite the query to use the materialized views instead
+of the original tables. Many query optimization techniques have relied on rewriting queries using views (E.g.~\cite{levy1995answering, gupta1995aggregate,
+goldstein2001optimizing,
+abiteboul1998complexity}), but discussing the details of the rewriting is beyond the scope of this survey.
 
-Despite the fact that materialized views provide a significant improvement in performance, we cannot simply create a materialized view for each common query.
-To begin with, materialized views consume disk storage which might be limited.
-The second, and main, problem with materialized views is their maintenance.
-Ensuring that the views have the most up to date data is tricky, and may outset the benefit of using materialized views if it is not properly designed.
-Finally, having several materialized views may increase the cost of searching for the appropriate view to use during the query optimization stage.
-Data warehouses heavily rely on materialized views since they have many queries with several joins.
-This will be discussed in details in Section~\ref{SEC-WAREHOUSES}.
-
-Many query optimization techniques have relied on rewriting queries using views (E.g.~\cite{levy1995answering, gupta1995aggregate, goldstein2001optimizing, abiteboul1998complexity}).
+Despite the fact that materialized views provide a significant improvement in performance, we cannot simply create a materialized view for each common query. To
+begin with, materialized views consume disk storage which might be limited. The second, and main, problem with materialized views is their maintenance. Ensuring
+that the views have the most up to date data is tricky, and may outset the benefit of using materialized views if it is not properly designed. Finally, having
+several materialized views may increase the cost of searching for the appropriate view to use during the query optimization stage. Data warehouses heavily rely
+on materialized views since they have many queries with several joins. This will be discussed in details in Section~\ref{SEC-WAREHOUSES}.
 
@@
 Relational Databases were first introduced by Codd~\cite{codd1970relational}
-In relational databases, users specify the actions they want to execute without having to worry about how these operations will actually be performed.
-This is where the role of physical design comes in.
-Physical design specifies the access paths used to get data from the tables.
-This includes which indices to use, and in which order to access the tables~\cite{finkelstein1988physical} and so on.
+In relational databases, users specify the actions they want to execute without having to worry about how these operations will actually be performed. This is
+where the role of physical design comes in. Physical design specifies the access paths used to get data from the tables. This includes which indices to use, and
+in which order to access the tables~\cite{finkelstein1988physical} and so on.
 
-There are many variations of B-trees.
-In relational databases, the most popular index is the $B^{+}-tree$.
-The $B^{+}-tree$ is a variant of the B  tree which is becoming the main indexing method supported by many databases such as DB2, Oracle, and SQL Server~cite{lightstone2007physical}.
-The main difference between the $B^{+}-trees$ and the B-tree is that only leaf nodes contain data pointers in a $B^{+}-tree$.
-(COMPARE COST HERE).
-In Prefix $B^{+}-trees$, instead of using the actual key value of a record, a prefix from the key value is used to decrease the storage size needed which will in turn allow more records to be stored per node thus decreasing the height of the tree.
+There are many variations of B-trees. In relational databases, the most popular index is the $B^{+}-tree$. The $B^{+}-tree$ is a variant of the B  tree which is
+becoming the main indexing method supported by many databases such as DB2, Oracle, and SQL Server~cite{lightstone2007physical}. The main difference between the
+$B^{+}-trees$ and the B-tree is that only leaf nodes contain data pointers in a $B^{+}-tree$. (COMPARE COST HERE). In Prefix $B^{+}-trees$, instead of using the
+actual key value of a record, a prefix from the key value is used to decrease the storage size needed which will in turn allow more records to be stored per
+node thus decreasing the height of the tree.
 
-The B+tree is the main indexing method used in current relational databases~\cite{lightstone2007physical}.
-B+trees have been successful in relational databases since data is ultimately recorded in files.
-Since B+trees have a high fanout, they allow less I/O operations to access a specific data record which is stored in a specific file.
+The B+tree is the main indexing method used in current relational databases~\cite{lightstone2007physical}. B+trees have been successful in relational databases
+since data is ultimately recorded in files. Since B+trees have a high fanout, they allow less I/O operations to access a specific data record which is stored in
+a specific file.
 
@@
 
-A Main Memory Database (MMDB) is one where the data resides in the main memory of the system rather than on a disk~\cite{garcia1992main}.
-It should be noted that this is different from caching.
-Disk based database systems use the main memory to cache query results.
-However, MMDB's primary copy of the data resides in main memory.
-MMDB have many implementation challenges that have been addressed throughout the research community.
+A Main Memory Database (MMDB) is one where the data resides in the main memory of the system rather than on a disk~\cite{garcia1992main}. It should be noted
+that this is different from caching. Disk based database systems use the main memory to cache query results. However, MMDB's primary copy of the data resides in
+main memory. MMDB have many implementation challenges that have been addressed throughout the research community.
 
-Garcia-Molina and Salem~\cite{garcia1992main} mention some of the challenges involved.
-One is concurrency control.
-In disk based databases, locks are tracked through a hash table, but the actual objects on disk do not contain lock information.
-On the other hand, in MMDB, lock status is part of the object itself since it is cheap to keep a number of bits for that.
-The next challenge is access methods.
-In disk-based databases, B-Trees are used to index the data for faster access.
-B-Trees have a short bushy nature to try to decrease the height of the tree to decrease the number of I/O accesses.
-Since I/O access is not a problem in MMDB, longer tree structures are used since its cheap to access main memory.
-Another point is that data values do not need to be stored in the index itself since they will be stored in main memory anyways so there are no performance gains from storing them in the index itself.
+Garcia-Molina and Salem~\cite{garcia1992main} mention some of the challenges involved. One is concurrency control. In disk based databases, locks are tracked
+through a hash table, but the actual objects on disk do not contain lock information. On the other hand, in MMDB, lock status is part of the object itself since
+it is cheap to keep a number of bits for that. The next challenge is access methods. In disk-based databases, B-Trees are used to index the data for faster
+access. B-Trees have a short bushy nature to try to decrease the height of the tree to decrease the number of I/O accesses. Since I/O access is not a problem in
+MMDB, longer tree structures are used since its cheap to access main memory. Another point is that data values do not need to be stored in the index itself
+since they will be stored in main memory anyways so there are no performance gains from storing them in the index itself.
 
@@
 Two considerations are taken when designing index structures for MMDB. The first is that the data resides in main memory and not on disk, and so many of the
-considerations taken for I/O operations is no longer there. The second is having index structures cache conscious. Lots of research work has shown that cache
-performance is very critical in MMDBs~\cite{rao1999cache, boncz1999database}. This is because the big difference between processor speed and main memory access
-speed. Thus, a cache miss is still relatively expensive in main memory databases. Accordingly, we will see that many of the index data structures proposed for
-main memory databases focused on being cache conscious or cache sensitive.
+considerations taken for I/O operations is no longer there. Instead of focusing on disk access and disk storage, the data structures for main memory
+databases should focus on the efficient use of CPU cycles and memory~\cite{lehman1986study}. The second is having index structures cache conscious. Lots of
+research work has shown that cache performance is very critical in MMDBs~\cite{boncz1999database,rao1999cache}. This is because the big difference between
+processor speed and main memory access speed. Thus, a cache miss is still relatively expensive in main memory databases. Accordingly, we will see that many of
+the index data structures proposed for main memory databases focused on being cache conscious or cache sensitive. In summary, a main memory index should be
+able to reduce overall computation time while using as little memory as possible~\cite{lehman1986study}. It should also be noted that unlike disk based systems
+where it is advantageous to store actual attribute values in the index, main memory systems do not require that. Instead, pointers to the actual attribute
+values can be used. This, in turn, will reduce the size of the index which is preferable in this case.
 
@@
 
-\subsubsection*{$B^{+}-Trees$}
-The $B^{+}-Tree$, explained in the previous section, is originally designed for disk based database systems.
-However, it has also been found useful for main memory databases (REF) as it is cache conscious.
+\subsubsection*{B Trees}
+Although B Trees, explained in the previous section, are originally designed for disk based database systems, they were found to also be useful in
+MMDB~\cite{lehman1986study}. In disk based systems, the B+ Tree is more used than the regular B Tree. However, in main memory databases, the B Tree would be
+more suitable since there is no gain from keeping all the data in the leaves. In a main memory system, this would only waste space without improving
+performance.
 
 \subsubsection*{T Trees}
-Lehmen and Carey~\cite{lehman1986study} study the different features needed for indices of MMDB.
-When designing data structures for MMDB, we are mainly concerned with the efficient use of CPU cycles and memory.
-Accordingly, the main goal of an index structure for MMDB is to minimize the overall computation time while using the minimum memory possible.
-Since all the database data is in main memory, index structures do not need to store actual values like those in disk-oriented databases.
-Instead, a pointer to the data instead of the actual data can be stored.
 Lehmen and Carey introduce a new index structure, the T Tree, which combines the features from AVL Trees and B Trees that are suited to main memory.
-Figure~\ref{fig:ttree} shows the proposed structure of the T Tree.
-They run some experiments on data residing in main memory to compare the performance of the T Tree to existing structures.
-All structures were modified to contain pointers to data values rather than the data values themselves.
-The structures included were  AVL Trees, simple arrays, B Trees, Chained Bucket Hashing, Extendible Hashing, Linear Hashing and Modified Linear Hashing.
-Their experiments showed that for unordered data, Modified Linear Hashing gave the best performance, and for ordered data, T Trees gave the best performance for a mix of searches, inserts and deletes.
+Figure~\ref{fig:ttree} shows the proposed structure of the T Tree. They run some experiments on data residing in main memory to compare the performance of the T
+Tree to existing structures. All structures were modified to contain pointers to data values rather than the data values themselves. The structures
+compared included  AVL Trees, simple arrays, B Trees, Chained Bucket Hashing, Extendible Hashing, Linear Hashing and Modified Linear Hashing. Their
+experiments showed that for unordered data, Modified Linear Hashing gave the best performance, and for ordered data, T Trees gave the best performance for a mix
+of searches, inserts and deletes. This is mainly because a T node contains many elements which results in good update and storage characteristics.
 
@@
 
-T Trees were the main index structure used for main memory databases for some time after their proposal.
-However, it was discovered that B+-trees outperform T-Trees on modern processors because of the growth of cache miss latency relative to processor speed~\cite{rao1999cache,lee2007cst}.
-This is because the height of the tree is high which makes the total number of memory accesses from the root to the leaf node higher.
-The other reason is that the node size is not aligned with the cache line size.
-The other problem with T-Trees is that there is a big waste of space in the cache because unnecessary data is brought into the cache.
+T Trees were the main index structure used for main memory databases for some time after their proposal. However, it was discovered that B+-trees outperform
+T-Trees on modern processors because of the growth of cache miss latency relative to processor speed~\cite{rao1999cache,lee2007cst}. This is because the height
+of the tree is high which makes the total number of memory accesses from the root to the leaf node higher. The other reason is that the node size is not aligned
+with the cache line size. The other problem with T-Trees is that there is a big waste of space in the cache because unnecessary data is brought into the cache.
 Additionally, record pointers stored in the tree take up a lot of space.
 
-To resolve these issues, Lee et al.~
-cite{lee2007cst} propose the Cache Sensitive T-Trees (CST-Trees).
-This is achieved in the following ways.
-First, a binary search tree whose node values are the maximum key of each node in the T Tre is constructed.
-That way, for a given value, the node containing this value could be quickly located through searching the binary tree.
-Second, the need to store pointers is eliminated by storing the tree in an array and locating the necessary nodes through index calculation.
-Finally, node sizes are aligned with cache line size such that there are no cache misses when accessing data in each binary search tree in the array.
+To resolve these issues, Lee et al.~\cite{lee2007cst} propose the Cache Sensitive T-Trees (CST-Trees). This is achieved in the following ways. First, a binary
+search tree whose node values are the maximum key of each node in the T Tre is constructed. That way, for a given value, the node containing this value could be
+quickly located through searching the binary tree. Second, the need to store pointers is eliminated by storing the tree in an array and locating the necessary
+nodes through index calculation. Finally, node sizes are aligned with cache line size such that there are no cache misses when accessing data in each binary
+search tree in the array.
 
-\subsubsection*{Cache Sensitive $B^{+}-trees$ ($CSB^{+}-trees)$}
+\subsubsection*{Cache Sensitive B+ trees (CSB+ Trees)}
 
-Rao and Ross~\cite{rao1999cache} show that $B^{+}-trees$ are more cache conscious than binary search trees and T-Trees which means they are more suited to main memory databases.
-However, traditional B+-Trees still had poor cache performance.
-Therefore, they propose Cache Sensitive B+-Trees ($CSB^{+}-Trees$)~\cite{rao2000making} as an index structure for main memory.
-To do so, they eliminated most of the child pointers and had more keys in each node to improve locality and reduce tree height~\cite{luan2009prefetching}.
-Since the number of cache misses in search operations is proportional to the height of the tree, $CSB^{+}-trees$ have fewer cache misses, and thus better performance.
+Rao and Ross~\cite{rao1999cache} show that $B^{+}-trees$ are more cache conscious than binary search trees and T-Trees which means they are more suited to main
+memory databases. However, traditional B+-Trees still had poor cache performance. Therefore, they propose Cache Sensitive B+-Trees
+($CSB^{+}-Trees$)~\cite{rao2000making} as an index structure for main memory. To do so, they eliminated most of the child pointers and had more keys in each
+node to improve locality and reduce tree height~\cite{luan2009prefetching}. Since the number of cache misses in search operations is proportional to the height
+of the tree, $CSB^{+}-trees$ have fewer cache misses, and thus better performance.
 
@@
 
-Chen et al.~\cite{chen2001improving} propose using prefetching to improve the performance of $B^{+}-trees$.
-Since current computer systems can prefetch different data simultaneously, the author take advantage of this.
-The node size of the $B^{+}-tree$ is made wider than the cache line size.
-This way, several cache misses can be overlapped at the same time thus improving performance.
+Chen et al.~\cite{chen2001improving} propose using prefetching to improve the performance of $B^{+}-trees$. Since current computer systems can prefetch
+different data simultaneously, the author take advantage of this. The node size of the $B^{+}-tree$ is made wider than the cache line size. This way, several
+cache misses can be overlapped at the same time thus improving performance.
 
@@
 \label{SEC-WAREHOUSES}
-A data warehouse is a subject-oriented, integrated, time-variant, and non-volatile collection of data and decision support technologies, aimed at enabling the knowledge worker (executive, manager, analyst) to make better and faster decisions \cite{560407, 248616}.
-More than half of IT executives named Data warehousing as the highest-priority post-millennium project for them \cite{sen2005comparison}.
-The value of data warehousing for an organization depends on the organization's need for reliable, consolidated, unique and integrated reporting and analysis of its data, at different levels of aggregation.
+A data warehouse is a subject-oriented, integrated, time-variant, and non-volatile collection of data and decision support technologies, aimed at enabling the
+knowledge worker (executive, manager, analyst) to make better and faster decisions \cite{560407, 248616}. More than half of IT executives named Data warehousing
+as the highest-priority post-millennium project for them \cite{sen2005comparison}. The value of data warehousing for an organization depends on the
+organization's need for reliable, consolidated, unique and integrated reporting and analysis of its data, at different levels of aggregation.
 
-Data warehousing has been shown useful in many industries: manufacturing (for order shipment and customer support), retail (for user profiling and inventory management), financial services (for claims analysis, risk analysis, credit card analysis, and fraud detection), transportation (for fleet management), telecommunications (for call analysis and fraud detection), utilities (for power usage analysis), and healthcare (for outcomes analysis) \cite{248616}.
+Data warehousing has been shown useful in many industries: manufacturing (for order shipment and customer support), retail (for user profiling and inventory
+management), financial services (for claims analysis, risk analysis, credit card analysis, and fraud detection), transportation (for fleet management),
+telecommunications (for call analysis and fraud detection), utilities (for power usage analysis), and healthcare (for outcomes analysis) \cite{248616}.
 
-Data warehouses tend to be extremely large, in fact it is quite possible for a data warehouse to store tens of petabytes of data and load tens of terabytes of data everyday \cite{thusoo2010data}.
-Datta et al.
-\cite {628286} notes that the information in a data warehouse is usually multidimensional in nature, requiring the capability to view the data from a variety of perspectives.
-Aggregated and summarized data become more crucial than detailed records in such environment.
-Therefore, the workloads are query intensive with mostly ad hoc, complex queries, often requiring computationally expensive operations such as scans, joins, and aggregation.
-Performing such operations on large amounts of data, like in the case of data warehousing, complicates the situation further.
-Moreover, the results have to be delivered interactively to the business analyst using the system.
+Data warehouses tend to be extremely large, in fact it is quite possible for a data warehouse to store tens of petabytes of data and load tens of terabytes of
+data everyday \cite{thusoo2010data}. Datta et al. \cite {628286} notes that the information in a data warehouse is usually multidimensional in nature, requiring
+the capability to view the data from a variety of perspectives. Aggregated and summarized data become more crucial than detailed records in such environment.
+Therefore, the workloads are query intensive with mostly ad hoc, complex queries, often requiring computationally expensive operations such as scans, joins, and
+aggregation. Performing such operations on large amounts of data, like in the case of data warehousing, complicates the situation further. Moreover, the results
+have to be delivered interactively to the business analyst using the system.
 
-The data warehouse supports on-line analytical processing (OLAP), the functional and performance requirements of which are quite different from those of the on-line transaction processing (OLTP) applications traditionally supported by the operational databases.
-Given the above challenges, the data warehouse is usually maintained separately from the organization’s operational databases \cite{248616}.
-Query throughput and response times are more important than transaction throughput.
+The data warehouse supports on-line analytical processing (OLAP), the functional and performance requirements of which are quite different from those of the
+on-line transaction processing (OLTP) applications traditionally supported by the operational databases. Given the above challenges, the data warehouse is
+usually maintained separately from the organization’s operational databases \cite{248616}. Query throughput and response times are more important than
+transaction throughput.
 
@@
 
-Supporting the multidimensional data models and operations typical of OLAP requires special data organization, access methods, and implementation methods, not generally provided by DBMSs targeted for OLTP.
+Supporting the multidimensional data models and operations typical of OLAP requires special data organization, access methods, and implementation methods, not
+generally provided by DBMSs targeted for OLTP.
 
@@
 \paragraph{Tree-based Indexes}
-B-tree (B+ trees but commonly referred to as B-trees) \cite{253268}, R-tree \cite{602266, Cheung20011}, K-D-B-tree \cite{582321}, BV-tree \cite{223796}, UB-tree \cite{bayer1997universal}
+B-tree (B+ trees but commonly referred to as B-trees) \cite{253268}, R-tree \cite{602266, Cheung20011}, K-D-B-tree \cite{582321}, BV-tree \cite{223796}, UB-tree
+\cite{bayer1997universal}
 
@@
 
-XML (eXtensible Markup Language) is increasingly being used as a data exchange format.
-Accordingly, the ability to store and manage XML documents is needed.
-Salminen and Tompa~\cite{salminen2001requirements} define an XML database as ``a collection of XML documents
-and their parts, maintained by a system having capabilities to
-manage and control the collection itself and the information
-represented by that collection.'' Salminena and Tompa also highlight some of the requirements of an XML database system.
-These requirements include dealing with namespaces, Internet resources, querying parts of an XML document, and transformations of XML documents.
+XML (eXtensible Markup Language) is increasingly being used as a data exchange format. Accordingly, the ability to store and manage XML documents is needed.
+Salminen and Tompa~\cite{salminen2001requirements} define an XML database as ``a collection of XML documents and their parts, maintained by a system
+having capabilities to manage and control the collection itself and the information represented by that collection.'' Salminena and Tompa also highlight some of
+the requirements of an XML database system. These requirements include dealing with namespaces, Internet resources, querying parts of an XML document, and
+transformations of XML documents.
 
-There are two ways in which an XML database system can be designed.
-The first is to have the database internally translate the XML document into relational tables (XML-enabled databases).
-The second is to have data structures that can persistently store XML documents (native XML databases).
-In this section, we will discuss how the database's physical design needs to be adapted in both cases.
-Other challenges such as expressing and executing queries and updates are faced in XML database design, but these are beyond the scope of this survey.
+There are two ways in which an XML database system can be designed. The first is to have the database internally translate the XML document into relational
+tables (XML-enabled databases). The second is to have data structures that can persistently store XML documents (native XML databases). In this section, we will
+discuss how the database's physical design needs to be adapted in both cases. Other challenges such as expressing and executing queries and updates are faced in
+XML database design, but these are beyond the scope of this survey.
 
@@
 
-Despite the special nature of XML documents, the same index structures used in relational databases can still be utilized in XML databases.
-However, some adjustments may need to be made to fit the nature of the XML documents.
-For XML-enabled databases, XML documents are usually represented as relational tables, and then indexed similar to other tables.
-Pal et al.~\cite{pal2004indexing} describe how this is done in Microsoft SQL Server 2005.
-To start with, a new data type called `XML' was introduced.
-This data type could contain values of complete XML documents, or just fragments of XML data.
-First, the different nodes in the XML data are labelled with the ORDPATH~\cite{o2004ordpaths} mechanism.
-The XML data is then shredded into a relational table with five main columns:ORDPATH, TAG, NODE\_TYPE, VALUE, and PATH\_ID.
-This information is then stored in a $B^{+}-tree$ as the primary index of the XML type.
-Additionally, a secondary index can be created on any of the columns of the primary index to speed things up.
+Despite the special nature of XML documents, the same index structures used in relational databases can still be utilized in XML databases. However, some
+adjustments may need to be made to fit the nature of the XML documents. For XML-enabled databases, XML documents are usually represented as relational tables,
+and then indexed similar to other tables. Pal et al.~\cite{pal2004indexing} describe how this is done in Microsoft SQL Server 2005. To start with, a new data
+type called `XML' was introduced. This data type could contain values of complete XML documents, or just fragments of XML data. First, the different nodes in
+the XML data are labelled with the ORDPATH~\cite{o2004ordpaths} mechanism. The XML data is then shredded into a relational table with five main columns:ORDPATH,
+TAG, NODE\_TYPE, VALUE, and PATH\_ID. This information is then stored in a $B^{+}-tree$ as the primary index of the XML type. Additionally, a secondary index
+can be created on any of the columns of the primary index to speed things up.
 
 For native XML databases such as eXist~\cite{meier2009exist} or TIMBER~\cite{jagadish2002timber}, $B^{+}-trees$ were still used to index the XML documents.
-However, in contrast to XML-enabled databases, the XML data is not first stripped into a relational table.
-eXist uses a number schema to assign unique identifiers to each node in the XML document.
-These unique number identifiers allow the determination of any node's parent, sibling or possible child nodes.
-Four index files are created for XML data where all the indexes are based on $B^{+}-trees$.
-One index manages the collection hierarchy, one index collects nodes in a paged file and associates unique node identifiers to the actual nodes, one index indexes elements and attributes, and the last index keeps track of word occurrences and is used by the fulltext search extensions.
+However, in contrast to XML-enabled databases, the XML data is not first stripped into a relational table. eXist uses a number schema to assign unique
+identifiers to each node in the XML document. These unique number identifiers allow the determination of any node's parent, sibling or possible child nodes.
+Four index files are created for XML data where all the indexes are based on $B^{+}-trees$. One index manages the collection hierarchy, one index collects nodes
+in a paged file and associates unique node identifiers to the actual nodes, one index indexes elements and attributes, and the last index keeps track of word
+occurrences and is used by the fulltext search extensions.
 
-TIMBER also uses a similar numbering schema.
-There are minor differences between the two numbering techniques, but the main idea is the same.
-The numbering is based on three values of a node: its start label, its end label, and its level (i.e.
-nested depth).
-On the other hand, Sedna~\cite{taranov2010sedna} uses a different reasoning to produce its numbering schema.
-Sedna assignes string identifiers to nodes such that they are lexically ordered according to the position of the node.
-Despite the differences in the ways the numbering is constructed, it seems that that no novel indexing structure is needed for XML databases.
-The problem is rather how to assign the nodes unique identifiers (the numbers or labels in the numbering schema) that can be used as identifiers in the index to reflect the XML document structure.
+TIMBER also uses a similar numbering schema. There are minor differences between the two numbering techniques, but the main idea is the same. The numbering is
+based on three values of a node: its start label, its end label, and its level (i.e. nested depth). On the other hand, Sedna~\cite{taranov2010sedna} uses a
+different reasoning to produce its numbering schema. Sedna assignes string identifiers to nodes such that they are lexically ordered according to the position
+of the node. Despite the differences in the ways the numbering is constructed, it seems that that no novel indexing structure is needed for XML databases. The
+problem is rather how to assign the nodes unique identifiers (the numbers or labels in the numbering schema) that can be used as identifiers in the index to
+reflect the XML document structure.
 
@@
 
-Since XPath query processing is expensive, materialized views come in handy in XML databases.
-Balmin et al.~\cite{balmin2004framework} propose a framework for using XPath materialized views for XML query processing.
-(TALK ABOUT IT MORE)
+Since XPath query processing is expensive, materialized views come in handy in XML databases. Balmin et al.~\cite{balmin2004framework} propose a framework for
+using XPath materialized views for XML query processing. (TALK ABOUT IT MORE)
 
@@
 
-Oracle Database uses B tree indexes and their subtypes such as Index-organized tables, reverse key indexes, descending indexes, B-tree cluster indexes.
-It also uses Bitmap and bitmap join indexes, function based indexes and domain based indexes.
+Oracle Database uses B tree indexes and their subtypes such as Index-organized tables, reverse key indexes, descending indexes, B-tree cluster indexes. It also
+uses Bitmap and bitmap join indexes, function based indexes and domain based indexes.
 

### 2010-07-09 cs848-db/Paper.tex

@@
 
-The first physical database design decision to be considered is the choice of indexes to be implemented.
-The concept of indices has been around for a long time.
+The first physical database design decision to be considered is the choice of indexes to be implemented. The concept of indices has been around for a long time.
 Just like and index at the end of the book is provided to help find certain content faster, indices in an information system help find content in tables faster.
-The main idea of an index is to have each record in a table have a unique identifier (primary key) and then organize these identifiers in a certain way that allows for fast access of a specific record.
-Of course, indexing can be done on any column in the database and not necessarily the primary key.
-There are many type of indexes that have been used in information systems.
+The main idea of an index is to have each record in a table have a unique identifier (primary key) and then organize these identifiers in a certain way that
+allows for fast access of a specific record. Of course, indexing can be done on any column in the database and not necessarily the primary key. There are many
+type of indexes that have been used in information systems.
 
-B-trees (short for Balanced Trees) are the most used database index structure.
-The B-tree is essentially similar to the traditional Binary Search Tree, but instead of having one value per node, a B-tee can have many values per node~
-cite{comer1979ubiquitous} as shown in Figure (PUT A FIGURE FOR A B TREE HERE).
-B-trees are suitable for relational databases since the cost of retrieval in a B-tree is at most proportional to: $log_{d}\frac{n+1}{2}$.
-The cost of insertion and deletion is at most proportional to $log_{d}n$ due to the possibility of progressing back up the tree to balance it after an insertion or a deletion.
+B-trees (short for Balanced Trees) are the most used database index structure.  The B-tree is essentially similar to the traditional Binary Search Tree, but
+instead of having one value per node, a B-tee can have many values per node~\cite{comer1979ubiquitous} as shown in Figure (PUT A FIGURE FOR A B TREE HERE).
+B-trees are suitable for relational databases since the cost of retrieval in a B-tree is at most proportional to: $log_{d}\frac{n+1}{2}$. The cost of insertion
+and deletion is at most proportional to $log_{d}n$ due to the possibility of progressing back up the tree to balance it after an insertion or a deletion.
 Although B-trees do well in retrieval, deletion and insertion, they do not perform well in sequential search.
 
-Different variations of B-trees have been used.
-For example, $B^{+}-trees$ are now the main method of indexing in Disk Based Relational Databases (DRDB).
-These will be discussed in Section~\ref{SEC-DDRDB}.
-The T Tree has been proposed for main memory databases, and will be discussed in Section~\ref{SEC-MMDB}.
-Cache sensitive indexes were then proposed.
-A variation of T Trees, Cache Sensitive T-Trees (CST-Trees) was introduced, and is also discussed in Section~\ref{SEC-MMDB}.
-On the same note, Cache Conscious $B^{+}-trees$ and Prefetching $J^{+}-Trees$ are also discussed in Section~\ref{SEC-MMDB}.
+Different variations of B-trees have been used. For example, $B^{+}-trees$ are now the main method of indexing in Disk Based Relational Databases (DRDB). These
+will be discussed in Section~\ref{SEC-DDRDB}. The T Tree has been proposed for main memory databases, and will be discussed in Section~\ref{SEC-MMDB}. Cache
+sensitive indexes were then proposed. A variation of T Trees, Cache Sensitive T-Trees (CST-Trees) was introduced, and is also discussed in
+Section~\ref{SEC-MMDB}. On the same note, Cache Conscious $B^{+}-trees$ and Prefetching $J^{+}-Trees$ are also discussed in Section~\ref{SEC-MMDB}.
 
-Of course, there are other indexing techniques used such as hash tables, and bitmap indices.
-For example, Bitmap indices are commonly used in Data Warehouses, and will be discussed in details in Section~\ref{SEC-WAREHOUSES}.
+Of course, there are other indexing techniques used such as hash tables, and bitmap indices. For example, Bitmap indices are commonly used in Data Warehouses,
+and will be discussed in details in Section~\ref{SEC-WAREHOUSES}.
 
@@
 
-Two considerations are taken when designing index structures for MMDB.
-The first is that the data resides in main memory and not on disk, and so many of the considerations taken for I/O operations is no longer there.
-The second is having index structures cache conscious.
-Lots of research work has shown that cache performance is very critical in MMDBs~\cite{rao1999cache, boncz1999database}.
-This is because the big difference between processor speed and main memory access speed.
-Thus, a cache miss is still relatively expensive in main memory databases.
-Accordingly, we will see that many of the index data structures proposed for main memory databases focused on being cache conscious or cache sensitive.
+Two considerations are taken when designing index structures for MMDB. The first is that the data resides in main memory and not on disk, and so many of the
+considerations taken for I/O operations is no longer there. Instead of focusing on disk access and disk storage, the data structures for main memory
+databases should focus on the efficient use of CPU cycles and memory~\cite{lehman1986study}. The second is having index structures cache conscious. Cache
+memories improve performance by holding recently referenced data~\cite{smith1982cache}. If the memory reference is found in the cache, then execution proceeds
+at processor speed. Otherwise, the data has to be fetched from main memory. With the big difference between processor speed and main memory speed, cache misses
+are still relatively expensive. Lots of research work has shown that cache performance is very critical in MMDBs~\cite{boncz1999database,rao1999cache}. This is
+because the big difference between processor speed and main memory access speed. Thus, a cache miss is still relatively expensive in main memory databases.
+Accordingly, we will see that many of the index data structures proposed for main memory databases focused on being cache conscious or cache sensitive. In
+summary, a main memory index should be able to reduce overall computation time while using as little memory as possible~\cite{lehman1986study}. It should also
+be noted that unlike disk based systems where it is advantageous to store actual attribute values in the index, main memory systems do not require that.
+Instead, pointers to the actual attribute values can be used. This, in turn, will reduce the size of the index which is preferable in this case.
 
-\subsubsection*{$B^{+}-Trees$}
-The $B^{+}-Tree$, explained in the previous section, is originally designed for disk based database systems.
-However, it has also been found useful for main memory databases (REF) as it is cache conscious.
+cite~\cite{dewitt1984implementation}.
+
+\subsubsection*{B Trees}
+Although B Trees, explained in the previous section, are originally designed for disk based database systems, they were found to also be useful in
+MMDB~\cite{lehman1986study}. In disk based systems, the B+ Tree is more used than the regular B Tree. However, in main memory databases, the B Tree would be
+more suitable since there is no gain from keeping all the data in the leaves. In a main memory system, this would only waste space without improving
+performance. In a B or B+ Tree, multiple keys are used to search within a node. This means that if the node fits in a cache line, this cache load can satisfy
+more than more comparison leading to better cache utilization~\cite{rao1999cache}. Therefore, B Trees in general have good cache behavior.
+
+However, the cache behavior of B+ Trees could still be improved. Therefore, Rao and Ross~\cite{rao2000making} propose Cache Sensitive B+-Trees
+($CSB^{+}-Trees$)~\cite{rao2000making} as an index structure for main memory. To do so, they eliminated most of the child pointers and had more keys in each
+node to improve locality and reduce tree height~\cite{luan2009prefetching}. Since the number of cache misses in search operations is proportional to the height
+of the tree, $CSB^{+}-trees$ have fewer cache misses, and thus better performance.
 
 \subsubsection*{T Trees}
-Lehmen and Carey~\cite{lehman1986study} study the different features needed for indices of MMDB.
-When designing data structures for MMDB, we are mainly concerned with the efficient use of CPU cycles and memory.
-Accordingly, the main goal of an index structure for MMDB is to minimize the overall computation time while using the minimum memory possible.
-Since all the database data is in main memory, index structures do not need to store actual values like those in disk-oriented databases.
-Instead, a pointer to the data instead of the actual data can be stored.
-Lehmen and Carey introduce a new index structure, the T Tree, which combines the features from AVL Trees and B Trees that are suited to main memory.
-Figure~\ref{fig:ttree} shows the proposed structure of the T Tree.
-They run some experiments on data residing in main memory to compare the performance of the T Tree to existing structures.
-All structures were modified to contain pointers to data values rather than the data values themselves.
-The structures included were  AVL Trees, simple arrays, B Trees, Chained Bucket Hashing, Extendible Hashing, Linear Hashing and Modified Linear Hashing.
-Their experiments showed that for unordered data, Modified Linear Hashing gave the best performance, and for ordered data, T Trees gave the best performance for a mix of searches, inserts and deletes.
+Lehmen and Carey~\cite{lehman1986study} introduce a new index structure, the T Tree, which combines the features from AVL Trees and B Trees that are suited to
+main memory. Figure~\ref{fig:ttree} shows the proposed structure of the T Tree. They run some experiments on data residing in main memory to compare the
+performance of the T
+Tree to existing structures. All structures were modified to contain pointers to data values rather than the data values themselves. The structures
+compared included  AVL Trees, simple arrays, B Trees, Chained Bucket Hashing, Extendible Hashing, Linear Hashing and Modified Linear Hashing. Their
+experiments showed that for unordered data, Modified Linear Hashing gave the best performance, and for ordered data, T Trees gave the best performance for a mix
+of searches, inserts and deletes. This is mainly because a T node contains many elements which results in good update and storage characteristics. The T Tree
+was the first index structure specifically designed for main memory databases.
+
+Although T Trees have been specifically designed for MMDB, they did not consider the cache behavior at that time~\cite{rao1999cache}. Although many keys are
+fit into one node, only the two end keys are used for comparison leading to low node utilization. Similarly, array binary search trees have the same problem.
 
@@
 
-\subsubsection*{Cache Sensitive Search Trees (CSS-Trees)}
-\cite{rao1999cache}.
-
-\subsubsection*{Cache Sensitive T-Trees (CST-Trees)}
-
-T Trees were the main index structure used for main memory databases for some time after their proposal.
-However, it was discovered that B+-trees outperform T-Trees on modern processors because of the growth of cache miss latency relative to processor speed~\cite{rao1999cache,lee2007cst}.
-This is because the height of the tree is high which makes the total number of memory accesses from the root to the leaf node higher.
-The other reason is that the node size is not aligned with the cache line size.
-The other problem with T-Trees is that there is a big waste of space in the cache because unnecessary data is brought into the cache.
+T Trees were the main index structure used for main memory databases for some time after their proposal. However, it was discovered that B+-trees outperform
+T-Trees on modern processors because of the growth of cache miss latency relative to processor speed~\cite{rao1999cache,lee2007cst}. This is because the height
+of the tree is high which makes the total number of memory accesses from the root to the leaf node higher. The other reason is that the node size is not aligned
+with the cache line size. Another problem with T-Trees is that there is a big waste of space in the cache because unnecessary data is brought into the cache.
 Additionally, record pointers stored in the tree take up a lot of space.
 
-To resolve these issues, Lee et al.~
-cite{lee2007cst} propose the Cache Sensitive T-Trees (CST-Trees).
-This is achieved in the following ways.
-First, a binary search tree whose node values are the maximum key of each node in the T Tre is constructed.
-That way, for a given value, the node containing this value could be quickly located through searching the binary tree.
-Second, the need to store pointers is eliminated by storing the tree in an array and locating the necessary nodes through index calculation.
-Finally, node sizes are aligned with cache line size such that there are no cache misses when accessing data in each binary search tree in the array.
+To resolve these issues, Lee et al.~\cite{lee2007cst} propose the Cache Sensitive T-Trees (CST-Trees). This is achieved in the following ways. First, a binary
+search tree whose node values are the maximum key of each node in the T Tre is constructed. That way, for a given value, the node containing this value could be
+quickly located through searching the binary tree. Second, the need to store pointers is eliminated by storing the tree in an array and locating the necessary
+nodes through index calculation. Finally, node sizes are aligned with cache line size such that there are no cache misses when accessing data in each binary
+search tree in the array.
 
-\subsubsection*{Cache Sensitive $B^{+}-trees$ ($CSB^{+}-trees)$}
+\subsubsection*{Cache Sensitive Search Trees (CSS-Trees)}
 
-Rao and Ross~\cite{rao1999cache} show that $B^{+}-trees$ are more cache conscious than binary search trees and T-Trees which means they are more suited to main memory databases.
-However, traditional B+-Trees still had poor cache performance.
-Therefore, they propose Cache Sensitive B+-Trees ($CSB^{+}-Trees$)~\cite{rao2000making} as an index structure for main memory.
-To do so, they eliminated most of the child pointers and had more keys in each node to improve locality and reduce tree height~\cite{luan2009prefetching}.
-Since the number of cache misses in search operations is proportional to the height of the tree, $CSB^{+}-trees$ have fewer cache misses, and thus better performance.
+Binary Search Trees have poor cache behavior when the array is much bigger than the cache. Rao and Ross~\cite{rao1999cache} propose the Cache Sensitive Search
+Tree (CSS-Tree) to improve this cache behavior. To do that, the search tree is stored as a directory in an array. The number of keys per node is chosen such
+that the whole node fits in a cache line. This improves local searching within a node to just one cache miss. They also hard-code the traversal within a node
+such that finding the next node happens in $log_{m+1}n$ times rather than $log_{2}n$ times in a binary search tree. However, CSS Trees are only suitable for
+read environments since they eliminate nearly all child pointers, thus removing the support for incremental update.
 
-\subsubsection*{$Prefetching B^{+}-Trees (pB^{+}-trees)$}
+\subsubsection*{Prefetching B+ Trees (pB+ Trees)}
 
-Chen et al.~\cite{chen2001improving} propose using prefetching to improve the performance of $B^{+}-trees$.
-Since current computer systems can prefetch different data simultaneously, the author take advantage of this.
-The node size of the $B^{+}-tree$ is made wider than the cache line size.
-This way, several cache misses can be overlapped at the same time thus improving performance.
+Chen et al.~\cite{chen2001improving} propose using prefetching to improve the performance of B+ Trees. Since current computer systems can prefetch
+different data simultaneously, the authors take advantage of this. The idea here is to increase the node size of B+ Trees and employ prefectching at the same
+time. In a B+ Tree, every time we move down a level in the tree, a cache miss occurs. Accordingly, prefetching larger size nodes means that we can overlap
+multiple cache misses before they occur.
 
@@
 
+\subsection{Partitioning}
+
+Partitioning is used in disk based systems to divide the data into the physical pages in a way that will yield the best performance. This could be through
+vertical partitioning or horizontal partitioning, as previously explained. In main memory databases, we do not have the issue of expensive disk access which
+makes partitioning in memory unnecessary as it will not save any costs. However, partionining is needed in the secondary storage of the data on
+disk~\cite{gruenwald1990database}. Disk storage is still used as a backup of main memory databases. When a crash occurs, a reload of the database from disk or
+archive memory takes place. In order to avoid page faults caused by referencing data that is still being reloaded, the reloading process must be designed to be
+efficient in loading the important data such that the database users are not affected. Gruenwald and Eich~\cite{gruenwald1990database,gruenwald1990choosing}
+
+\subsection{Clustering}
+Clustering in disk based systems is important because sequential I/O access is cheaper than random or dispersed I/O access. This is not the case with main
+memory, and so clustering is not needed in MMDB~\cite{garcia1992main, moldovan2008databases}. Therefore, components of one object may be spread across memory
+without impacting performance.
+
 \subsection{Data Warehouses}
 \label{SEC-WAREHOUSES}
-A data warehouse is a subject-oriented, integrated, time-variant, and non-volatile collection of data and decision support technologies, aimed at enabling the knowledge worker (executive, manager, analyst) to make better and faster decisions \cite{560407, 248616}.
-More than half of IT executives named Data warehousing as the highest-priority post-millennium project for them \cite{sen2005comparison}.
-The value of data warehousing for an organization depends on the organization's need for reliable, consolidated, unique and integrated reporting and analysis of its data, at different levels of aggregation.
+A data warehouse is a subject-oriented, integrated, time-variant, and non-volatile collection of data and decision support technologies, aimed at enabling the
+knowledge worker (executive, manager, analyst) to make better and faster decisions \cite{560407, 248616}. More than half of IT executives named Data warehousing
+as the highest-priority post-millennium project for them \cite{sen2005comparison}. The value of data warehousing for an organization depends on the
+organization's need for reliable, consolidated, unique and integrated reporting and analysis of its data, at different levels of aggregation.
 
-Data warehousing has been shown useful in many industries: manufacturing (for order shipment and customer support), retail (for user profiling and inventory management), financial services (for claims analysis, risk analysis, credit card analysis, and fraud detection), transportation (for fleet management), telecommunications (for call analysis and fraud detection), utilities (for power usage analysis), and healthcare (for outcomes analysis) \cite{248616}.
+Data warehousing has been shown useful in many industries: manufacturing (for order shipment and customer support), retail (for user profiling and inventory
+management), financial services (for claims analysis, risk analysis, credit card analysis, and fraud detection), transportation (for fleet management),
+telecommunications (for call analysis and fraud detection), utilities (for power usage analysis), and healthcare (for outcomes analysis) \cite{248616}.
 
-Data warehouses tend to be extremely large, in fact it is quite possible for a data warehouse to store tens of petabytes of data and load tens of terabytes of data everyday \cite{thusoo2010data}.
-Datta et al.
-\cite {628286} notes that the information in a data warehouse is usually multidimensional in nature, requiring the capability to view the data from a variety of perspectives.
-Aggregated and summarized data become more crucial than detailed records in such environment.
-Therefore, the workloads are query intensive with mostly ad hoc, complex queries, often requiring computationally expensive operations such as scans, joins, and aggregation.
-Performing such operations on large amounts of data, like in the case of data warehousing, complicates the situation further.
-Moreover, the results have to be delivered interactively to the business analyst using the system.
+Data warehouses tend to be extremely large, in fact it is quite possible for a data warehouse to store tens of petabytes of data and load tens of terabytes of
+data everyday \cite{thusoo2010data}. Datta et al. \cite {628286} notes that the information in a data warehouse is usually multidimensional in nature, requiring
+the capability to view the data from a variety of perspectives. Aggregated and summarized data become more crucial than detailed records in such environment.
+Therefore, the workloads are query intensive with mostly ad hoc, complex queries, often requiring computationally expensive operations such as scans, joins, and
+aggregation. Performing such operations on large amounts of data, like in the case of data warehousing, complicates the situation further. Moreover, the results
+have to be delivered interactively to the business analyst using the system.
 
-The data warehouse supports on-line analytical processing (OLAP), the functional and performance requirements of which are quite different from those of the on-line transaction processing (OLTP) applications traditionally supported by the operational databases.
-Given the above challenges, the data warehouse is usually maintained separately from the organization’s operational databases \cite{248616}.
-Query throughput and response times are more important than transaction throughput.
+The data warehouse supports on-line analytical processing (OLAP), the functional and performance requirements of which are quite different from those of the
+on-line transaction processing (OLTP) applications traditionally supported by the operational databases. Given the above challenges, the data warehouse is
+usually maintained separately from the organization’s operational databases \cite{248616}. Query throughput and response times are more important than
+transaction throughput.
 

### 2010-07-09 cs848-db/Paper.tex

@@
 
+Chen et al.~\cite{chen2001improving} propose using prefetching to improve the performance of B+ Trees. Since current computer systems can prefetch
+different data simultaneously, the authors take advantage of this. The idea here is to increase the node size of B+ Trees and employ prefectching at the same
+time. In a B+ Tree, every time we move down a level in the tree, a cache miss occurs. Accordingly, prefetching larger sizes nodes means that we can overlap
+multiple cache misses before they occur. This is because in one cache mix, extra data is retrieved in parallel with the originally requested data if it can be
+predicted sufficiently early. Figure~\ref{fig:prefetching} shows how prefetching retrieves the same data in less time. This method, however, does not
+eliminate the full cache miss latency with every level in the tree.
+
+\begin{figure}[!t]
+\centering
+\includegraphics[width=7cm]{figs/prefetching.png}
+\caption{Performance of various B+ Tree searches where a cache miss to memory takes 150 cycles and a subsequent access can begin 10 cycles
+later~\cite{chen2001improving}}
+\label{fig:prefetching}
+\end{figure}
+
 \subsubsection*{T Trees}
@@
 \centering
-\includegraphics[width=9cm]{figs/Ttree.png}
+\includegraphics[width=7cm]{figs/Ttree.png}
 \caption{T Tree by Lehmen and Carey~\cite{lehman1986study}}
@@
 
-\subsubsection*{Prefetching B+ Trees (pB+ Trees)}
-
-Chen et al.~\cite{chen2001improving} propose using prefetching to improve the performance of B+ Trees. Since current computer systems can prefetch
-different data simultaneously, the authors take advantage of this. The idea here is to increase the node size of B+ Trees and employ prefectching at the same
-time. In a B+ Tree, every time we move down a level in the tree, a cache miss occurs. Accordingly, prefetching larger size nodes means that we can overlap
-multiple cache misses before they occur.
-
-\subsubsection*{$J^{+}-trees$ and $pJ^{+}-trees$}
+\subsubsection*{J+-trees and pJ+-trees}
+\cite{luan2009prefetching}
 

### 2010-07-09 trunk/cs848-db/Paper.tex
finished intro for dw

@@
 considerations taken for I/O operations is no longer there. Instead of focusing on disk access and disk storage, the data structures for main memory
-databases should focus on the efficient use of CPU cycles and memory~\cite{lehman1986study}. The second is having index structures cache conscious. Lots of
-research work has shown that cache performance is very critical in MMDBs~\cite{boncz1999database,rao1999cache}. This is because the big difference between
-processor speed and main memory access speed. Thus, a cache miss is still relatively expensive in main memory databases. Accordingly, we will see that many of
-the index data structures proposed for main memory databases focused on being cache conscious or cache sensitive. In summary, a main memory index should be
-able to reduce overall computation time while using as little memory as possible~\cite{lehman1986study}. It should also be noted that unlike disk based systems
-where it is advantageous to store actual attribute values in the index, main memory systems do not require that. Instead, pointers to the actual attribute
-values can be used. This, in turn, will reduce the size of the index which is preferable in this case.
+databases should focus on the efficient use of CPU cycles and memory~\cite{lehman1986study}. The second is having index structures cache conscious. Cache
+memories improve performance by holding recently referenced data~\cite{smith1982cache}. If the memory reference is found in the cache, then execution proceeds
+at processor speed. Otherwise, the data has to be fetched from main memory. With the big difference between processor speed and main memory speed, cache misses
+are still relatively expensive. Lots of research work has shown that cache performance is very critical in MMDBs~\cite{boncz1999database,rao1999cache}. This is
+because the big difference between processor speed and main memory access speed. Thus, a cache miss is still relatively expensive in main memory databases.
+Accordingly, we will see that many of the index data structures proposed for main memory databases focused on being cache conscious or cache sensitive. In
+summary, a main memory index should be able to reduce overall computation time while using as little memory as possible~\cite{lehman1986study}. It should also
+be noted that unlike disk based systems where it is advantageous to store actual attribute values in the index, main memory systems do not require that.
+Instead, pointers to the actual attribute values can be used. This, in turn, will reduce the size of the index which is preferable in this case.
 
@@
 more suitable since there is no gain from keeping all the data in the leaves. In a main memory system, this would only waste space without improving
-performance.
+performance. In a B or B+ Tree, multiple keys are used to search within a node. This means that if the node fits in a cache line, this cache load can satisfy
+more than more comparison leading to better cache utilization~\cite{rao1999cache}. Therefore, B Trees in general have good cache behavior.
+
+However, the cache behavior of B+ Trees could still be improved. Therefore, Rao and Ross~\cite{rao2000making} propose Cache Sensitive B+-Trees
+($CSB^{+}-Trees$)~\cite{rao2000making} as an index structure for main memory. To do so, they eliminated most of the child pointers and had more keys in each
+node to improve locality and reduce tree height~\cite{luan2009prefetching}. Since the number of cache misses in search operations is proportional to the height
+of the tree, $CSB^{+}-trees$ have fewer cache misses, and thus better performance.
+
+Chen et al.~\cite{chen2001improving} propose using prefetching to improve the performance of B+ Trees. Since current computer systems can prefetch
+different data simultaneously, the authors take advantage of this. The idea here is to increase the node size of B+ Trees and employ prefectching at the same
+time. In a B+ Tree, every time we move down a level in the tree, a cache miss occurs. Accordingly, prefetching larger sizes nodes means that we can overlap
+multiple cache misses before they occur. This is because in one cache mix, extra data is retrieved in parallel with the originally requested data if it can be
+predicted sufficiently early. Figure~\ref{fig:prefetching} shows how prefetching retrieves the same data in less time. This method, however, does not
+eliminate the full cache miss latency with every level in the tree.
+
+\begin{figure}[!t]
+\centering
+\includegraphics[width=7cm]{figs/prefetching.png}
+\caption{Performance of various B+ Tree searches where a cache miss to memory takes 150 cycles and a subsequent access can begin 10 cycles
+later~\cite{chen2001improving}}
+\label{fig:prefetching}
+\end{figure}
 
 \subsubsection*{T Trees}
-Lehmen and Carey introduce a new index structure, the T Tree, which combines the features from AVL Trees and B Trees that are suited to main memory.
-Figure~\ref{fig:ttree} shows the proposed structure of the T Tree. They run some experiments on data residing in main memory to compare the performance of the T
+Lehmen and Carey~\cite{lehman1986study} introduce a new index structure, the T Tree, which combines the features from AVL Trees and B Trees that are suited to
+main memory. Figure~\ref{fig:ttree} shows the proposed structure of the T Tree. They run some experiments on data residing in main memory to compare the
+performance of the T
 Tree to existing structures. All structures were modified to contain pointers to data values rather than the data values themselves. The structures
@@
 experiments showed that for unordered data, Modified Linear Hashing gave the best performance, and for ordered data, T Trees gave the best performance for a mix
-of searches, inserts and deletes. This is mainly because a T node contains many elements which results in good update and storage characteristics.
+of searches, inserts and deletes. This is mainly because a T node contains many elements which results in good update and storage characteristics. The T Tree
+was the first index structure specifically designed for main memory databases.
+
+Although T Trees have been specifically designed for MMDB, they did not consider the cache behavior at that time~\cite{rao1999cache}. Although many keys are
+fit into one node, only the two end keys are used for comparison leading to low node utilization. Similarly, array binary search trees have the same problem.
 
@@
 
-\subsubsection*{Cache Sensitive Search Trees (CSS-Trees)}
-\cite{rao1999cache}.
-
-\subsubsection*{Cache Sensitive T-Trees (CST-Trees)}
-
 T Trees were the main index structure used for main memory databases for some time after their proposal. However, it was discovered that B+-trees outperform
@@
 of the tree is high which makes the total number of memory accesses from the root to the leaf node higher. The other reason is that the node size is not aligned
-with the cache line size. The other problem with T-Trees is that there is a big waste of space in the cache because unnecessary data is brought into the cache.
+with the cache line size. Another problem with T-Trees is that there is a big waste of space in the cache because unnecessary data is brought into the cache.
 Additionally, record pointers stored in the tree take up a lot of space.
@@
 
-\subsubsection*{Cache Sensitive B+ trees (CSB+ Trees)}
+\subsubsection*{Cache Sensitive Search Trees (CSS-Trees)}
 
-Rao and Ross~\cite{rao1999cache} show that $B^{+}-trees$ are more cache conscious than binary search trees and T-Trees which means they are more suited to main
-memory databases. However, traditional B+-Trees still had poor cache performance. Therefore, they propose Cache Sensitive B+-Trees
-($CSB^{+}-Trees$)~\cite{rao2000making} as an index structure for main memory. To do so, they eliminated most of the child pointers and had more keys in each
-node to improve locality and reduce tree height~\cite{luan2009prefetching}. Since the number of cache misses in search operations is proportional to the height
-of the tree, $CSB^{+}-trees$ have fewer cache misses, and thus better performance.
+Binary Search Trees have poor cache behavior when the array is much bigger than the cache. Rao and Ross~\cite{rao1999cache} propose the Cache Sensitive Search
+Tree (CSS-Tree) to improve this cache behavior. To do that, the search tree is stored as a directory in an array. The number of keys per node is chosen such
+that the whole node fits in a cache line. This improves local searching within a node to just one cache miss. They also hard-code the traversal within a node
+such that finding the next node happens in $log_{m+1}n$ times rather than $log_{2}n$ times in a binary search tree. However, CSS Trees are only suitable for
+read environments since they eliminate nearly all child pointers, thus removing the support for incremental update.
 
-\subsubsection*{$Prefetching B^{+}-Trees (pB^{+}-trees)$}
-
-Chen et al.~\cite{chen2001improving} propose using prefetching to improve the performance of $B^{+}-trees$. Since current computer systems can prefetch
-different data simultaneously, the author take advantage of this. The node size of the $B^{+}-tree$ is made wider than the cache line size. This way, several
-cache misses can be overlapped at the same time thus improving performance.
-
-\subsubsection*{$J^{+}-trees$ and $pJ^{+}-trees$}
+\subsubsection*{J+-trees and pJ+-trees}
+\cite{luan2009prefetching}
 
@@
 \label{SEC-WAREHOUSES}
-A data warehouse is a subject-oriented, integrated, time-variant, and non-volatile collection of data and decision support technologies, aimed at enabling the
-knowledge worker (executive, manager, analyst) to make better and faster decisions \cite{560407, 248616}. More than half of IT executives named Data warehousing
-as the highest-priority post-millennium project for them \cite{sen2005comparison}. The value of data warehousing for an organization depends on the
-organization's need for reliable, consolidated, unique and integrated reporting and analysis of its data, at different levels of aggregation.
+A data warehouse is a subject-oriented, integrated, time-variant, and non-volatile collection of data and decision support technologies, aimed at enabling the knowledge worker (executive, manager, analyst) to make better and faster decisions \cite{560407, 248616}.
+More than half of IT executives named Data warehousing as the highest-priority post-millennium project for them \cite{sen2005comparison}.
+The value of data warehousing for an organization depends on the organization's need for reliable, consolidated, unique and integrated reporting and analysis of its data, at different levels of aggregation.
 
@@
 
-Data warehouses tend to be extremely large, in fact it is quite possible for a data warehouse to store tens of petabytes of data and load tens of terabytes of
-data everyday \cite{thusoo2010data}. Datta et al. \cite {628286} notes that the information in a data warehouse is usually multidimensional in nature, requiring
-the capability to view the data from a variety of perspectives. Aggregated and summarized data become more crucial than detailed records in such environment.
-Therefore, the workloads are query intensive with mostly ad hoc, complex queries, often requiring computationally expensive operations such as scans, joins, and
-aggregation. Performing such operations on large amounts of data, like in the case of data warehousing, complicates the situation further. Moreover, the results
-have to be delivered interactively to the business analyst using the system.
+Data warehouses tend to be extremely large, in fact it is quite possible for a data warehouse to store tens of petabytes of data while loading tens of terabytes of data everyday \cite{thusoo2010data}.
+Datta et al.
+\cite {628286} notes that the information in a data warehouse is usually multidimensional in nature, requiring the capability to view the data from a variety of perspectives.
+Aggregated and summarized data become more crucial than detailed records in such environment.
+Therefore, the workloads are query intensive with mostly ad hoc, complex queries, often requiring computationally expensive operations such as scans, joins, and aggregation.
+Performing such operations on large amounts of data, like in the case of data warehousing, complicates the situation further.
+Moreover, the results have to be delivered interactively to the business analyst using the system.
 
-The data warehouse supports on-line analytical processing (OLAP), the functional and performance requirements of which are quite different from those of the
-on-line transaction processing (OLTP) applications traditionally supported by the operational databases. Given the above challenges, the data warehouse is
-usually maintained separately from the organization’s operational databases \cite{248616}. Query throughput and response times are more important than
-transaction throughput.
+Although data in a data warehouse is extracted and loaded from multiple on-line transaction processing (OLTP) data sources (including DB2, Oracle, IMS databases, and flat files) using Extract, Transfer, and Load (ETL) tools (see Figure \ref{fig:dw}), a data warehouse is usually maintained seperately from the organization's operational databases \cite{sen2005comparison, 248616}.
+This architecture can be justified owing to the fact that operational databases are finely tuned to satisfy known OLTP requirements and functionalities which are quite different from that of on-line analytical processing (OLAP) which is supported by data warehousing.
+Analyzing data for decision support usually requires consolidating data from many heterogeneous sources of varying quality, or use inconsistent representations, codes and formats.
+Moreover, understanding trends or making predictions requires historical data, whereas operational databases store only current data.
+In OLAP, there's a need to support multidimensional data models and operations which requires special data organization, access methods, and implementation methods, not generally provided by DBMSs targeted for OLTP \cite{248616}.
+Finally, in data warehousing, query throughput and response times are more important than transaction throughput which is the major performance metric for operational databases.
 
-OLAP vs OLTP
-
-OLTP \cite{sen2005comparison}
-
-Data warehouses, in contrast, are targeted for decision support. Historical, summarized and consolidated data is more important than detailed, individual records.
-
-The workloads are query intensive with mostly ad hoc, complex queries that can access millions of records and perform a lot of scans, joins, and aggregates.
-Query throughput and response times are more important than transaction throughput.
-
-Operational databases are finely tuned to support known OLTP workloads.
-
-Understanding trends or making predictions requires historical data, whereas operational databases store only current data.
-
-Decision support usually requires consolidating data from many heterogeneous sources of varying quality, or use inconsistent representations, codes and formats.
-
-Supporting the multidimensional data models and operations typical of OLAP requires special data organization, access methods, and implementation methods, not
-generally provided by DBMSs targeted for OLTP.
+\begin{figure}[!t]
+\centering
+\includegraphics[width=9cm]{figs/dw.png}
+\caption{Data Warehousing Architecture \cite{248616}}
+\label{fig:dw}
+\end{figure}
 
@@
 
-- joins in data warehouses are usually expensive since the fact table (the largest one) participates in every join \cite{628286}
+Joins in data warehouses are usually expensive since the fact table (the largest one) participates in every join \cite{628286}
 

### 2010-07-09 cs848-db/Paper.tex

@@
 
-A Main Memory Database (MMDB) is one where the data resides in the main memory of the system rather than on a disk~\cite{garcia1992main}. It should be noted
-that this is different from caching. Disk based database systems use the main memory to cache query results. However, MMDB's primary copy of the data resides in
-main memory. MMDB have many implementation challenges that have been addressed throughout the research community.
+A Main Memory Database (MMDB) (sometimes referred to as in-memory database) is one where the data resides in the main memory of the system rather than on a
+disk~\cite{garcia1992main}. It should be noted that this is different from caching. Disk based database systems use the main memory to cache query results.
+However, MMDB's primary copy of the data resides in main memory. MMDB have many implementation challenges that have been addressed throughout the research
+community.
 
@@
 Instead, pointers to the actual attribute values can be used. This, in turn, will reduce the size of the index which is preferable in this case.
+Table~\ref{tab:mmdbindexsumm} summarizes the features of the different index structures discussed in this section with respect to MMDB.
 
-cite~\cite{dewitt1984implementation}.
+\begin{table}[t!]
+\centering
+\scalebox{0.7}{
+\begin{tabular}{p{2cm}|p{6cm}|p{6cm}}
+\hline
+ & \textbf{Features suitable for MMDB} & \textbf{Cache Consciousness}\\
+\hline\hline
+B Trees &
+\begin{itemize}\item Good storage utilization
+\item Reasonably quick searching
+\item Fast updating
+\end{itemize} 
+& \begin{itemize}\item Reasonable cache behavior if node fits in cache line\end{itemize}\\
+\hline
+B+ Trees & 
+\begin{itemize}
+\item Wastes space since all data is stored in the leaves
+\item Reasonably quick searching
+\item Fast updating
+\end{itemize} &
+\begin{itemize}
+\item Reasonable cache behavior if node fits in cache line\end{itemize}\\
+\hline
+CSB+ Trees & \begin{itemize}
+              \item Same features of B+ Trees
+             \end{itemize} &
+\begin{itemize}
+ \item Good cache performance due to improved locality \& lower tree height
+\end{itemize}\\
+\hline
+pB+ Trees & \begin{itemize}
+              \item Same features of B+ Trees
+             \end{itemize} &
+\begin{itemize}
+ \item Improves cache miss performance by prefetching more data in each cache miss
+\item Node size is also increased leading to the same benefits of CSB+ Trees
+\end{itemize}\\
+\hline
+T Tree &  \begin{itemize}
+              \item Contain pointers to data values instead of the values themselves leading to better space utilization
+\item Node size is bigger leading to previous mentioned advantages
+             \end{itemize} &
+\begin{itemize}
+ \item Cache behavior was not considered at time of design
+\end{itemize}\\
+\hline
+CST & \begin{itemize}
+       \item Fast traversal in $log_{m+1}n$
+      \end{itemize} &
+\begin{itemize}
+ \item Good cache behavior as number of key nodes per line is chosen according to cache line miss.
+\end{itemize}\\
+\hline
+J+ Tree & \begin{itemize}
+       \item 
+      \end{itemize} &
+\begin{itemize}
+ \item 
+\end{itemize}\\
+\hline
+pJ+ Tree & \begin{itemize}
+       \item Fast traversal in $log_{m+1}n$
+      \end{itemize} &
+\begin{itemize}
+ \item
+\end{itemize}\\
+\hline
+\end{tabular}
+}
+\caption{Summary of index structures for Main Memory Databases (MMDB)}
+\label{tab:mmdbindexsumm}
+\end{table}
+
+%cite~\cite{dewitt1984implementation}.
 
@@
 MMDB~\cite{lehman1986study}. In disk based systems, the B+ Tree is more used than the regular B Tree. However, in main memory databases, the B Tree would be
-more suitable since there is no gain from keeping all the data in the leaves. In a main memory system, this would only waste space without improving
+more suitable since there is no gain from keeping all the datGood storage utilizationa in the leaves. In a main memory system, this would only waste space
+without improving
 performance. In a B or B+ Tree, multiple keys are used to search within a node. This means that if the node fits in a cache line, this cache load can satisfy
-more than more comparison leading to better cache utilization~\cite{rao1999cache}. Therefore, B Trees in general have good cache behavior.
+more than more comparison leading to better cache utilization~\cite{rao1999cache}. Therefore, B and B+ Trees in general have reasonable cache behavior.
 
@@
 
-Chen et al.~\cite{chen2001improving} propose using prefetching to improve the performance of B+ Trees. Since current computer systems can prefetch
-different data simultaneously, the authors take advantage of this. The idea here is to increase the node size of B+ Trees and employ prefectching at the same
-time. In a B+ Tree, every time we move down a level in the tree, a cache miss occurs. Accordingly, prefetching larger sizes nodes means that we can overlap
-multiple cache misses before they occur. This is because in one cache mix, extra data is retrieved in parallel with the originally requested data if it can be
-predicted sufficiently early. Figure~\ref{fig:prefetching} shows how prefetching retrieves the same data in less time. This method, however, does not
-eliminate the full cache miss latency with every level in the tree.
+Chen et al.~\cite{chen2001improving} propose using prefetching to improve the performance of B+ Trees. They call the new index structure prefetching B+ Tree
+(pB+ Tree). Since current computer systems can prefetch different data simultaneously, the authors take advantage of this. The idea here is to increase the node
+size of B+ Trees and employ prefectching at the same time. In a B+ Tree, every time we move down a level in the tree, a cache miss occurs. Accordingly,
+prefetching larger sizes nodes means that we can overlap multiple cache misses before they occur. This is because in one cache miss, extra data is retrieved in
+parallel with the originally requested data if it can be predicted sufficiently early. Figure~\ref{fig:prefetching} shows how prefetching retrieves the same
+data in less time. This method, however, does not eliminate the full cache miss latency with every level in the tree.
 
@@
 To resolve these issues, Lee et al.~\cite{lee2007cst} propose the Cache Sensitive T-Trees (CST-Trees). This is achieved in the following ways. First, a binary
-search tree whose node values are the maximum key of each node in the T Tre is constructed. That way, for a given value, the node containing this value could be
-quickly located through searching the binary tree. Second, the need to store pointers is eliminated by storing the tree in an array and locating the necessary
-nodes through index calculation. Finally, node sizes are aligned with cache line size such that there are no cache misses when accessing data in each binary
+search tree whose node values are the maximum key of each node in the T Tree is constructed. That way, for a given value, the node containing this value could
+be quickly located through searching the binary tree. Second, the need to store pointers is eliminated by storing the tree in an array and locating the
+necessary nodes through index calculation. Finally, node sizes are aligned with cache line size such that there are no cache misses when accessing data in each
+binary
 search tree in the array.
@@
 that the whole node fits in a cache line. This improves local searching within a node to just one cache miss. They also hard-code the traversal within a node
-such that finding the next node happens in $log_{m+1}n$ times rather than $log_{2}n$ times in a binary search tree. However, CSS Trees are only suitable for
-read environments since they eliminate nearly all child pointers, thus removing the support for incremental update.
+such that finding the next node happens in $log_{m+1}n$ times rather than $log_{2}n$ times in a binary search tree where $m$ is the number of keys per node.
+However, CSS Trees are only suitable for read environments since they eliminate nearly all child pointers, thus removing the support for incremental update.
 
@@
 
-\subsection{Partitioning}
+\subsubsection{Materialized Views}
+?????????????????????
+
+\subsubsection{Partitioning}
 
@@
 
-\subsection{Clustering}
+\subsubsection{Clustering}
 Clustering in disk based systems is important because sequential I/O access is cheaper than random or dispersed I/O access. This is not the case with main

### 2010-07-10 trunk/cs848-db/Paper.tex

@@
 database performance is likely to suffer, and users have to wait for a long time to get the results back. If we know these queries beforehand, it makes sense to
-simply store the results of the queries on disk instead of recomputing them each time. This is exactly what materialized views aim to do. By precalculating the
+simply store the results of the queries on disk instead of recomputing them each time. Oracle 8i (SOURCE) introduced materialized views to fit this purpose. By
+precalculating the
 results of a complex query, and storing them in a table on disk, the new table with the results is very likely to be much smaller than the original tables
@@
 
-cite~\cite{dewitt1984implementation}.
+In this section, we talk about three families of index structures used in MMDB: B Trees, T Trees, and Binary Search Trees. There are more recent index
+structures developed for MMDB such as J+ Trees~\cite{luan2009prefetching}, but which have not gained much popularity, and so we do not discuss them in our
+survey. Table~\ref{tab:mmdbindexsumm} summarizes the features of the different index structures discussed in this section with respect to MMDB.
+
+\begin{table}[t!]
+\centering
+\scalebox{0.7}{
+\begin{tabular}{p{2cm}|p{6cm}|p{6cm}}
+\hline
+ & \textbf{Features suitable for MMDB} & \textbf{Cache Consciousness}\\
+\hline\hline
+B Trees &
+\begin{itemize}\item Good storage utilization
+\item Reasonably quick searching
+\item Fast updating
+\end{itemize} 
+& \begin{itemize}\item Reasonable cache behavior if node fits in cache line\end{itemize}\\
+\hline
+B+ Trees & 
+\begin{itemize}
+\item Wastes space since all data is stored in the leaves
+\item Reasonably quick searching
+\item Fast updating
+\end{itemize} &
+\begin{itemize}
+\item Reasonable cache behavior if node fits in cache line\end{itemize}\\
+\hline
+CSB+ Trees & \begin{itemize}
+              \item Same features of B+ Trees
+             \end{itemize} &
+\begin{itemize}
+ \item Good cache performance due to improved locality \& lower tree height
+\end{itemize}\\
+\hline
+pB+ Trees & \begin{itemize}
+              \item Same features of B+ Trees
+             \end{itemize} &
+\begin{itemize}
+ \item Improves cache miss performance by prefetching more data in each cache miss
+\item Node size is also increased leading to the same benefits of CSB+ Trees
+\end{itemize}\\
+\hline
+T Tree &  \begin{itemize}
+              \item Contain pointers to data values instead of the values themselves leading to better space utilization
+\item Node size is bigger leading to previous mentioned advantages
+             \end{itemize} &
+\begin{itemize}
+ \item Cache behavior was not considered at time of design
+\end{itemize}\\
+\hline
+CST & \begin{itemize}
+       \item Fast traversal in $log_{m+1}n$
+      \end{itemize} &
+\begin{itemize}
+ \item Good cache behavior as number of key nodes per line is chosen according to cache line miss.
+\end{itemize}\\
+%\hline
+%J+ Tree & \begin{itemize}
+%       \item 
+%      \end{itemize} &
+%\begin{itemize}
+% \item 
+%\end{itemize}\\
+%\hline
+%pJ+ Tree & \begin{itemize}
+%       \item Fast traversal in $log_{m+1}n$
+%      \end{itemize} &
+%\begin{itemize}
+% \item
+%\end{itemize}\\
+%\hline
+\end{tabular}
+}
+\caption{Summary of index structures for Main Memory Databases (MMDB)}
+\label{tab:mmdbindexsumm}
+\end{table}
+
+%cite~\cite{dewitt1984implementation}.
 
@@
 that the whole node fits in a cache line. This improves local searching within a node to just one cache miss. They also hard-code the traversal within a node
-such that finding the next node happens in $log_{m+1}n$ times rather than $log_{2}n$ times in a binary search tree. However, CSS Trees are only suitable for
-read environments since they eliminate nearly all child pointers, thus removing the support for incremental update.
+such that finding the next node happens in $log_{m+1}n$ times rather than $log_{2}n$ times in a binary search tree where $m$ is the number of keys per node.
+However, CSS Trees are only suitable for read environments since they eliminate nearly all child pointers, thus removing the support for incremental update.
 
-\subsubsection*{J+-trees and pJ+-trees}
-\cite{luan2009prefetching}
+\subsubsection{Materialized Views}
+From our literature survey, it seems that materialized views are not used in MMDB. This makes sense since the problems materialized views try to address are
+not present in MMDB. That is, the cost of computing complicated join queries is much less in MMDB due to the low cost of accessing the data in memory.
+Accordingly, the maintenance cost of materialized views in MMDB will outweigh their minimal benefit.
 
-\subsection{Partitioning}
+\subsubsection{Partitioning}
 

### 2010-07-10 cs848-db/Paper.tex

@@
 Instead, pointers to the actual attribute values can be used. This, in turn, will reduce the size of the index which is preferable in this case.
-Table~\ref{tab:mmdbindexsumm} summarizes the features of the different index structures discussed in this section with respect to MMDB.
+
+In this section, we talk about three families of index structures used in MMDB: B Trees, T Trees, and Binary Search Trees. There are more recent index
+structures developed for MMDB such as J+ Trees~\cite{luan2009prefetching}, but which have not gained much popularity, and so we do not discuss them in our
+survey. Table~\ref{tab:mmdbindexsumm} summarizes the features of the different index structures discussed in this section with respect to MMDB.
 
@@
 \hline
+
 pB+ Trees & \begin{itemize}
@@
 \begin{itemize}
- \item Good cache behavior as number of key nodes per line is chosen according to cache line miss.
+ \item Good cache behavior as number of key nodes per line is chosen according to cache line miss
 \end{itemize}\\
 \hline
-J+ Tree & \begin{itemize}
-       \item 
+CSS Trees &  \begin{itemize}
+       \item Fast traversal in $log_{m+1}n$ ($m$ is the number of keys per node)
+\item Mainly suitable for read environments
       \end{itemize} &
 \begin{itemize}
- \item 
+ \item Good cache behavior since $m$ is chosen to fit in the cache line size
 \end{itemize}\\
-\hline
-pJ+ Tree & \begin{itemize}
-       \item Fast traversal in $log_{m+1}n$
-      \end{itemize} &
-\begin{itemize}
- \item
-\end{itemize}\\
-\hline
+
 \end{tabular}
@@
 MMDB~\cite{lehman1986study}. In disk based systems, the B+ Tree is more used than the regular B Tree. However, in main memory databases, the B Tree would be
-more suitable since there is no gain from keeping all the datGood storage utilizationa in the leaves. In a main memory system, this would only waste space
-without improving
-performance. In a B or B+ Tree, multiple keys are used to search within a node. This means that if the node fits in a cache line, this cache load can satisfy
-more than more comparison leading to better cache utilization~\cite{rao1999cache}. Therefore, B and B+ Trees in general have reasonable cache behavior.
+more suitable from a storage perspective since there is no gain from keeping all the data in the leaves. In a main memory system, this
+would only waste space without improving performance. On the other hand, the B+ Tree uses multiple keys to search within a node. This means that
+if the node fits in a cache line, this cache load can satisfy more than more comparison leading to better cache utilization~\cite{rao1999cache}. Therefore, B+
+Trees in general have reasonable cache behavior.
 
@@
 
-\subsubsection*{J+-trees and pJ+-trees}
-\cite{luan2009prefetching}
-
 \subsubsection{Materialized Views}
-?????????????????????
+From our literature survey, it seems that materialized views are not used in MMDB. This makes sense since the problems materialized views try to address are
+not present in MMDB. That is, the cost of computing complicated join queries is much less in MMDB due to the low cost of accessing the data in memory.
+Accordingly, the maintenance cost of materialized views in MMDB will outweigh their minimal benefit.
 
@@
 vertical partitioning or horizontal partitioning, as previously explained. In main memory databases, we do not have the issue of expensive disk access which
-makes partitioning in memory unnecessary as it will not save any costs. However, partionining is needed in the secondary storage of the data on
+makes partitioning in memory unnecessary as it will not save any costs. However, partitioning is needed in the secondary storage of the data on
 disk~\cite{gruenwald1990database}. Disk storage is still used as a backup of main memory databases. When a crash occurs, a reload of the database from disk or
 archive memory takes place. In order to avoid page faults caused by referencing data that is still being reloaded, the reloading process must be designed to be
-efficient in loading the important data such that the database users are not affected. Gruenwald and Eich~\cite{gruenwald1990database,gruenwald1990choosing}
+efficient in loading the important data such that the database users are not affected. An efficient MMDB reload takes into consideration the number of I/Os for
+reload and the number of main memory references during transaction processing. Gruenwald and Eich~\cite{gruenwald1990database,gruenwald1990choosing} show that
+horizontal partitioning is the most suitable when the database performs more modifications than tuple deletions and more selections that projections and joins.
+Single vertical partitioning is chosen in the other cases. They also show that if we just concentrate on the reload performance (i.e the number of I/Os for
+reload), then single vertical partitioning is always the best choice.
 

### 2010-07-10 trunk/cs848-db/Paper.tex
cleanup

@@
 \begin{itemize}
- \item Good cache behavior as number of key nodes per line is chosen according to cache line miss.
+ \item Good cache behavior as number of key nodes per line is chosen according to cache line miss
 \end{itemize}\\
-%\hline
-%J+ Tree & \begin{itemize}
-%       \item 
-%      \end{itemize} &
-%\begin{itemize}
-% \item 
-%\end{itemize}\\
-%\hline
-%pJ+ Tree & \begin{itemize}
-%       \item Fast traversal in $log_{m+1}n$
-%      \end{itemize} &
-%\begin{itemize}
-% \item
-%\end{itemize}\\
-%\hline
+\hline
+CSS Trees &  \begin{itemize}
+       \item Fast traversal in $log_{m+1}n$ ($m$ is the number of keys per node)
+\item Mainly suitable for read environments
+      \end{itemize} &
+\begin{itemize}
+ \item Good cache behavior since $m$ is chosen to fit in the cache line size
+\end{itemize}\\
+
 \end{tabular}
@@
 
+\cite{Cheung20011}
+ROLAP
+          o Data is stored in relational databases.
+          o support for multi-dimensional view of data is achieved thru star scheme
+          o Support extensions to SQL.
+          o Efficiently implement the multidimensional data model and operations.
+          o Ex: Microstrategy.
+MOLAP
+          o Directly store multidimensional data in special data structures (arrays).
+          o Implement the OLAP operations over these special data structures.
+          o Snowflake Schema. \cite{kimball2009data}
+          o Fact Constellations.
+          o Ex: Essbase (Arbor), statistical DBMSes.
+
+hierarchical data clusters \cite{karayannidis2008hierarchical}
+
 star schema
@@
 \cite{355309}
-
-\subsubsection{Data Cubes}
-\cite{Cheung20011}
-ROLAP
-          o Data is stored in relational databases.
-          o Support extensions to SQL.
-          o Efficiently implement the multidimensional data model and operations.
-          o Ex: Microstrategy.
-MOLAP
-          o Directly store multidimensional data in special data structures (arrays).
-          o Implement the OLAP operations over these special data structures.
-          o Star Schema.
-          o Snowflake Schema. \cite{kimball2009data}
-          o Fact Constellations.
-          o Ex: Essbase (Arbor), statistical DBMSes.
-
-hierarchical data clusters \cite{karayannidis2008hierarchical}
 

### 2010-07-10 cs848-db/Paper.tex

@@
 
-\subsection{Data Compression}
+\subsection{Other Physical Design Techniques}
 
-\subsection{Other Techniques}
-
-Striping, mirroring, denormalization.
+There are other physical design methods to improve the performance of the information systems~\cite{lightstone2007physical}. For example, data compression can
+be used to make more data fit into a fixed amount of disk space such that it can be access faster. Data striping is used to distribute data that is accessed
+together across multiple disks. This allows the data to be retrieved faster through parallelism. Database reliability is also improved through mirorring which
+involves duplicating the data on multiple disks. Finally, refining the global schema to reflect query and transaction requirements is also sometimes used. This
+is called denormalization.
 
@@
 Given the different physical design methods explained in the previous section, we now look at how these methods apply to different types of information systems.
-Mainly, how they apply to Disk-based relational database systems, main memory relational database systems, data warehouses, and XML databases.
+Mainly, how they apply to Disk-based relational database systems, main memory relational database systems, data warehouses, and XML databases. We mainly focus
+on the first four techniques discussed in Section~\ref{SEC-ELEMENTS} as these are the main design decisions involved in physical design.
 
@@
 
-\subsubsection{Type of Indexes Used}
+\subsubsection{Index Structures}
 
@@
 
-\subsubsection{Indexes}
+\subsubsection{Index Structures}
 \paragraph{Tree-based Indexes}

### 2010-07-10 cs848-db/Paper.tex

@@
 
-\subsection{Indexes}
+\subsection{Index Structures}
 
@@
 MMDB, longer tree structures are used since its cheap to access main memory. Another point is that data values do not need to be stored in the index itself
-since they will be stored in main memory anyways so there are no performance gains from storing them in the index itself.
+since they will be stored in main memory anyways so there are no performance gains from storing them in the index itself. Table~\ref{tab:mmdbsumm} summarizes
+the physical design elements of MMDBs.
+
+\begin{table}[t!]
+\centering
+\scalebox{0.7}{
+\begin{tabular}{p{3cm}|p{6cm}|p{6cm}}
+\hline
+ \textbf{Physical Design Element} & \textbf{Desired Features} & \textbf{Methods/Structures Used}\\
+\hline\hline
+Index Structures & \begin{itemize}
+                    \item No need to store actual values in index \item Larger node size that is aligned with cache line size for cache consciousness
+                   \end{itemize} & \begin{itemize}
+                                    \item B Tree
+\item B+ Tree
+\item CSB+ Tree
+\item pB+ Tree
+\item T Tree
+\item CST Tree
+\item CSS Tree
+                                   \end{itemize}\\
+\hline
+Materialized Views & \begin{itemize} \item Not needed since processing and memory access is cheap \end{itemize}& N/A\\
+\hline
+Partitioning & \begin{itemize} \item Only needed in secondary storage to speed up reloading in case of a crash \end{itemize} & \begin{itemize}
+                                                                                            \item Horizontal Partitioning
+\item Single Vertical Partitioning
+                                                                                           \end{itemize}\\
+\hline
+Clustering & \begin{itemize} \item Not needed since sequential random or dispersed main memory access is not more expensive than sequential access \end{itemize}
+& N/A
+
+\end{tabular}
+}
+\caption{Summary of physical design of Main Memory Databases (MMDB)}
+\label{tab:mmdbsumm}
+\end{table}
 
@@
 \hline
- & \textbf{Features suitable for MMDB} & \textbf{Cache Consciousness}\\
+ \textbf{Index Structure} & \textbf{Features suitable for MMDB} & \textbf{Cache Consciousness}\\
 \hline\hline
@@
 CST & \begin{itemize}
-       \item Fast traversal in $log_{m+1}n$
+       \item Binary search tree for maximum value of each node leads to faster search
+\item More space efficient by removing pointers \& using array storage. Index calculation used to allocate values
       \end{itemize} &
 \begin{itemize}
- \item Good cache behavior as number of key nodes per line is chosen according to cache line miss
+ \item Node size is aligned with cache line size to avoid misses
 \end{itemize}\\

### 2010-07-10 trunk/cs848-db/Paper.tex

@@
 
+This includes which indices to use, and in which order to access the tables~\cite{finkelstein1988physical} and so on.
+
 This survey presents the different physical design patterns that have been used in different types of information systems. The systems discussed in this survey
@@
 
-Relational Databases were first introduced by Codd~\cite{codd1970relational}
-In relational databases, users specify the actions they want to execute without having to worry about how these operations will actually be performed. This is
-where the role of physical design comes in. Physical design specifies the access paths used to get data from the tables. This includes which indices to use, and
-in which order to access the tables~\cite{finkelstein1988physical} and so on.
+Relational Databases were first introduced by Codd~\cite{codd1970relational}. In relational databases, users specify the actions they want to execute without
+having to worry about how these operations will actually be performed. This is where the role of physical design comes in. Physical design specifies the access
+paths used to get data from the tables.
+
+\subsubsection{Index Structures}
 
@@
 MMDB, longer tree structures are used since its cheap to access main memory. Another point is that data values do not need to be stored in the index itself
-since they will be stored in main memory anyways so there are no performance gains from storing them in the index itself.
+since they will be stored in main memory anyways so there are no performance gains from storing them in the index itself. Table~\ref{tab:mmdbsumm} summarizes
+the physical design elements of MMDBs.
 
-\subsubsection{Type of Indexes Used}
+\begin{table}[t!]
+\centering
+\scalebox{0.7}{
+\begin{tabular}{p{3cm}|p{6cm}|p{6cm}}
+\hline
+ \textbf{Physical Design Element} & \textbf{Desired Features} & \textbf{Methods/Structures Used}\\
+\hline\hline
+Index Structures & \begin{itemize}
+                    \item No need to store actual values in index \item Larger node size that is aligned with cache line size for cache consciousness
+                   \end{itemize} & \begin{itemize}
+                                    \item B Tree
+\item B+ Tree
+\item CSB+ Tree
+\item pB+ Tree
+\item T Tree
+\item CST Tree
+\item CSS Tree
+                                   \end{itemize}\\
+\hline
+Materialized Views & \begin{itemize} \item Not needed since processing and memory access is cheap \end{itemize}& N/A\\
+\hline
+Partitioning & \begin{itemize} \item Only needed in secondary storage to speed up reloading in case of a crash \end{itemize} & \begin{itemize}
+                                                                                            \item Horizontal Partitioning
+\item Single Vertical Partitioning
+                                                                                           \end{itemize}\\
+\hline
+Clustering & \begin{itemize} \item Not needed since sequential random or dispersed main memory access is not more expensive than sequential access \end{itemize}
+& N/A
+
+\end{tabular}
+}
+\caption{Summary of physical design of Main Memory Databases (MMDB)}
+\label{tab:mmdbsumm}
+\end{table}
+
+\subsubsection{Index Structures}
 

### 2010-07-12 trunk/cs848-db/Paper.tex

@@
 
+Besides indexing the actual nodes, path indexing is also common in XML databases since usually queries search for a specific path. This is done by clustering
+together all the IDs of nodes on a given path, and creating an index for them~\cite{milo1999index, arion2008path}.
+
 \subsubsection{Materialized Views}
 
-Since XPath query processing is expensive, materialized views come in handy in XML databases. Balmin et al.~\cite{balmin2004framework} propose a framework for
+The main idea behind using materialized views is to rewrite queries using these views so that they run more efficiently. Given the semi-structured nature of
+XML data, and the expressiveness of most XML querying languages, view selection for optimization becomes a more complicated
+problem~\cite{tang2009materialized}. Therefore, the choice of which views to materialize becomes more complicated in XML database. That is, of course, besides
+the difficulty of selecting which views to use for rewriting. However, this problem is not within the scope this survey. Since XPath query processing is
+expensive, materialized views come in handy in XML databases. Balmin et al.~\cite{balmin2004framework} propose a framework for
 using XPath materialized views for XML query processing. (TALK ABOUT IT MORE)
+
+\subsubsection{Partitioning}
+In XML-enabled databases, inlining during the shredding of the data such that children nodes that occur only once are stored with the parent node produces
+better partitioned data~\cite{tatarinov2002storing}. Inlining shreds a document into sets of tables according to the document schema which leads to better
+performance because queries tend to access less data. Such inlining is similar to horizontal partitioning~\cite{ramanath2003searching}.
 

### 2010-07-12 trunk/cs848-db/Paper.tex

@@
 
-The main idea behind using materialized views is to rewrite queries using these views so that they run more efficiently. Given the semi-structured nature of
+The main idea behind using materialized views is to rewrite queries using these views so that they run more efficiently. This is very important in XML
+Databases since XQuery and XPath queries are complicated~\cite{arion2007structured}. The semi-structured nature of
 XML data, and the expressiveness of most XML querying languages, view selection for optimization becomes a more complicated
-problem~\cite{tang2009materialized}. Therefore, the choice of which views to materialize becomes more complicated in XML database. That is, of course, besides
-the difficulty of selecting which views to use for rewriting. However, this problem is not within the scope this survey. Since XPath query processing is
-expensive, materialized views come in handy in XML databases. Balmin et al.~\cite{balmin2004framework} propose a framework for
-using XPath materialized views for XML query processing. (TALK ABOUT IT MORE)
+problem~\cite{tang2009materialized}. Therefore, the choice of which views to materialize becomes more complicated in XML database. Much work has been
+done on rewriting XML queries using views~\cite{arion2007structured, balmin2004framework, aouiche2006clustering, tang2008multiple}. However, this problem is not
+within the scope this survey.
 
@@
 better partitioned data~\cite{tatarinov2002storing}. Inlining shreds a document into sets of tables according to the document schema which leads to better
-performance because queries tend to access less data. Such inlining is similar to horizontal partitioning~\cite{ramanath2003searching}.
+performance because queries tend to access less data. Such inlining is similar to horizontal partitioning~\cite{ramanath2003searching}. Additionally, it is
+common to horizontally partition the mapped relational tables such that nodes of the same type are stored together~\cite{amer2002overview}.
+
+\subsubsection{Clustering}
+Clustering in XML databases follows the heuristic that sub-elements are likely to be queries with an element~\cite{jagadish2002timber}. Accordingly, elements
+and their subelements are clustered together. In general, storing the XML data in document order it thought to be the best technique for efficient querying.
+Additionally, Lian et al.~\cite{lian2004efficient} suggest clustering XML documents by structural similarity to produce more efficient access. This is done by
+representing the documents as a structure graph (s-graph), and then identifying clusters based on similarity using a clustering algorithm.
+Nayak~\cite{nayak2008fast} also proposes a similar clustering mechanism for XML data. Her technique is also based on structural similarity, but it is more
+focused on the level structure of the XML documents.
 

### 2010-07-12 trunk/cs848-db/Paper.tex

@@
 \usepackage{graphicx}
+\usepackage{amssymb}
 
 \begin{document}
-%
-% paper title
-% can use linebreaks \\ within to get better formatting as desired
+
 \title{A Survey of Physical Design Techniques of Information Systems}
-
-% author names and affiliations
-% use a multiple column layout for up to three different
-% affiliations
-%\author{\IEEEauthorblockN{Karim Ali, Sarah Nadi}
-%\IEEEauthorblockA{School of Computer Science\\
-%University of Waterloo\\
-%Email: {{karim, snadi}}@uwaterloo.ca}
-%}
 
@@
 
+MySQL mainly uses B Trees except for spatial data where the R Tree is used (http://dev.mysql.com/doc//refman/5.0/en/mysql-indexes.html)
+
+\sussubsection{Paritioning}
+
+MySQL only supports horizontal partitioning.
+
 \subsection{Main Memory Relational Database System (MMDB)}
@@
 structures developed for MMDB such as J+ Trees~\cite{luan2009prefetching}, but which have not gained much popularity, and so we do not discuss them in our
-survey. Table~\ref{tab:mmdbindexsumm} summarizes the features of the different index structures discussed in this section with respect to MMDB.
+survey. Table~\ref{tab:mmdbindexsumm} summarizes the features of the different index structures discussed in this section with respect to MMDB. The T Tree is
+the most used index for MMDB (E.g. MySQL CLuster). However, there are other indexes used as well. For example, IBM's solidDB uses Trie (or prefix tree) for its
+indexing. For their purposes, it served as a good index for main memory since it eliminates the need for many key comparisons~\cite{ibmsoliddb}.
 
@@
 necessary nodes through index calculation. Finally, node sizes are aligned with cache line size such that there are no cache misses when accessing data in each
-binary
-search tree in the array.
+binary search tree in the array.
 
@@
 in a paged file and associates unique node identifiers to the actual nodes, one index indexes elements and attributes, and the last index keeps track of word
-occurrences and is used by the fulltext search extensions.
+occurrences and is used by the fulltext search extensions. Oracle uses a similar technique for indexing its XML content where an XMLIndex table is created for
+XML entries~\cite{oracleXML}. This table contains a unique identifier for the XPath leading to the node, the row id of the table used to store the XML data,an
+order key that identifies the hierarchal positioning of the node, a locator key used for fragment extraction, and the text value of the node.
 
@@
 
+Two-Tier Storage DBMS (using disk based and main memory dbs)~\cite{eo2008two}.
+
 \section{Conclusions}
 \label{SEC-CONCL}
-
-% use section* for acknowledgement
-\section*{Acknowledgment}
-
-The authors would like to thank...
 

### 2010-07-13 trunk/cs848-db/Paper.tex

@@
 
+\date{}
 % make the title area
@@
 
-Database users write queries and updates using a "user-friendly" language such as SQL without having to worry about how the underlying data is stored. Physical
+Relational Database Systems have been the most popular type of Information Systems for decades. When we think of designing a database systems, we usually think
+of which data we need to keep track of and how will we properly organize it into meaningful tables. However, this is not the only phase in database design.
+All of our designed tables need to be physically stored on disk. How these tables will be stored and accessed greatly affects the database performance. Physical
 database design is concerned with the actual storage of data on disk. This means how the files are organized, and how they are accessed. Data is stored in the
 form of records in files which reside on a physical disk. In this sense, we try to find the most efficient way possible to store and access the data such that
-queries are answered and executed efficiently. This is reflected in designing proper indexing techniques, partitioning data for more efficient access, designing
-materialized views, data clustering, data compression, striping, mirroring and denormalization~\cite{lightstone2007physical}.
-
-This includes which indices to use, and in which order to access the tables~\cite{finkelstein1988physical} and so on.
+queries are answered and executed efficiently. In the traditional relational database system, this involved selecting proper indexing techniques,
+partitioning the data for more efficient access, materializing the results of some queries, clustering and compressing the data, and many other
+techniques~\cite{fikelstein1988physical,lightstone2007physical}. The physical design aspects of relational databases have been studied in much detail. However,
+over time, different types of Information Systems have emerged that have different characteristics. For example, most relational databases have been disk-based
+where the primary copy of the data resides on disk. This is as opposed to Main Memory Relational Databases (MMDB) where the primary copy of the data resides in
+main memory and no disk operations are needed to query the data. Such differences pose new requirements for the physical design of these systems.
 
 This survey presents the different physical design patterns that have been used in different types of information systems. The systems discussed in this survey
-are disk based relational databases, main memory relational databases, data warehouses, and XML databases. The aim of this survey is not to explain the details
-of every data structure used in physical design, but rather to compare how the different data structures have been used in these systems. This includes
-explaining how alterations or additions are made to data structures to fit the specific needs of every system. For a complete reference on the details of the
-data structures used in physical database design, please see (CITE HERE).
+are disk based relational databases (traditional database systems), main memory relational databases, data warehouses, and XML databases. The aim of this survey
+is not to explain the details of every data structure used in physical design, but rather to compare how the different data structures have been used in these
+systems. This includes explaining how alterations or additions are made to data structures to fit the specific needs of every system. For a complete reference
+on the details of the data structures used in physical database design, please see (CITE HERE). Using the different physical design methods used in disk based
+relational database systems, we explore how these methods apply to the other systems and where the differences and similarities lie.
 
-The rest of the survey is organized as follows. Section~\ref{SEC-ELEMENTS} first explains different elements of physical database design. For example, this
-includes explaining what indexes are and briefly mentioning what the different types of indexes used in information systems are. After explaining the different
-elements of physical database design, Section~\ref{SEC-DIFFSYS} discusses how each of the four systems uses these elements to improve its efficiency.
-Sections~\ref{SEC-DDRDB},~\ref{SEC-MMDB},~\ref{SEC-WAREHOUSES},~\ref{SEC-XML} discuss Disk Based Databases, Main Memory Databases, Data Warehouses, and XML
-Databases respectively. Throughout these sections, comparisons are made as to how the different elements have been modified to fit the needs of each system.
-Section~\ref{SEC-AUTO} then presents physical database design can be automated. Section~\ref{SEC-OPEN} mentions some of the open problems of physical database
-design, and suggests possible solutions. Finally Section~\ref{SEC-CONCL} summarizes and concludes this summary.
+The rest of the survey is organized as follows. Section~\ref{SEC-ELEMENTS} first explains different elements of physical database design based on traditional
+disk-based databases. Section~\ref{SEC-DIFFSYS} then discusses how each of the three other systems uses these elements, as applicable, to improve its
+efficiency.Sections~\ref{SEC-MMDB},~\ref{SEC-XML},~\ref{SEC-WAREHOUSES}, discuss Main Memory Databases, XML Databases, and Data Warehouses
+respectively. Throughout these sections, comparisons are made as to how the different elements have been modified to fit the needs of each system.
+Section~\ref{SEC-AUTO} then presents how the choice of the different physical design techniques can be automated. Section~\ref{SEC-OPEN} mentions some of the
+open problems of physical database design, and suggests possible solutions. Finally Section~\ref{SEC-CONCL} summarizes and concludes this survey.
 
@@
 \label{SEC-ELEMENTS}
+
+Relational Databases were first introduced by Codd~\cite{codd1970relational}. Disk based Relational Database Systems (RDBMS) are the most common and popular
+type of Information Systems. Since they have been extensively researched and have been around for many decades, we use them as the basis of our comparison. In
+this section, we describe the different elements of physical design in RDBMS. These elements will also be examined in the other systems to see how they apply
+to them.
 
@@
 B-trees (short for Balanced Trees) are the most used database index structure.  The B-tree is essentially similar to the traditional Binary Search Tree, but
-instead of having one value per node, a B-tee can have many values per node~\cite{comer1979ubiquitous} as shown in Figure (PUT A FIGURE FOR A B TREE HERE).
+instead of havinsummarizesg one value per node, a B-tee can have many values per node~\cite{comer1979ubiquitous} as shown in Figure (PUT A FIGURE FOR A B TREE
+HERE).
 B-trees are suitable for relational databases since the cost of retrieval in a B-tree is at most proportional to: $log_{d}\frac{n+1}{2}$. The cost of insertion
@@
 and will be discussed in details in Section~\ref{SEC-WAREHOUSES}.
+
+There are many variations of B-trees. In relational databases, the most popular index is the $B^{+}-tree$. The $B^{+}-tree$ is a variant of the B  tree which is
+becoming the main indexing method supported by many databases such as DB2, Oracle, and SQL Server~cite{lightstone2007physical}. The main difference between the
+$B^{+}-trees$ and the B-tree is that only leaf nodes contain data pointers in a $B^{+}-tree$. (COMPARE COST HERE). In Prefix $B^{+}-trees$, instead of using the
+actual key value of a record, a prefix from the key value is used to decrease the storage size needed which will in turn allow more records to be stored per
+node thus decreasing the height of the tree.
+
+The B+tree is the main indexing method used in current relational databases~\cite{lightstone2007physical}. B+trees have been successful in relational databases
+since data is ultimately recorded in files. Since B+trees have a high fanout, they allow less I/O operations to access a specific data record which is stored in
+a specific file.
+
+MySQL mainly uses B Trees except for spatial data where the R Tree is used (http://dev.mysql.com/doc//refman/5.0/en/mysql-indexes.html)
+
+Oracle Database uses B tree indexes and their subtypes such as Index-organized tables, reverse key indexes, descending indexes, B-tree cluster indexes. It also
+uses Bitmap and bitmap join indexes, function based indexes and domain based indexes.
+
+MySQL mostly uses B trees, except for spatial data where R trees are used and hash indexes are used for memory tables.
+
+mention clustered vs. unclustered ?
 
@@
 on materialized views since they have many queries with several joins. This will be discussed in details in Section~\ref{SEC-WAREHOUSES}.
+There are many variations of B-trees. In relational databases, the most popular index is the $B^{+}-tree$. The $B^{+}-tree$ is a variant of the B  tree which is
+becoming the main indexing method supported by many databases such as DB2, Oracle, and SQL Server~cite{lightstone2007physical}. The main difference between the
+$B^{+}-trees$ and the B-tree is that only leaf nodes contain data pointers in a $B^{+}-tree$. (COMPARE COST HERE). In Prefix $B^{+}-trees$, instead of using the
+actual key value of a record, a prefix from the key value is used to decrease the storage size needed which will in turn allow more records to be stored per
+node thus decreasing the height of the tree.
+
+The B+tree is the main indexing method used in current relational databases~\cite{lightstone2007physical}. B+trees have been successful in relational databases
+since data is ultimately recorded in files. Since B+trees have a high fanout, they allow less I/O operations to access a specific data record which is stored in
+a specific file.
 
@@
 
+Put table from database about the diff horizontal partitioning 
+MySQL only supports horizontal partitioning.
+
 \subsection{Clustering}
@@
 page, for example. In this case, only this page will need to be read. As
-people realized the potential of clustering, multidimensional clustering was introduced where data is clustered based on different criteria at the same time.
+people realized the potential of clustering, multidimensional clustering was introduced where data is clustered based on different criteria at the same time
+(TALK MORE).
 
@@
 Given the different physical design methods explained in the previous section, we now look at how these methods apply to different types of information systems.
-Mainly, how they apply to Disk-based relational database systems, main memory relational database systems, data warehouses, and XML databases. We mainly focus
+Mainly, how they apply to Main Memory Relational Database Systems (MMDB), Data Warehouses, and XML databases. We mainly focus
 on the first four techniques discussed in Section~\ref{SEC-ELEMENTS} as these are the main design decisions involved in physical design.
-
-\subsection{Disk Based Relational Database System (DRDB)}
-\label{SEC-DDRDB}
-
-Relational Databases were first introduced by Codd~\cite{codd1970relational}. In relational databases, users specify the actions they want to execute without
-having to worry about how these operations will actually be performed. This is where the role of physical design comes in. Physical design specifies the access
-paths used to get data from the tables.
-
-\subsubsection{Index Structures}
-
-There are many variations of B-trees. In relational databases, the most popular index is the $B^{+}-tree$. The $B^{+}-tree$ is a variant of the B  tree which is
-becoming the main indexing method supported by many databases such as DB2, Oracle, and SQL Server~cite{lightstone2007physical}. The main difference between the
-$B^{+}-trees$ and the B-tree is that only leaf nodes contain data pointers in a $B^{+}-tree$. (COMPARE COST HERE). In Prefix $B^{+}-trees$, instead of using the
-actual key value of a record, a prefix from the key value is used to decrease the storage size needed which will in turn allow more records to be stored per
-node thus decreasing the height of the tree.
-
-The B+tree is the main indexing method used in current relational databases~\cite{lightstone2007physical}. B+trees have been successful in relational databases
-since data is ultimately recorded in files. Since B+trees have a high fanout, they allow less I/O operations to access a specific data record which is stored in
-a specific file.
-
-MySQL mainly uses B Trees except for spatial data where the R Tree is used (http://dev.mysql.com/doc//refman/5.0/en/mysql-indexes.html)
-
-\sussubsection{Paritioning}
-
-MySQL only supports horizontal partitioning.
 
@@
 
+\subsection{XML Databases}
+\label{SEC-XML}
+
+XML (eXtensible Markup Language) is increasingly being used as a data exchange format. Accordingly, the ability to store and manage XML documents is needed.
+Salminen and Tompa~\cite{salminen2001requirements} define an XML database as ``a collection of XML documents and their parts, maintained by a system
+having capabilities to manage and control the collection itself and the information represented by that collection.'' Salminena and Tompa also highlight some of
+the requirements of an XML database system. These requirements include dealing with namespaces, Internet resources, querying parts of an XML document, and
+transformations of XML documents.
+
+There are two ways in which an XML database system can be designed. The first is to have the database internally translate the XML document into relational
+tables (XML-enabled databases). The second is to have data structures that can persistently store XML documents (native XML databases). In this section, we will
+discuss how the database's physical design needs to be adapted in both cases. Other challenges such as expressing and executing queries and updates are faced in
+XML database design, but these are beyond the scope of this survey.
+
+%eXist~\cite{meier2009exist} and TIMBER~\cite{jagadish2002timber} are two of the famous research efforts to design a native XML database.
+
+\subsubsection{Index Structures}
+
+Despite the special nature of XML documents, the same index structures used in relational databases can still be utilized in XML databases. However, some
+adjustments may need to be made to fit the nature of the XML documents. For XML-enabled databases, XML documents are usually represented as relational tables,
+and then indexed similar to other tables. Pal et al.~\cite{pal2004indexing} describe how this is done in Microsoft SQL Server 2005. To start with, a new data
+type called `XML' was introduced. This data type could contain values of complete XML documents, or just fragments of XML data. First, the different nodes in
+the XML data are labelled with the ORDPATH~\cite{o2004ordpaths} mechanism. The XML data is then shredded into a relational table with five main columns:ORDPATH,
+TAG, NODE\_TYPE, VALUE, and PATH\_ID. This information is then stored in a $B^{+}-tree$ as the primary index of the XML type. Additionally, a secondary index
+can be created on any of the columns of the primary index to speed things up.
+
+For native XML databases such as eXist~\cite{meier2009exist} or TIMBER~\cite{jagadish2002timber}, $B^{+}-trees$ were still used to index the XML documents.
+However, in contrast to XML-enabled databases, the XML data is not first stripped into a relational table. eXist uses a number schema to assign unique
+identifiers to each node in the XML document. These unique number identifiers allow the determination of any node's parent, sibling or possible child nodes.
+Four index files are created for XML data where all the indexes are based on $B^{+}-trees$. One index manages the collection hierarchy, one index collects nodes
+in a paged file and associates unique node identifiers to the actual nodes, one index indexes elements and attributes, and the last index keeps track of word
+occurrences and is used by the fulltext search extensions. Oracle uses a similar technique for indexing its XML content where an XMLIndex table is created for
+XML entries~\cite{oracleXML}. This table contains a unique identifier for the XPath leading to the node, the row id of the table used to store the XML data,an
+order key that identifies the hierarchal positioning of the node, a locator key used for fragment extraction, and the text value of the node.
+
+TIMBER also uses a similar numbering schema. There are minor differences between the two numbering techniques, but the main idea is the same. The numbering is
+based on three values of a node: its start label, its end label, and its level (i.e. nested depth). On the other hand, Sedna~\cite{taranov2010sedna} uses a
+different reasoning to produce its numbering schema. Sedna assignes string identifiers to nodes such that they are lexically ordered according to the position
+of the node. Despite the differences in the ways the numbering is constructed, it seems that that no novel indexing structure is needed for XML databases. The
+problem is rather how to assign the nodes unique identifiers (the numbers or labels in the numbering schema) that can be used as identifiers in the index to
+reflect the XML document structure.
+
+Besides indexing the actual nodes, path indexing is also common in XML databases since usually queries search for a specific path. This is done by clustering
+together all the IDs of nodes on a given path, and creating an index for them~\cite{milo1999index, arion2008path}.
+
+\subsubsection{Materialized Views}
+
+The main idea behind using materialized views is to rewrite queries using these views so that they run more efficiently. This is very important in XML
+Databases since XQuery and XPath queries are complicated~\cite{arion2007structured}. The semi-structured nature of
+XML data, and the expressiveness of most XML querying languages, view selection for optimization becomes a more complicated
+problem~\cite{tang2009materialized}. Therefore, the choice of which views to materialize becomes more complicated in XML database. Much work has been
+done on rewriting XML queries using views~\cite{arion2007structured, balmin2004framework, aouiche2006clustering, tang2008multiple}. However, this problem is not
+within the scope this survey.
+
+\subsubsection{Partitioning}
+In XML-enabled databases, inlining during the shredding of the data such that children nodes that occur only once are stored with the parent node produces
+better partitioned data~\cite{tatarinov2002storing}. Inlining shreds a document into sets of tables according to the document schema which leads to better
+performance because queries tend to access less data. Such inlining is similar to horizontal partitioning~\cite{ramanath2003searching}. Additionally, it is
+common to horizontally partition the mapped relational tables such that nodes of the same type are stored together~\cite{amer2002overview}.
+
+\subsubsection{Clustering}
+Clustering in XML databases follows the heuristic that sub-elements are likely to be queries with an element~\cite{jagadish2002timber}. Accordingly, elements
+and their subelements are clustered together. In general, storing the XML data in document order it thought to be the best technique for efficient querying.
+Additionally, Lian et al.~\cite{lian2004efficient} suggest clustering XML documents by structural similarity to produce more efficient access. This is done by
+representing the documents as a structure graph (s-graph), and then identifying clusters based on similarity using a clustering algorithm.
+Nayak~\cite{nayak2008fast} also proposes a similar clustering mechanism for XML data. Her technique is also based on structural similarity, but it is more
+focused on the level structure of the XML documents.
+
 \subsection{Data Warehouses}
@@
 
-\subsection{XML Databases}
-\label{SEC-XML}
-
-XML (eXtensible Markup Language) is increasingly being used as a data exchange format. Accordingly, the ability to store and manage XML documents is needed.
-Salminen and Tompa~\cite{salminen2001requirements} define an XML database as ``a collection of XML documents and their parts, maintained by a system
-having capabilities to manage and control the collection itself and the information represented by that collection.'' Salminena and Tompa also highlight some of
-the requirements of an XML database system. These requirements include dealing with namespaces, Internet resources, querying parts of an XML document, and
-transformations of XML documents.
-
-There are two ways in which an XML database system can be designed. The first is to have the database internally translate the XML document into relational
-tables (XML-enabled databases). The second is to have data structures that can persistently store XML documents (native XML databases). In this section, we will
-discuss how the database's physical design needs to be adapted in both cases. Other challenges such as expressing and executing queries and updates are faced in
-XML database design, but these are beyond the scope of this survey.
-
-%eXist~\cite{meier2009exist} and TIMBER~\cite{jagadish2002timber} are two of the famous research efforts to design a native XML database.
-
-\subsubsection{Index Structures}
-
-Despite the special nature of XML documents, the same index structures used in relational databases can still be utilized in XML databases. However, some
-adjustments may need to be made to fit the nature of the XML documents. For XML-enabled databases, XML documents are usually represented as relational tables,
-and then indexed similar to other tables. Pal et al.~\cite{pal2004indexing} describe how this is done in Microsoft SQL Server 2005. To start with, a new data
-type called `XML' was introduced. This data type could contain values of complete XML documents, or just fragments of XML data. First, the different nodes in
-the XML data are labelled with the ORDPATH~\cite{o2004ordpaths} mechanism. The XML data is then shredded into a relational table with five main columns:ORDPATH,
-TAG, NODE\_TYPE, VALUE, and PATH\_ID. This information is then stored in a $B^{+}-tree$ as the primary index of the XML type. Additionally, a secondary index
-can be created on any of the columns of the primary index to speed things up.
-
-For native XML databases such as eXist~\cite{meier2009exist} or TIMBER~\cite{jagadish2002timber}, $B^{+}-trees$ were still used to index the XML documents.
-However, in contrast to XML-enabled databases, the XML data is not first stripped into a relational table. eXist uses a number schema to assign unique
-identifiers to each node in the XML document. These unique number identifiers allow the determination of any node's parent, sibling or possible child nodes.
-Four index files are created for XML data where all the indexes are based on $B^{+}-trees$. One index manages the collection hierarchy, one index collects nodes
-in a paged file and associates unique node identifiers to the actual nodes, one index indexes elements and attributes, and the last index keeps track of word
-occurrences and is used by the fulltext search extensions. Oracle uses a similar technique for indexing its XML content where an XMLIndex table is created for
-XML entries~\cite{oracleXML}. This table contains a unique identifier for the XPath leading to the node, the row id of the table used to store the XML data,an
-order key that identifies the hierarchal positioning of the node, a locator key used for fragment extraction, and the text value of the node.
-
-TIMBER also uses a similar numbering schema. There are minor differences between the two numbering techniques, but the main idea is the same. The numbering is
-based on three values of a node: its start label, its end label, and its level (i.e. nested depth). On the other hand, Sedna~\cite{taranov2010sedna} uses a
-different reasoning to produce its numbering schema. Sedna assignes string identifiers to nodes such that they are lexically ordered according to the position
-of the node. Despite the differences in the ways the numbering is constructed, it seems that that no novel indexing structure is needed for XML databases. The
-problem is rather how to assign the nodes unique identifiers (the numbers or labels in the numbering schema) that can be used as identifiers in the index to
-reflect the XML document structure.
-
-Besides indexing the actual nodes, path indexing is also common in XML databases since usually queries search for a specific path. This is done by clustering
-together all the IDs of nodes on a given path, and creating an index for them~\cite{milo1999index, arion2008path}.
-
-\subsubsection{Materialized Views}
-
-The main idea behind using materialized views is to rewrite queries using these views so that they run more efficiently. This is very important in XML
-Databases since XQuery and XPath queries are complicated~\cite{arion2007structured}. The semi-structured nature of
-XML data, and the expressiveness of most XML querying languages, view selection for optimization becomes a more complicated
-problem~\cite{tang2009materialized}. Therefore, the choice of which views to materialize becomes more complicated in XML database. Much work has been
-done on rewriting XML queries using views~\cite{arion2007structured, balmin2004framework, aouiche2006clustering, tang2008multiple}. However, this problem is not
-within the scope this survey.
-
-\subsubsection{Partitioning}
-In XML-enabled databases, inlining during the shredding of the data such that children nodes that occur only once are stored with the parent node produces
-better partitioned data~\cite{tatarinov2002storing}. Inlining shreds a document into sets of tables according to the document schema which leads to better
-performance because queries tend to access less data. Such inlining is similar to horizontal partitioning~\cite{ramanath2003searching}. Additionally, it is
-common to horizontally partition the mapped relational tables such that nodes of the same type are stored together~\cite{amer2002overview}.
-
-\subsubsection{Clustering}
-Clustering in XML databases follows the heuristic that sub-elements are likely to be queries with an element~\cite{jagadish2002timber}. Accordingly, elements
-and their subelements are clustered together. In general, storing the XML data in document order it thought to be the best technique for efficient querying.
-Additionally, Lian et al.~\cite{lian2004efficient} suggest clustering XML documents by structural similarity to produce more efficient access. This is done by
-representing the documents as a structure graph (s-graph), and then identifying clusters based on similarity using a clustering algorithm.
-Nayak~\cite{nayak2008fast} also proposes a similar clustering mechanism for XML data. Her technique is also based on structural similarity, but it is more
-focused on the level structure of the XML documents.
-
-\section{Real Databases (incorporate in previous section)}
-
-Oracle Database uses B tree indexes and their subtypes such as Index-organized tables, reverse key indexes, descending indexes, B-tree cluster indexes. It also
-uses Bitmap and bitmap join indexes, function based indexes and domain based indexes.
-
-MySQL mostly uses B trees, except for spatial data where R trees are used and hash indexes are used for memory tables.
-
 \section{Automating Physical Database Design}
 \label{SEC-AUTO}
+
+Choosing which indexes to create, which views to materialize, and how to partition data are hard choices. They are highly dependent on the schema used as well
+as the query workload. The number of combinations of choices that can be made is very large making this task quite tedious and difficult for database
+administrators. Accordingly, researchers have been inclined to find ways to automate these decisions. There were many research efforts done in this
+area(E.g.~\cite{finkelstein1988physical, hammer76selec, frank1992selec}), but most of these efforts suffered from either examining too many configurations, not
+accounting for the query workload, or not considering if the query optimizer will benefit from the proposed configurations or not. Chaudhuri and
+Narasayya~\cite{chadhuri1997selec} were the first to propose an efficient automatic index selection tool that is dependent on the query workload and uses the
+query optimizer to evaluate the proposed configuration. They find the best configuration for the workload by examining each query individually to produce
+single-column indexes then combining these configurations. They then run a greedy algorithm to select the best configuration with the least cost (as determined
+by the query optimizer). They then try to further optimize the configuration by looking for possible multicolumn indexes as a combination of the single-column
+indexes already found. One of the main strengths of this technique was that all these indexes were virtually created such that the query optimizer could use
+them, but without having to incur the cost of physically creating them. This solution was part of the research at Microsoft and was tested in Microsoft SQL
+Server 7.0.
+
+Similarly, IBM~\cite{schiefer1999db2, valentin2000select} suggested a similar technique for their database with some changes such as selecting the order of
+columns within an index based on the order in which the query optimizer evaluates the predicates.
 
@@
 
-\bibliographystyle{plain}
+\bibliographystyle{IEEEtran}
 \bibliography{references}

### 2010-07-14 cs848-db/Paper.tex

@@
 
-Database users write queries and updates using a "user-friendly" language such as SQL without having to worry about how the underlying data is stored. Physical
+Relational Database Systems have been the most popular type of Information Systems for decades. When we think of designing a database systems, we usually think
+of which data we need to keep track of and how will we properly organize it into meaningful tables. However, this is not the only phase in database design.
+All of our designed tables need to be physically stored on disk. How these tables will be stored and accessed greatly affects the database performance. Physical
 database design is concerned with the actual storage of data on disk. This means how the files are organized, and how they are accessed. Data is stored in the
 form of records in files which reside on a physical disk. In this sense, we try to find the most efficient way possible to store and access the data such that
-queries are answered and executed efficiently. This is reflected in designing proper indexing techniques, partitioning data for more efficient access, designing
-materialized views, data clustering, data compression, striping, mirroring and denormalization~\cite{lightstone2007physical}.
+queries are answered and executed efficiently. In the traditional relational database system, this involved selecting proper indexing techniques,
+partitioning the data for more efficient access, materializing the results of some queries, clustering and compressing the data, and many other
+techniques~\cite{fikelstein1988physical,lightstone2007physical}. The physical design aspects of relational databases have been studied in much detail. However,
+over time, different types of Information Systems have emerged that have different characteristics. For example, most relational databases have been disk-based
+where the primary copy of the data resides on disk. This is as opposed to Main Memory Relational Databases (MMDB) where the primary copy of the data resides in
+main memory and no disk operations are needed to query the data. Such differences pose new requirements for the physical design of these systems.
 
 This survey presents the different physical design patterns that have been used in different types of information systems. The systems discussed in this survey
-are disk based relational databases, main memory relational databases, data warehouses, and XML databases. The aim of this survey is not to explain the details
-of every data structure used in physical design, but rather to compare how the different data structures have been used in these systems. This includes
-explaining how alterations or additions are made to data structures to fit the specific needs of every system. For a complete reference on the details of the
-data structures used in physical database design, please see (CITE HERE).
+are disk based relational databases (traditional database systems), main memory relational databases, data warehouses, and XML databases. The aim of this survey
+is not to explain the details of every data structure used in physical design, but rather to compare how the different data structures have been used in these
+systems. This includes explaining how alterations or additions are made to data structures to fit the specific needs of every system. For a complete reference
+on the details of the data structures used in physical database design, please see (CITE HERE). Using the different physical design methods used in disk based
+relational database systems, we explore how these methods apply to the other systems and where the differences and similarities lie.
 
-The rest of the survey is organized as follows. Section~\ref{SEC-ELEMENTS} first explains different elements of physical database design. For example, this
-includes explaining what indexes are and briefly mentioning what the different types of indexes used in information systems are. After explaining the different
-elements of physical database design, Section~\ref{SEC-DIFFSYS} discusses how each of the four systems uses these elements to improve its efficiency.
-Sections~\ref{SEC-DDRDB},~\ref{SEC-MMDB},~\ref{SEC-WAREHOUSES},~\ref{SEC-XML} discuss Disk Based Databases, Main Memory Databases, Data Warehouses, and XML
-Databases respectively. Throughout these sections, comparisons are made as to how the different elements have been modified to fit the needs of each system.
-Section~\ref{SEC-AUTO} then presents physical database design can be automated. Section~\ref{SEC-OPEN} mentions some of the open problems of physical database
-design, and suggests possible solutions. Finally Section~\ref{SEC-CONCL} summarizes and concludes this summary.
+The rest of the survey is organized as follows. Section~\ref{SEC-ELEMENTS} first explains different elements of physical database design based on traditional
+disk-based databases. Section~\ref{SEC-DIFFSYS} then discusses how each of the three other systems uses these elements, as applicable, to improve its
+efficiency.Sections~\ref{SEC-MMDB},~\ref{SEC-XML},~\ref{SEC-WAREHOUSES}, discuss Main Memory Databases, XML Databases, and Data Warehouses
+respectively. Throughout these sections, comparisons are made as to how the different elements have been modified to fit the needs of each system.
+Section~\ref{SEC-AUTO} then presents how the choice of the different physical design techniques can be automated. Section~\ref{SEC-OPEN} mentions some of the
+open problems of physical database design, and suggests possible solutions. Finally Section~\ref{SEC-CONCL} summarizes and concludes this survey.
 
@@
 Given the different physical design methods explained in the previous section, we now look at how these methods apply to different types of information systems.
-Mainly, how they apply to Disk-based relational database systems, main memory relational database systems, data warehouses, and XML databases. We mainly focus
+Mainly, how they apply to Main Memory Relational Database Systems (MMDB), Data Warehouses, and XML databases. We mainly focus
 on the first four techniques discussed in Section~\ref{SEC-ELEMENTS} as these are the main design decisions involved in physical design.
-
-\subsection{Disk Based Relational Database System (DRDB)}
-\label{SEC-DDRDB}
-
-Relational Databases were first introduced by Codd~\cite{codd1970relational}
-In relational databases, users specify the actions they want to execute without having to worry about how these operations will actually be performed. This is
-where the role of physical design comes in. Physical design specifies the access paths used to get data from the tables. This includes which indices to use, and
-in which order to access the tables~\cite{finkelstein1988physical} and so on.
-
-There are many variations of B-trees. In relational databases, the most popular index is the $B^{+}-tree$. The $B^{+}-tree$ is a variant of the B  tree which is
-becoming the main indexing method supported by many databases such as DB2, Oracle, and SQL Server~cite{lightstone2007physical}. The main difference between the
-$B^{+}-trees$ and the B-tree is that only leaf nodes contain data pointers in a $B^{+}-tree$. (COMPARE COST HERE). In Prefix $B^{+}-trees$, instead of using the
-actual key value of a record, a prefix from the key value is used to decrease the storage size needed which will in turn allow more records to be stored per
-node thus decreasing the height of the tree.
-
-The B+tree is the main indexing method used in current relational databases~\cite{lightstone2007physical}. B+trees have been successful in relational databases
-since data is ultimately recorded in files. Since B+trees have a high fanout, they allow less I/O operations to access a specific data record which is stored in
-a specific file.
 
@@
 without impacting performance.
+
+\subsection{XML Databases}
+\label{SEC-XML}
+
+XML (eXtensible Markup Language) is increasingly being used as a data exchange format. Accordingly, the ability to store and manage XML documents is needed.
+Salminen and Tompa~\cite{salminen2001requirements} define an XML database as ``a collection of XML documents and their parts, maintained by a system
+having capabilities to manage and control the collection itself and the information represented by that collection.'' Salminena and Tompa also highlight some of
+the requirements of an XML database system. These requirements include dealing with namespaces, Internet resources, querying parts of an XML document, and
+transformations of XML documents.
+
+There are two ways in which an XML database system can be designed. The first is to have the database internally translate the XML document into relational
+tables (XML-enabled databases). The second is to have data structures that can persistently store XML documents (native XML databases). In this section, we will
+discuss how the database's physical design needs to be adapted in both cases. Other challenges such as expressing and executing queries and updates are faced in
+XML database design, but these are beyond the scope of this survey.
+
+%eXist~\cite{meier2009exist} and TIMBER~\cite{jagadish2002timber} are two of the famous research efforts to design a native XML database.
+
+\subsubsection{Index Structures}
+
+Despite the special nature of XML documents, the same index structures used in relational databases can still be utilized in XML databases. However, some
+adjustments may need to be made to fit the nature of the XML documents. For XML-enabled databases, XML documents are usually represented as relational tables,
+and then indexed similar to other tables. Pal et al.~\cite{pal2004indexing} describe how this is done in Microsoft SQL Server 2005. To start with, a new data
+type called `XML' was introduced. This data type could contain values of complete XML documents, or just fragments of XML data. First, the different nodes in
+the XML data are labelled with the ORDPATH~\cite{o2004ordpaths} mechanism. The XML data is then shredded into a relational table with five main columns:ORDPATH,
+TAG, NODE\_TYPE, VALUE, and PATH\_ID. This information is then stored in a $B^{+}-tree$ as the primary index of the XML type. Additionally, a secondary index
+can be created on any of the columns of the primary index to speed things up.
+
+For native XML databases such as eXist~\cite{meier2009exist} or TIMBER~\cite{jagadish2002timber}, $B^{+}-trees$ were still used to index the XML documents.
+However, in contrast to XML-enabled databases, the XML data is not first stripped into a relational table. eXist uses a number schema to assign unique
+identifiers to each node in the XML document. These unique number identifiers allow the determination of any node's parent, sibling or possible child nodes.
+Four index files are created for XML data where all the indexes are based on $B^{+}-trees$. One index manages the collection hierarchy, one index collects nodes
+in a paged file and associates unique node identifiers to the actual nodes, one index indexes elements and attributes, and the last index keeps track of word
+occurrences and is used by the fulltext search extensions. Oracle uses a similar technique for indexing its XML content where an XMLIndex table is created for
+XML entries~\cite{oracleXML}. This table contains a unique identifier for the XPath leading to the node, the row id of the table used to store the XML data,an
+order key that identifies the hierarchal positioning of the node, a locator key used for fragment extraction, and the text value of the node.
+
+TIMBER also uses a similar numbering schema. There are minor differences between the two numbering techniques, but the main idea is the same. The numbering is
+based on three values of a node: its start label, its end label, and its level (i.e. nested depth). On the other hand, Sedna~\cite{taranov2010sedna} uses a
+different reasoning to produce its numbering schema. Sedna assignes string identifiers to nodes such that they are lexically ordered according to the position
+of the node. Despite the differences in the ways the numbering is constructed, it seems that that no novel indexing structure is needed for XML databases. The
+problem is rather how to assign the nodes unique identifiers (the numbers or labels in the numbering schema) that can be used as identifiers in the index to
+reflect the XML document structure.
+
+Besides indexing the actual nodes, path indexing is also common in XML databases since usually queries search for a specific path. This is done by clustering
+together all the IDs of nodes on a given path, and creating an index for them~\cite{milo1999index, arion2008path}.
+
+talk about Tamino (http://www.rpbourret.com/xml/XMLAndDatabases.htm)~\cite{bourret2003xml}
+
+\subsubsection{Materialized Views}
+
+The main idea behind using materialized views is to rewrite queries using these views so that they run more efficiently. This is very important in XML
+Databases since XQuery and XPath queries are complicated~\cite{arion2007structured}. The semi-structured nature of
+XML data, and the expressiveness of most XML querying languages, view selection for optimization becomes a more complicated
+problem~\cite{tang2009materialized}. Therefore, the choice of which views to materialize becomes more complicated in XML database. Much work has been
+done on rewriting XML queries using views~\cite{arion2007structured, balmin2004framework, aouiche2006clustering, tang2008multiple}. However, this problem is not
+within the scope this survey.
+
+\subsubsection{Partitioning}
+In XML-enabled databases, inlining during the shredding of the data such that children nodes that occur only once are stored with the parent node produces
+better partitioned data~\cite{tatarinov2002storing}. Inlining shreds a document into sets of tables according to the document schema which leads to better
+performance because queries tend to access less data. Such inlining is similar to horizontal partitioning~\cite{ramanath2003searching}. Additionally, it is
+common to horizontally partition the mapped relational tables such that nodes of the same type are stored together~\cite{amer2002overview}.
+
+\subsubsection{Clustering}
+Clustering in XML databases follows the heuristic that sub-elements are likely to be queries with an element~\cite{jagadish2002timber}. Accordingly, elements
+and their subelements are clustered together. In general, storing the XML data in document order it thought to be the best technique for efficient querying.
+Additionally, Lian et al.~\cite{lian2004efficient} suggest clustering XML documents by structural similarity to produce more efficient access. This is done by
+representing the documents as a structure graph (s-graph), and then identifying clusters based on similarity using a clustering algorithm.
+Nayak~\cite{nayak2008fast} also proposes a similar clustering mechanism for XML data. Her technique is also based on structural similarity, but it is more
+focused on the level structure of the XML documents.
 
@@
 
-\subsection{XML Databases}
-\label{SEC-XML}
-
-XML (eXtensible Markup Language) is increasingly being used as a data exchange format. Accordingly, the ability to store and manage XML documents is needed.
-Salminen and Tompa~\cite{salminen2001requirements} define an XML database as ``a collection of XML documents and their parts, maintained by a system
-having capabilities to manage and control the collection itself and the information represented by that collection.'' Salminena and Tompa also highlight some of
-the requirements of an XML database system. These requirements include dealing with namespaces, Internet resources, querying parts of an XML document, and
-transformations of XML documents.
-
-There are two ways in which an XML database system can be designed. The first is to have the database internally translate the XML document into relational
-tables (XML-enabled databases). The second is to have data structures that can persistently store XML documents (native XML databases). In this section, we will
-discuss how the database's physical design needs to be adapted in both cases. Other challenges such as expressing and executing queries and updates are faced in
-XML database design, but these are beyond the scope of this survey.
-
-%eXist~\cite{meier2009exist} and TIMBER~\cite{jagadish2002timber} are two of the famous research efforts to design a native XML database.
-
-\subsubsection{Index Structures}
-
-Despite the special nature of XML documents, the same index structures used in relational databases can still be utilized in XML databases. However, some
-adjustments may need to be made to fit the nature of the XML documents. For XML-enabled databases, XML documents are usually represented as relational tables,
-and then indexed similar to other tables. Pal et al.~\cite{pal2004indexing} describe how this is done in Microsoft SQL Server 2005. To start with, a new data
-type called `XML' was introduced. This data type could contain values of complete XML documents, or just fragments of XML data. First, the different nodes in
-the XML data are labelled with the ORDPATH~\cite{o2004ordpaths} mechanism. The XML data is then shredded into a relational table with five main columns:ORDPATH,
-TAG, NODE\_TYPE, VALUE, and PATH\_ID. This information is then stored in a $B^{+}-tree$ as the primary index of the XML type. Additionally, a secondary index
-can be created on any of the columns of the primary index to speed things up.
-
-For native XML databases such as eXist~\cite{meier2009exist} or TIMBER~\cite{jagadish2002timber}, $B^{+}-trees$ were still used to index the XML documents.
-However, in contrast to XML-enabled databases, the XML data is not first stripped into a relational table. eXist uses a number schema to assign unique
-identifiers to each node in the XML document. These unique number identifiers allow the determination of any node's parent, sibling or possible child nodes.
-Four index files are created for XML data where all the indexes are based on $B^{+}-trees$. One index manages the collection hierarchy, one index collects nodes
-in a paged file and associates unique node identifiers to the actual nodes, one index indexes elements and attributes, and the last index keeps track of word
-occurrences and is used by the fulltext search extensions.
-
-TIMBER also uses a similar numbering schema. There are minor differences between the two numbering techniques, but the main idea is the same. The numbering is
-based on three values of a node: its start label, its end label, and its level (i.e. nested depth). On the other hand, Sedna~\cite{taranov2010sedna} uses a
-different reasoning to produce its numbering schema. Sedna assignes string identifiers to nodes such that they are lexically ordered according to the position
-of the node. Despite the differences in the ways the numbering is constructed, it seems that that no novel indexing structure is needed for XML databases. The
-problem is rather how to assign the nodes unique identifiers (the numbers or labels in the numbering schema) that can be used as identifiers in the index to
-reflect the XML document structure.
-
-\subsubsection{Materialized Views}
-
-Since XPath query processing is expensive, materialized views come in handy in XML databases. Balmin et al.~\cite{balmin2004framework} propose a framework for
-using XPath materialized views for XML query processing. (TALK ABOUT IT MORE)
-
-\section{Real Databases (incorporate in previous section)}
-
-Oracle Database uses B tree indexes and their subtypes such as Index-organized tables, reverse key indexes, descending indexes, B-tree cluster indexes. It also
-uses Bitmap and bitmap join indexes, function based indexes and domain based indexes.
-
-MySQL mostly uses B trees, except for spatial data where R trees are used and hash indexes are used for memory tables.
-
 \section{Automating Physical Database Design}
 \label{SEC-AUTO}
+
+Choosing which indexes to create, which views to materialize, and how to partition data are hard choices. They are highly dependent on the schema used as well
+as the query workload. The number of combinations of choices that can be made is very large making this task quite tedious and difficult for database
+administrators. Accordingly, researchers have been inclined to find ways to automate these decisions. There were many research efforts done in this
+area(E.g.~\cite{finkelstein1988physical, hammer76selec, frank1992selec}), but most of these efforts suffered from either examining too many configurations, not
+accounting for the query workload, or not considering if the query optimizer will benefit from the proposed configurations or not. Chaudhuri and
+Narasayya~\cite{chadhuri1997selec} were the first to propose an efficient automatic index selection tool that is dependent on the query workload and uses the
+query optimizer to evaluate the proposed configuration. They find the best configuration for the workload by examining each query individually to produce
+single-column indexes then combining these configurations. They then run a greedy algorithm to select the best configuration with the least cost (as determined
+by the query optimizer). They then try to further optimize the configuration by looking for possible multicolumn indexes as a combination of the single-column
+indexes already found. One of the main strengths of this technique was that all these indexes were virtually created such that the query optimizer could use
+them, but without having to incur the cost of physically creating them. This solution was part of the research at Microsoft and was tested in Microsoft SQL
+Server 7.0.
+
+Similarly, IBM~\cite{schiefer1999db2, valentin2000select} suggested a similar technique for their database with some changes such as selecting the order of
+columns within an index based on the order in which the query optimizer evaluates the predicates.
 
@@
 
+Two-Tier Storage DBMS (using disk based and main memory dbs)~\cite{eo2008two}.
+
 \section{Conclusions}
@@
 
-% use section* for acknowledgement
-\section*{Acknowledgment}
-
-The authors would like to thank...
-
 %\bibliographystyle{IEEEtran}
 
-\bibliographystyle{plain}
+\bibliographystyle{IEEEtran}
 \bibliography{references}

### 2010-07-23 trunk/cs848-db/Paper.tex

@@
-%\documentclass[conference]{IEEEtran}
 \documentclass[12pt,a4paper]{article}
 \usepackage{fullpage}
+\usepackage{multirow}
 
@@
 \usepackage{amssymb}
+\usepackage{url}
 
@@
 
-Relational Database Systems have been the most popular type of Information Systems for decades. When we think of designing a database systems, we usually think
-of which data we need to keep track of and how will we properly organize it into meaningful tables. However, this is not the only phase in database design.
-All of our designed tables need to be physically stored on disk. How these tables will be stored and accessed greatly affects the database performance. Physical
-database design is concerned with the actual storage of data on disk. This means how the files are organized, and how they are accessed. Data is stored in the
+Relational Database Management Systems (RDBMS) have been the most popular type of Information Systems for decades. When we think of designing a database
+system, we usually think of which data we need to keep track of and how will we properly organize it into meaningful tables. However, this is not the only phase
+in database design. All of the designed tables need to be physically stored on disk. How these tables will be stored and accessed greatly affects the database
+performance. Physical database design is concerned with the actual storage of data on disk. This means how the files are organized, and how they are accessed.
+Data is stored in the
 form of records in files which reside on a physical disk. In this sense, we try to find the most efficient way possible to store and access the data such that
@@
 is not to explain the details of every data structure used in physical design, but rather to compare how the different data structures have been used in these
-systems. This includes explaining how alterations or additions are made to data structures to fit the specific needs of every system. For a complete reference
-on the details of the data structures used in physical database design, please see (CITE HERE). Using the different physical design methods used in disk based
-relational database systems, we explore how these methods apply to the other systems and where the differences and similarities lie.
+systems. This includes explaining how alterations or additions are made to data structures to fit the specific needs of every system. Using the different
+physical design methods used in disk based
+RDBMSs, we explore how these methods apply to the other systems and where the differences and similarities lie.
 
-The rest of the survey is organized as follows. Section~\ref{SEC-ELEMENTS} first explains different elements of physical database design based on traditional
-disk-based databases. Section~\ref{SEC-DIFFSYS} then discusses how each of the three other systems uses these elements, as applicable, to improve its
-efficiency.Sections~\ref{SEC-MMDB},~\ref{SEC-XML},~\ref{SEC-WAREHOUSES}, discuss Main Memory Databases, XML Databases, and Data Warehouses
+The rest of the survey is organized as follows. Section~\ref{SEC-ELEMENTS} first explains different elements of physical database design used in traditional
+disk-based databases. Section~\ref{SEC-DIFFSYS} then discusses how each of the three other systems uses these elements, as
+applicable. Sections~\ref{SEC-MMDB},~\ref{SEC-XML},~\ref{SEC-WAREHOUSES}, discuss Main Memory Databases, XML Databases, and Data Warehouses
 respectively. Throughout these sections, comparisons are made as to how the different elements have been modified to fit the needs of each system.
-Section~\ref{SEC-AUTO} then presents how the choice of the different physical design techniques can be automated. Section~\ref{SEC-OPEN} mentions some of the
+Section~\ref{SEC-OPEN} mentions some of the
 open problems of physical database design, and suggests possible solutions. Finally Section~\ref{SEC-CONCL} summarizes and concludes this survey.
@@
 
-Relational Databases were first introduced by Codd~\cite{codd1970relational}. Disk based Relational Database Systems (RDBMS) are the most common and popular
+Relational Databases were first introduced by Codd~\cite{codd1970relational}. Disk based Relational Database Management Systems (RDBMS) are the most common and
+popular
 type of Information Systems. Since they have been extensively researched and have been around for many decades, we use them as the basis of our comparison. In
@@
 
-B-trees (short for Balanced Trees) are the most used database index structure.  The B-tree is essentially similar to the traditional Binary Search Tree, but
-instead of havinsummarizesg one value per node, a B-tee can have many values per node~\cite{comer1979ubiquitous} as shown in Figure (PUT A FIGURE FOR A B TREE
-HERE).
-B-trees are suitable for relational databases since the cost of retrieval in a B-tree is at most proportional to: $log_{d}\frac{n+1}{2}$. The cost of insertion
+Btrees (short for Balanced Trees) are the most used database index structure.  The Btree is essentially similar to the traditional Binary Search Tree, but
+instead of having one value per node, a B-tee can have many values per node~\cite{comer1979ubiquitous}. B-trees are suitable for relational databases since the
+cost of retrieval in a B-tree is at most proportional to: $log_{d}\frac{n+1}{2}$. The cost of insertion
 and deletion is at most proportional to $log_{d}n$ due to the possibility of progressing back up the tree to balance it after an insertion or a deletion.
@@
 
-Different variations of B-trees have been used. For example, $B^{+}-trees$ are now the main method of indexing in Disk Based Relational Databases (DRDB). These
-will be discussed in Section~\ref{SEC-DDRDB}. The T Tree has been proposed for main memory databases, and will be discussed in Section~\ref{SEC-MMDB}. Cache
-sensitive indexes were then proposed. A variation of T Trees, Cache Sensitive T-Trees (CST-Trees) was introduced, and is also discussed in
-Section~\ref{SEC-MMDB}. On the same note, Cache Conscious $B^{+}-trees$ and Prefetching $J^{+}-Trees$ are also discussed in Section~\ref{SEC-MMDB}.
+Different variations of B-trees have been used. For example, B+trees are now the main method of indexing in current disk-based Relational
+Databases~\cite{lightstone2007physical} such as DB2, Oracle, and SQL Server. The main difference between the
+B+tree and the Btree is that only leaf nodes contain data pointers in a B+tree, and leaf nodes contain pointers to each other as shown in
+Figure~\ref{fig:b+tree}. In Prefix B+trees, instead of using the
+actual key value of a record, a prefix from the key value is used to decrease the storage size needed which will in turn allow more records to be stored per
+node thus decreasing the height of the tree. B+trees have been successful in relational databases since data is ultimately recorded in files. Since B+trees have
+a high fanout, they allow less I/O operations to access a specific data record which is stored in a specific file.
 
-Of course, there are other indexing techniques used such as hash tables, and bitmap indices. For example, Bitmap indices are commonly used in Data Warehouses,
-and will be discussed in details in Section~\ref{SEC-WAREHOUSES}.
+Of course, there are other indexing techniques used such as hash tables, and bitmap indices. EXPLAAAAAAAAIN
 
-There are many variations of B-trees. In relational databases, the most popular index is the $B^{+}-tree$. The $B^{+}-tree$ is a variant of the B  tree which is
-becoming the main indexing method supported by many databases such as DB2, Oracle, and SQL Server~cite{lightstone2007physical}. The main difference between the
-$B^{+}-trees$ and the B-tree is that only leaf nodes contain data pointers in a $B^{+}-tree$. (COMPARE COST HERE). In Prefix $B^{+}-trees$, instead of using the
-actual key value of a record, a prefix from the key value is used to decrease the storage size needed which will in turn allow more records to be stored per
-node thus decreasing the height of the tree.
+briefly discuss clustered vs. unclustered ?
 
-The B+tree is the main indexing method used in current relational databases~\cite{lightstone2007physical}. B+trees have been successful in relational databases
-since data is ultimately recorded in files. Since B+trees have a high fanout, they allow less I/O operations to access a specific data record which is stored in
-a specific file.
-
-MySQL mainly uses B Trees except for spatial data where the R Tree is used (http://dev.mysql.com/doc//refman/5.0/en/mysql-indexes.html)
-
-Oracle Database uses B tree indexes and their subtypes such as Index-organized tables, reverse key indexes, descending indexes, B-tree cluster indexes. It also
-uses Bitmap and bitmap join indexes, function based indexes and domain based indexes.
-
-MySQL mostly uses B trees, except for spatial data where R trees are used and hash indexes are used for memory tables.
-
-mention clustered vs. unclustered ?
+\begin{figure}[!t]
+\centering
+\includegraphics[width=7cm]{figs/b+tree.png}
+\caption{B+tree}
+\label{fig:b+tree}
+\end{figure}
 
@@
 database performance is likely to suffer, and users have to wait for a long time to get the results back. If we know these queries beforehand, it makes sense to
-simply store the results of the queries on disk instead of recomputing them each time. Oracle 8i (SOURCE) introduced materialized views to fit this purpose. By
+simply store the results of the queries on disk instead of recomputing them each time. Oracle 8i introduced materialized views to fit this purpose. By
 precalculating the
@@
 goldstein2001optimizing,
-abiteboul1998complexity}), but discussing the details of the rewriting is beyond the scope of this survey.
+abiteboul1998complexity}), but discussing the details of the rewriting process is beyond the scope of this survey.
 
-Despite the fact that materialized views provide a significant improvement in performance, we cannot simply create a materialized view for each common query. To
-begin with, materialized views consume disk storage which might be limited. The second, and main, problem with materialized views is their maintenance. Ensuring
+Despite the fact that materialized views provide a significant improvement in performance, we cannot simply create a materialized view for each frequent query.
+First, materialized views consume disk storage which might be limited. The second, and main, problem with materialized views is their maintenance.
+Ensuring
 that the views have the most up to date data is tricky, and may outset the benefit of using materialized views if it is not properly designed. Finally, having
-several materialized views may increase the cost of searching for the appropriate view to use during the query optimization stage. Data warehouses heavily rely
-on materialized views since they have many queries with several joins. This will be discussed in details in Section~\ref{SEC-WAREHOUSES}.
-There are many variations of B-trees. In relational databases, the most popular index is the $B^{+}-tree$. The $B^{+}-tree$ is a variant of the B  tree which is
-becoming the main indexing method supported by many databases such as DB2, Oracle, and SQL Server~cite{lightstone2007physical}. The main difference between the
-$B^{+}-trees$ and the B-tree is that only leaf nodes contain data pointers in a $B^{+}-tree$. (COMPARE COST HERE). In Prefix $B^{+}-trees$, instead of using the
-actual key value of a record, a prefix from the key value is used to decrease the storage size needed which will in turn allow more records to be stored per
-node thus decreasing the height of the tree.
-
-The B+tree is the main indexing method used in current relational databases~\cite{lightstone2007physical}. B+trees have been successful in relational databases
-since data is ultimately recorded in files. Since B+trees have a high fanout, they allow less I/O operations to access a specific data record which is stored in
-a specific file.
+several materialized views may increase the cost of searching for the appropriate view to use during the query optimization stage.
 
@@
 
-Partitioning is an important aspect of physical design as it reduces table scanning time. There are two main categories of partitioning: horizontal partitioning
+Partitioning is an important aspect of physical design as it reduces table scanning time~\cite{lightstone2007physical}. There are two main categories of
+partitioning: horizontal partitioning
 and vertical partitioning. Horizontal partitioning divides the table into sets of table rows where each row or record still has all the attributes of the table.
 For example, dividing the data by date where all data dating less than year 2000 lies in one partition while all data dating more lies in another. On the other
-hand, vertical partitioning reduces the width of the dataset by storing some attributes in one table, and some in another.
+hand, vertical partitioning reduces the width of the dataset by storing some attributes in one table, and some in another. Usually, less frequently queried
+columns are partitioned together to reduce the size of the table.
+
 The main advantage of any type of partitioning is to reduce the amount of time it takes to scan a table which in turn improves performance. For example, if a
 table has 1000 records, and a query is only using the first 20 records then having these 20 records in a partition containing 50 records will save time as
-opposed to examining all the 1000 records. Further classification in partitioning is single vs group horizontal or vertical
-partitioning~\cite{gruenwald1990database}. Horizontal partitioning just divides the data into partitions where each group contains complete tuples. Group
-horizontal partitioning groups tuples that are more frequently used together. Single vertical partitioning divides a relation into groups each of which contains
-only one type of attribute. Group vertical partitioning divides the data vertically such that attributes that are commonly required together are physically
-stored together.
+opposed to examining all the 1000 records. Horizontal partitioning is more widely used, and is implemented in all current DBMSes. Table~\ref{tab:horpartition}
+shows different categories of horizontal partitioning, and the situations they are suited to. Combining partitioning and indexing can often lead to better
+performance. For example, if List partitioning is used to divide a table by regions and then a bitmap index is used on that table, finding records in the same
+region or state can be much faster. Finally, partitioning the data into different physical devices to achieve parallelism when accessing it can greatly
+increase the performance of the database.
 
-Put table from database about the diff horizontal partitioning 
-MySQL only supports horizontal partitioning.
+NEED SOME REFERENCES HERE
+
+\begin{table}[!t]
+\scalebox{0.8}{
+ \begin{tabular}{p{2cm}||p{4cm}|p{4cm}|p{4cm}|p{4cm}}
+& \textbf{Range Partitioning} & \textbf{List Partitioning} & \textbf{Hash Partitioning} & \textbf{Composite Partitioning}\\
+\hline\hline
+\textbf{How?} & If key falls in range & List of keys & Hash function& Hybrid\\
+\hline
+\textbf{Why?} & Disk limits & Group point values & Range/List are N/A & According to needs\\
+\hline
+\textbf{Benefits} & \multicolumn{4}{c}{Partition elimination, Improved administration, Fast roll-in/out, and Implicit clustering }
+\\
+\hline
+\textbf{Example} & \begin{itemize}
+                    \item Zip code 
+		    \item Time range
+                   \end{itemize} &
+		    \begin{itemize}
+		    \item Regions
+		    \item States
+		    \end{itemize} &
+		    \begin{itemize}
+		    \item EmployeeId
+		    \item ProductId
+		    \end{itemize} & Hash then List \\
+\hline
+\textbf{RDBMS} & MySQL, DB2, Oracle, PostgreSQL, SQL Server & MySQL, Oracle, PostgreSQL, SQL Server & MySQL, Oracle & MySQL, Oracle\\
+
+\end{tabular}
+}
+\caption{Types of horizontal partitioning}
+\label{tab:horpartition}
+\end{table}
 
 \subsection{Clustering}
-When only records satisfying certain criteria are need from a table, it is often unnecessary to scan the whole table. For example, if only records that occurred
-in January 2010 are needed, then querying all the table would be an extremely unnecessary overhead. This is where clustering comes in. Clustering reflects how
-the records are physically located together on disk. Records that are likely to be queried together will be stored physically together to avoid querying the
-whole table. For the previously mentioned example, the data can be clustered by month, such that all the records of January 2010 are physically in the same
-page, for example. In this case, only this page will need to be read. As
-people realized the potential of clustering, multidimensional clustering was introduced where data is clustered based on different criteria at the same time
-(TALK MORE).
+Clustering reflects how the records are physically located together on disk. This is very important since sequential I/O access is much cheaper than random
+access. Therefore, records that are likely to be queried together should be stored physically together to decrease the number of random I/O accesses.
+MOREEEEEEEEEEEEEEEE
 
@@
 There are other physical design methods to improve the performance of the information systems~\cite{lightstone2007physical}. For example, data compression can
-be used to make more data fit into a fixed amount of disk space such that it can be access faster. Data striping is used to distribute data that is accessed
-together across multiple disks. This allows the data to be retrieved faster through parallelism. Database reliability is also improved through mirorring which
-involves duplicating the data on multiple disks. Finally, refining the global schema to reflect query and transaction requirements is also sometimes used. This
-is called denormalization.
+be used to make more data fit into a fixed amount of disk space such that it can be access faster. However, compression is only beneficial if the cost of
+executing the compression algorithm is lower than the I/O cost of accessing the data. Otherwise, compressing the data must just add an overhead without
+improving performance. Data striping is used to distribute data that is accessed together across multiple disks. This allows the data
+to be retrieved faster through parallelism. Striping has the same goals as partitioning. However, striping is more geared towards the hardware level
+which is beyond the control of the database while partitioning is within the control of the database. Database reliability is also improved through mirroring
+which involves duplicating the data on multiple disks. Finally, refining the global schema to reflect query and transaction requirements is also sometimes used.
+This is called denormalization which can be thought of as manual clustering of the data to improve performance.
 
@@
 
-A Main Memory Database (MMDB) (sometimes referred to as in-memory database) is one where the data resides in the main memory of the system rather than on a
+A Main Memory Database (MMDB) (sometimes referred to as in-memory database) is one where the primary copy of the data resides in the main memory of the system
+rather than on a
 disk~\cite{garcia1992main}. It should be noted that this is different from caching. Disk based database systems use the main memory to cache query results.
-However, MMDB's primary copy of the data resides in main memory. MMDB have many implementation challenges that have been addressed throughout the research
-community.
-
-Garcia-Molina and Salem~\cite{garcia1992main} mention some of the challenges involved. One is concurrency control. In disk based databases, locks are tracked
-through a hash table, but the actual objects on disk do not contain lock information. On the other hand, in MMDB, lock status is part of the object itself since
-it is cheap to keep a number of bits for that. The next challenge is access methods. In disk-based databases, B-Trees are used to index the data for faster
-access. B-Trees have a short bushy nature to try to decrease the height of the tree to decrease the number of I/O accesses. Since I/O access is not a problem in
-MMDB, longer tree structures are used since its cheap to access main memory. Another point is that data values do not need to be stored in the index itself
-since they will be stored in main memory anyways so there are no performance gains from storing them in the index itself. Table~\ref{tab:mmdbsumm} summarizes
+However, MMDB's primary copy of the data resides in main memory. Accessing data in main memory is faster than disk, and so MMDBs usually have better
+performance that disk based systems. However, main memory is more expensive and the size is limited in comparison to disk. However, MMDBs are essential to use
+where some of the collected data cannot, or will not, be stored on disk in the first place. For example, phone switch data may be needed for reasoning with
+other stored data, but it is captured in real time and will not necessarily be stored on disk. MMDB have many implementation challenges that have been addressed
+throughout the research community. These challenges will be discussed throughout this section. Table~\ref{tab:mmdbsumm} summarizes
 the physical design elements of MMDBs.
@@
 \hline
-Materialized Views & \begin{itemize} \item Not needed since processing and memory access is cheap \end{itemize}& N/A\\
+Materialized Views & \begin{itemize} \item Not highly utilized since processing and memory access is cheap \end{itemize}& N/A\\
 \hline
@@
                                                                                             \item Horizontal Partitioning
-\item Single Vertical Partitioning
+\item Vertical Partitioning
                                                                                            \end{itemize}\\
 \hline
-Clustering & \begin{itemize} \item Not needed since sequential random or dispersed main memory access is not more expensive than sequential access \end{itemize}
+Clustering & \begin{itemize} \item Not needed since random or dispersed main memory access is not more expensive than sequential access \end{itemize}
 & N/A
@@
 
-Two considerations are taken when designing index structures for MMDB. The first is that the data resides in main memory and not on disk, and so many of the
+Garcia-Molina and Salem~\cite{garcia1992main} mention data access methods as one of the main challenges involved in MMDBs. Two considerations are taken when
+designing index structures for MMDB. The first is that the data resides in main memory and not on disk, and so many of the
 considerations taken for I/O operations is no longer there. Instead of focusing on disk access and disk storage, the data structures for main memory
@@
 \hline
- \textbf{Index Structure} & \textbf{Features suitable for MMDB} & \textbf{Cache Consciousness}\\
+ \textbf{Index Structure} & \textbf{Features} & \textbf{Cache Consciousness}\\
 \hline\hline
@@
 \begin{itemize}
-\item Wastes space since all data is stored in the leaves
 \item Reasonably quick searching
@@
               \item Same features of B+ Trees
+\item Stores child nodes in an array. Stores pointer to first child only and uses index calculation to get the rest
              \end{itemize} &
 \begin{itemize}
- \item Good cache performance due to improved locality \& lower tree height
+ \item Good cache performance due to improved locality \& lower tree height (because of larger node size)
 \end{itemize}\\
@@
               \item Contain pointers to data values instead of the values themselves leading to better space utilization
-\item Node size is bigger leading to previous mentioned advantages
+\item Many sorted keys per node
+\item Only minimum and maximum keys used for comparisons
+\item Contains record pointers in every key
+\item Good update \& storage characteristics if Btrees.
+
              \end{itemize} &
@@
  \item Cache behavior was not considered at time of design
+\item Record pointers waste space in cache
 \end{itemize}\\
@@
  \item Node size is aligned with cache line size to avoid misses
+\item Higher usage of cached data in binary search tree
 \end{itemize}\\
@@
 if the node fits in a cache line, this cache load can satisfy more than more comparison leading to better cache utilization~\cite{rao1999cache}. Therefore, B+
-Trees in general have reasonable cache behavior.
+Trees in general have reasonable cache behavior, and variations of them have been adapted in MMDBs.
 
-However, the cache behavior of B+ Trees could still be improved. Therefore, Rao and Ross~\cite{rao2000making} propose Cache Sensitive B+-Trees
-($CSB^{+}-Trees$)~\cite{rao2000making} as an index structure for main memory. To do so, they eliminated most of the child pointers and had more keys in each
+However, the cache behavior of B+ Trees could still be improved. Therefore, Rao and Ross~\cite{rao2000making} propose the Cache Sensitive B+-Trees
+($CSB^{+}-Trees$)~\cite{rao2000making} as an index structure for main memory. To do so, they eliminated most of the child pointers by storing the child nodes
+in an array, and only keeping a pointed to the first child. Additionally, they used more keys in each
 node to improve locality and reduce tree height~\cite{luan2009prefetching}. Since the number of cache misses in search operations is proportional to the height
@@
 of searches, inserts and deletes. This is mainly because a T node contains many elements which results in good update and storage characteristics. The T Tree
-was the first index structure specifically designed for main memory databases.
-
-Although T Trees have been specifically designed for MMDB, they did not consider the cache behavior at that time~\cite{rao1999cache}. Although many keys are
-fit into one node, only the two end keys are used for comparison leading to low node utilization. Similarly, array binary search trees have the same problem.
+was the first index structure specifically designed for main memory databases. The T Tree has many sorted keys per node, and only uses the minimum and maximum
+keys for comparison. Additionally, every key contains a record pointer.
 
@@
 
-T Trees were the main index structure used for main memory databases for some time after their proposal. However, it was discovered that B+-trees outperform
+Although T Trees have been specifically designed for MMDB, they did not consider the cache behavior at that time since the gap between processor speed and main
+memory speed was not big then~\cite{rao1999cache}. Although many keys are
+fit into one node, only the two end keys are used for comparison leading to low node utilization. Similarly, array binary search trees have the same problem. 
+It was discovered that B+-trees outperform
 T-Trees on modern processors because of the growth of cache miss latency relative to processor speed~\cite{rao1999cache,lee2007cst}. This is because the height
 of the tree is high which makes the total number of memory accesses from the root to the leaf node higher. The other reason is that the node size is not aligned
-with the cache line size. Another problem with T-Trees is that there is a big waste of space in the cache because unnecessary data is brought into the cache.
-Additionally, record pointers stored in the tree take up a lot of space.
+with the cache line size. Another problem with T-Trees is that there is a big waste of space in the cache because unnecessary data which is not used (the
+record pointers) brought into the cache.
 
-To resolve these issues, Lee et al.~\cite{lee2007cst} propose the Cache Sensitive T-Trees (CST-Trees). This is achieved in the following ways. First, a binary
+To resolve these issues, Lee et al.~\cite{lee2007cst} propose the Cache Sensitive T-Trees (CST-Trees). First, a binary
 search tree whose node values are the maximum key of each node in the T Tree is constructed. That way, for a given value, the node containing this value could
 be quickly located through searching the binary tree. Second, the need to store pointers is eliminated by storing the tree in an array and locating the
-necessary nodes through index calculation. Finally, node sizes are aligned with cache line size such that there are no cache misses when accessing data in each
+necessary nodes through index calculation. This produces a better utilization of the cache by removing the unused data. Finally, node sizes are aligned with
+cache line size such that there are no cache misses when accessing data in each
 binary search tree in the array.
@@
 Tree (CSS-Tree) to improve this cache behavior. To do that, the search tree is stored as a directory in an array. The number of keys per node is chosen such
-that the whole node fits in a cache line. This improves local searching within a node to just one cache miss. They also hard-code the traversal within a node
+that the whole node fits in a cache line. This improves local searching within a node to just one cache miss. They also hard code the traversal within a node
 such that finding the next node happens in $log_{m+1}n$ times rather than $log_{2}n$ times in a binary search tree where $m$ is the number of keys per node.
@@
 not present in MMDB. That is, the cost of computing complicated join queries is much less in MMDB due to the low cost of accessing the data in memory.
-Accordingly, the maintenance cost of materialized views in MMDB will outweigh their minimal benefit.
+Accordingly, the maintenance cost of materialized views in MMDB will outweigh their minimal benefit. However, this does not mean that they cannot be used at
+all in MMDBs. As with all other physical design decisions, it rather depend on the benefits of implementing the technique. Accordingly, if the data is very
+large and may benefit from using materialized views, there is no reason not to use them.
 
@@
 horizontal partitioning is the most suitable when the database performs more modifications than tuple deletions and more selections that projections and joins.
-Single vertical partitioning is chosen in the other cases. They also show that if we just concentrate on the reload performance (i.e the number of I/Os for
-reload), then single vertical partitioning is always the best choice.
+Vertical partitioning is chosen in the other cases. They also show that if we just concentrate on the reload performance (i.e the number of I/Os for
+reload), then vertical partitioning is always the best choice.
 
@@
 having capabilities to manage and control the collection itself and the information represented by that collection.'' Salminena and Tompa also highlight some of
-the requirements of an XML database system. These requirements include dealing with namespaces, Internet resources, querying parts of an XML document, and
-transformations of XML documents.
+the requirements of an XML database system. These requirements include querying parts of an XML document, and
+transformations of XML documents. The main challenge in XML databases is path navigation since it is a very expensive operation.
 
 There are two ways in which an XML database system can be designed. The first is to have the database internally translate the XML document into relational
-tables (XML-enabled databases). The second is to have data structures that can persistently store XML documents (native XML databases). In this section, we will
-discuss how the database's physical design needs to be adapted in both cases. Other challenges such as expressing and executing queries and updates are faced in
-XML database design, but these are beyond the scope of this survey.
+tables (XML-enabled databases). The second is to have data structures that can persistently store XML documents (native XML databases). XML-enabled databases
+are suited for data-centric documents which have a regular structure without any mixed content~\cite{bourret2003xml}. On the other hand, for document-centric
+documents which have an irregular structure with lots of mixed content, having a native XML database is essential. Examples of XML-enabled databases include SQL
+Server 2005 and Oracle, and examples of native XML databases include Tamino~\cite{taminodb},
+eXist~\cite{meier2009exist}, and TIMBER~\cite{jagadish2002timber}.
 
-%eXist~\cite{meier2009exist} and TIMBER~\cite{jagadish2002timber} are two of the famous research efforts to design a native XML database.
+In this section, we will discuss how the database's physical design needs to be adapted in both cases. Other challenges such as expressing and executing queries
+and updates are faced in XML database design, but these are beyond the scope of this survey.
 
@@
 type called `XML' was introduced. This data type could contain values of complete XML documents, or just fragments of XML data. First, the different nodes in
-the XML data are labelled with the ORDPATH~\cite{o2004ordpaths} mechanism. The XML data is then shredded into a relational table with five main columns:ORDPATH,
+the XML data are labeled with the ORDPATH~\cite{o2004ordpaths} mechanism. The XML data is then shredded into a relational table with five main columns: ORDPATH,
 TAG, NODE\_TYPE, VALUE, and PATH\_ID. This information is then stored in a $B^{+}-tree$ as the primary index of the XML type. Additionally, a secondary index
@@
 
-For native XML databases such as eXist~\cite{meier2009exist} or TIMBER~\cite{jagadish2002timber}, $B^{+}-trees$ were still used to index the XML documents.
+For native XML databases such as eXist~\cite{meier2009exist} or TIMBER~\cite{jagadish2002timber}, B+trees are still used to index the XML documents.
 However, in contrast to XML-enabled databases, the XML data is not first stripped into a relational table. eXist uses a number schema to assign unique
@@
 in a paged file and associates unique node identifiers to the actual nodes, one index indexes elements and attributes, and the last index keeps track of word
-occurrences and is used by the fulltext search extensions. Oracle uses a similar technique for indexing its XML content where an XMLIndex table is created for
+occurrences and is used by the full text search extensions. Oracle uses a similar technique for indexing its XML content where an XMLIndex table is created for
 XML entries~\cite{oracleXML}. This table contains a unique identifier for the XPath leading to the node, the row id of the table used to store the XML data,an
-order key that identifies the hierarchal positioning of the node, a locator key used for fragment extraction, and the text value of the node.
+order key that identifies the hierarchal positioning of the node, a locater key used for fragment extraction, and the text value of the node.
 
@@
 based on three values of a node: its start label, its end label, and its level (i.e. nested depth). On the other hand, Sedna~\cite{taranov2010sedna} uses a
-different reasoning to produce its numbering schema. Sedna assignes string identifiers to nodes such that they are lexically ordered according to the position
-of the node. Despite the differences in the ways the numbering is constructed, it seems that that no novel indexing structure is needed for XML databases. The
-problem is rather how to assign the nodes unique identifiers (the numbers or labels in the numbering schema) that can be used as identifiers in the index to
-reflect the XML document structure.
+different reasoning to produce its numbering schema. Sedna assigns string identifiers to nodes such that they are lexically ordered according to the position
+of the node. Despite the differences in the ways the numbering is constructed, it seems that existing indexing techniques can be adapted for XML databases. The
+challenge is how to assign the nodes unique identifiers (the numbers or labels in the numbering schema) that can be used as identifiers in the index to
+reflect the XML document structure. Additionally, in native XML databases, documents are usually grouped into collections, and an index may be used to search
+within these collections. This is different from the node index which searches within a document.
 
-Besides indexing the actual nodes, path indexing is also common in XML databases since usually queries search for a specific path. This is done by clustering
-together all the IDs of nodes on a given path, and creating an index for them~\cite{milo1999index, arion2008path}.
+Besides indexing the actual nodes, path indexing is also common in XML databases since usually queries search for a specific path which can be very expensive.
+This is done by clustering together all the IDs of nodes on a given path, and creating an index for them~\cite{milo1999index, arion2008path}. Other techniques
+include that proposed by Cooper et al.~\cite{cooper2001} where they propose an indexing mechanism for paths called ``Index Fabric''. An Index Fabric is based on
+Patricia tries~\cite{knuth1998} which are used for string indexing, but is customized for disk-based systems. A path in an XML document is encoded into a unique
+string, and is then inserted into the fabric. The encoded string is usually very short, and so searching is relatively fast.
 
@@
 The main idea behind using materialized views is to rewrite queries using these views so that they run more efficiently. This is very important in XML
-Databases since XQuery and XPath queries are complicated~\cite{arion2007structured}. The semi-structured nature of
+Databases since XQuery and XPath queries are complicated~\cite{arion2007structured}. Accordingly, materialized views are used in XML databases to improve
+performance. The semi-structured nature of
 XML data, and the expressiveness of most XML querying languages, view selection for optimization becomes a more complicated
@@
 
-hierarchical data clusters \cite{karayannidis2008hierarchical}
+hierarchical data clusters \cite{karayannidis2008hierarchical}.
 
@@
 
-\section{Automating Physical Database Design}
-\label{SEC-AUTO}
-
-Choosing which indexes to create, which views to materialize, and how to partition data are hard choices. They are highly dependent on the schema used as well
-as the query workload. The number of combinations of choices that can be made is very large making this task quite tedious and difficult for database
-administrators. Accordingly, researchers have been inclined to find ways to automate these decisions. There were many research efforts done in this
-area(E.g.~\cite{finkelstein1988physical, hammer76selec, frank1992selec}), but most of these efforts suffered from either examining too many configurations, not
-accounting for the query workload, or not considering if the query optimizer will benefit from the proposed configurations or not. Chaudhuri and
-Narasayya~\cite{chadhuri1997selec} were the first to propose an efficient automatic index selection tool that is dependent on the query workload and uses the
-query optimizer to evaluate the proposed configuration. They find the best configuration for the workload by examining each query individually to produce
-single-column indexes then combining these configurations. They then run a greedy algorithm to select the best configuration with the least cost (as determined
-by the query optimizer). They then try to further optimize the configuration by looking for possible multicolumn indexes as a combination of the single-column
-indexes already found. One of the main strengths of this technique was that all these indexes were virtually created such that the query optimizer could use
-them, but without having to incur the cost of physically creating them. This solution was part of the research at Microsoft and was tested in Microsoft SQL
-Server 7.0.
-
-Similarly, IBM~\cite{schiefer1999db2, valentin2000select} suggested a similar technique for their database with some changes such as selecting the order of
-columns within an index based on the order in which the query optimizer evaluates the predicates.
-
 \section{Open Problems in Physical Database Design}

### 2010-07-24 trunk/cs848-db/Paper.tex

@@
 
-\section{Open Problems in Physical Database Design}
+\section{Open Research Problems in Physical Database Design}
 \label{SEC-OPEN}
+
+Choosing which indexes to create, which views to materialize, and how to partition data are hard choices. They are highly dependent on the schema used as well
+as the query workload. The number of combinations of choices that can be made is very large making this task quite tedious and difficult for database
+administrators. Accordingly, researchers have been inclined to find ways to automate these decisions. There were many early research efforts done in this
+area(E.g.~\cite{finkelstein1988physical, hammer76selec, frank1992selec, chaudhuri1997overview, schiefer1999db2, valentin2000select}), but most of these efforts
+focused on having a static workload. Accordingly, research shifted to more dynamic ways to automate physical design. Initial attempts used an dynamic, on-line
+approaches(E.g.~\cite{bruno2007online, sattler2003, schnaitter2007}) where they observed the changes in the workload and adjusted the design accordingly while
+attempting to predict future behavior based on the observed behavior. To avoid the problems of on-line design, researchers are also currently looking at how to
+perform this dynamic physical design off-line (E.g. ~\cite{hannes2008,agrawal2006}).
+
+A second question is how to characterize
+scenarios or classes of workloads for which constrained dynamic
+physical designs will be beneficial.
+
+Similarly, IBM~\cite{} suggested a similar technique for their database with some changes such as selecting the order of
+columns within an index based on the order in which the query optimizer evaluates the predicates.
 

### 2010-07-24 cs848-db/Paper.tex

@@
-%\documentclass[conference]{IEEEtran}
 \documentclass[12pt,a4paper]{article}
@@
 \usepackage{amssymb}
+\usepackage{url}
+\usepackage{multirow}
 
@@
 partitioning the data for more efficient access, materializing the results of some queries, clustering and compressing the data, and many other
-techniques~\cite{fikelstein1988physical,lightstone2007physical}. The physical design aspects of relational databases have been studied in much detail. However,
+techniques~\cite{finkelstein1988physical,lightstone2007physical}. The physical design aspects of relational databases have been studied in much detail. However,
 over time, different types of Information Systems have emerged that have different characteristics. For example, most relational databases have been disk-based
@@
 is not to explain the details of every data structure used in physical design, but rather to compare how the different data structures have been used in these
-systems. This includes explaining how alterations or additions are made to data structures to fit the specific needs of every system. For a complete reference
-on the details of the data structures used in physical database design, please see (CITE HERE). Using the different physical design methods used in disk based
-relational database systems, we explore how these methods apply to the other systems and where the differences and similarities lie.
+systems. This includes explaining how alterations or additions are made to data structures to fit the specific needs of every system. Using the different
+physical design methods used in disk based
+RDBMSs, we explore how these methods apply to the other systems and where the differences and similarities lie.
 
-The rest of the survey is organized as follows. Section~\ref{SEC-ELEMENTS} first explains different elements of physical database design based on traditional
-disk-based databases. Section~\ref{SEC-DIFFSYS} then discusses how each of the three other systems uses these elements, as applicable, to improve its
-efficiency.Sections~\ref{SEC-MMDB},~\ref{SEC-XML},~\ref{SEC-WAREHOUSES}, discuss Main Memory Databases, XML Databases, and Data Warehouses
+The rest of the survey is organized as follows. Section~\ref{SEC-ELEMENTS} first explains different elements of physical database design used in traditional
+disk-based databases. Section~\ref{SEC-DIFFSYS} then discusses how each of the three other systems uses these elements, as
+applicable. Sections~\ref{SEC-MMDB},~\ref{SEC-XML},~\ref{SEC-WAREHOUSES}, discuss Main Memory Databases, XML Databases, and Data Warehouses
 respectively. Throughout these sections, comparisons are made as to how the different elements have been modified to fit the needs of each system.
-Section~\ref{SEC-AUTO} then presents how the choice of the different physical design techniques can be automated. Section~\ref{SEC-OPEN} mentions some of the
-open problems of physical database design, and suggests possible solutions. Finally Section~\ref{SEC-CONCL} summarizes and concludes this survey.
+Section~\ref{SEC-OPEN} mentions some of the current
+open problems of physical database design. Finally Section~\ref{SEC-CONCL} summarizes and concludes this survey.
 
@@
 
-Partitioning is an important aspect of physical design as it reduces table scanning time. There are two main categories of partitioning: horizontal partitioning
+Partitioning is an important aspect of physical design as it reduces table scanning time~\cite{lightstone2007physical}. There are two main categories of
+partitioning: horizontal partitioning
 and vertical partitioning. Horizontal partitioning divides the table into sets of table rows where each row or record still has all the attributes of the table.
 For example, dividing the data by date where all data dating less than year 2000 lies in one partition while all data dating more lies in another. On the other
-hand, vertical partitioning reduces the width of the dataset by storing some attributes in one table, and some in another.
+hand, vertical partitioning reduces the width of the dataset by storing some attributes in one table, and some in another. Usually, less frequently queried
+columns are partitioned together to reduce the size of the table.
+
 The main advantage of any type of partitioning is to reduce the amount of time it takes to scan a table which in turn improves performance. For example, if a
 table has 1000 records, and a query is only using the first 20 records then having these 20 records in a partition containing 50 records will save time as
-opposed to examining all the 1000 records. Further classification in partitioning is single vs group horizontal or vertical
-partitioning~\cite{gruenwald1990database}. Horizontal partitioning just divides the data into partitions where each group contains complete tuples. Group
-horizontal partitioning groups tuples that are more frequently used together. Single vertical partitioning divides a relation into groups each of which contains
-only one type of attribute. Group vertical partitioning divides the data vertically such that attributes that are commonly required together are physically
-stored together.
+opposed to examining all the 1000 records. Horizontal partitioning is more widely used, and is implemented in all current DBMSes.
 
-Put table from database about the diff horizontal partitioning 
-MySQL only supports horizontal partitioning.
+%Table~\ref{tab:horpartition}
+shows different categories of horizontal partitioning, and the situations they are suited to. Combining partitioning and indexing can often lead to better
+performance. For example, if List partitioning is used to divide a table by regions and then a bitmap index is used on that table, finding records in the same
+region or state can be much faster. Finally, partitioning the data into different physical devices to achieve parallelism when accessing it can greatly
+increase the performance of the database.
+
+NEED SOME REFERENCES HERE
+
+\begin{table}[!t]
+\scalebox{0.7}{
+ \begin{tabular}{p{2cm}||p{4cm}|p{4cm}|p{4cm}|p{4cm}}
+& \textbf{Range Partitioning} & \textbf{List Partitioning} & \textbf{Hash Partitioning} & \textbf{Composite Partitioning}\\
+\hline\hline
+\textbf{How?} & If key falls in range & List of keys & Hash function& Hybrid\\
+\hline
+\textbf{Why?} & Disk limits & Group point values & Range/List are N/A & According to needs\\
+\hline
+\textbf{Benefits} & \multicolumn{4}{c}{Partition elimination, Improved administration, Fast roll-in/out, and Implicit clustering }\\
+\hline
+\textbf{Example} & \begin{itemize}
+                    \item Zip code 
+		    \item Time range
+                   \end{itemize} &
+		    \begin{itemize}
+		    \item Regions
+		    \item States
+		    \end{itemize} &
+		    \begin{itemize}
+		    \item EmployeeId
+		    \item ProductId
+		    \end{itemize} &
+		    Hash then List \\
+\hline
+\textbf{RDBMS} & MySQL, DB2, Oracle, PostgreSQL, SQL Server & MySQL, Oracle, PostgreSQL, SQL Server & MySQL, Oracle & MySQL, Oracle\\
+\end{tabular}
+}
+\caption{Types of horizontal partitioning}
+\label{tab:horpartition}
+\end{table}
 
 \subsection{Clustering}
-When only records satisfying certain criteria are need from a table, it is often unnecessary to scan the whole table. For example, if only records that occurred
-in January 2010 are needed, then querying all the table would be an extremely unnecessary overhead. This is where clustering comes in. Clustering reflects how
-the records are physically located together on disk. Records that are likely to be queried together will be stored physically together to avoid querying the
-whole table. For the previously mentioned example, the data can be clustered by month, such that all the records of January 2010 are physically in the same
-page, for example. In this case, only this page will need to be read. As
-people realized the potential of clustering, multidimensional clustering was introduced where data is clustered based on different criteria at the same time
-(TALK MORE).
+Clustering reflects how the records are physically located together on disk. This is very important since sequential I/O access is much cheaper than random
+access. Therefore, records that are likely to be queried together should be stored physically together to decrease the number of random I/O accesses.
+MOREEEEEEEEEEEEEEEE
 
@@
 
-A Main Memory Database (MMDB) (sometimes referred to as in-memory database) is one where the data resides in the main memory of the system rather than on a
+A Main Memory Database (MMDB) (sometimes referred to as in-memory database) is one where the primary copy of the data resides in the main memory of the system
+rather than on a
 disk~\cite{garcia1992main}. It should be noted that this is different from caching. Disk based database systems use the main memory to cache query results.
-However, MMDB's primary copy of the data resides in main memory. MMDB have many implementation challenges that have been addressed throughout the research
-community.
-
-Garcia-Molina and Salem~\cite{garcia1992main} mention some of the challenges involved. One is concurrency control. In disk based databases, locks are tracked
-through a hash table, but the actual objects on disk do not contain lock information. On the other hand, in MMDB, lock status is part of the object itself since
-it is cheap to keep a number of bits for that. The next challenge is access methods. In disk-based databases, B-Trees are used to index the data for faster
-access. B-Trees have a short bushy nature to try to decrease the height of the tree to decrease the number of I/O accesses. Since I/O access is not a problem in
-MMDB, longer tree structures are used since its cheap to access main memory. Another point is that data values do not need to be stored in the index itself
-since they will be stored in main memory anyways so there are no performance gains from storing them in the index itself. Table~\ref{tab:mmdbsumm} summarizes
+However, MMDB's primary copy of the data resides in main memory. Accessing data in main memory is faster than disk, and so MMDBs usually have better
+performance that disk based systems. However, main memory is more expensive and the size is limited in comparison to disk. However, MMDBs are essential to use
+where some of the collected data cannot, or will not, be stored on disk in the first place. For example, phone switch data may be needed for reasoning with
+other stored data, but it is captured in real time and will not necessarily be stored on disk. MMDB have many implementation challenges that have been addressed
+throughout the research community. These challenges will be discussed throughout this section. %Table~\ref{tab:mmdbsumm} summarizes
 the physical design elements of MMDBs.
@@
 \centering
-\scalebox{0.7}{
+\scalebox{0.9}{
 \begin{tabular}{p{3cm}|p{6cm}|p{6cm}}
@@
 structures developed for MMDB such as J+ Trees~\cite{luan2009prefetching}, but which have not gained much popularity, and so we do not discuss them in our
-survey. Table~\ref{tab:mmdbindexsumm} summarizes the features of the different index structures discussed in this section with respect to MMDB. The T Tree is
+survey. %Table~\ref{tab:mmdbindexsumm} summarizes the features of the different index structures discussed in this section with respect to MMDB. The T Tree is
 the most used index for MMDB (E.g. MySQL CLuster). However, there are other indexes used as well. For example, IBM's solidDB uses Trie (or prefix tree) for its
@@
 \centering
-\scalebox{0.7}{
+\scalebox{0.6}{
 \begin{tabular}{p{2cm}|p{6cm}|p{6cm}}
 \hline
- \textbf{Index Structure} & \textbf{Features suitable for MMDB} & \textbf{Cache Consciousness}\\
+ \textbf{Index Structure} & \textbf{Features} & \textbf{Cache Consciousness}\\
 \hline\hline
@@
 based on three values of a node: its start label, its end label, and its level (i.e. nested depth). On the other hand, Sedna~\cite{taranov2010sedna} uses a
-different reasoning to produce its numbering schema. Sedna assignes string identifiers to nodes such that they are lexically ordered according to the position
-of the node. Despite the differences in the ways the numbering is constructed, it seems that that no novel indexing structure is needed for XML databases. The
-problem is rather how to assign the nodes unique identifiers (the numbers or labels in the numbering schema) that can be used as identifiers in the index to
-reflect the XML document structure.
+different reasoning to produce its numbering schema. Sedna assigns string identifiers to nodes such that they are lexically ordered according to the position
+of the node. Despite the differences in the ways the numbering is constructed, it seems that existing indexing techniques can be adapted for XML databases. The
+challenge is how to assign the nodes unique identifiers (the numbers or labels in the numbering schema) that can be used as identifiers in the index to
+reflect the XML document structure. Additionally, in native XML databases, documents are usually grouped into collections, and an index may be used to search
+within these collections. This is different from the node index which searches within a document.
 
-Besides indexing the actual nodes, path indexing is also common in XML databases since usually queries search for a specific path. This is done by clustering
-together all the IDs of nodes on a given path, and creating an index for them~\cite{milo1999index, arion2008path}.
-
-talk about Tamino (http://www.rpbourret.com/xml/XMLAndDatabases.htm)~\cite{bourret2003xml}
+Besides indexing the actual nodes, path indexing is also common in XML databases since usually queries search for a specific path which can be very expensive.
+This is done by clustering together all the IDs of nodes on a given path, and creating an index for them~\cite{milo1999index, arion2008path}. Other techniques
+include that proposed by Cooper et al.~\cite{cooper2001} where they propose an indexing mechanism for paths called ``Index Fabric''. An Index Fabric is based on
+Patricia tries~\cite{knuth1998} which are used for string indexing, but is customized for disk-based systems. A path in an XML document is encoded into a unique
+string, and is then inserted into the fabric. The encoded string is usually very short, and so searching is relatively fast.
 
@@
 \label{SEC-WAREHOUSES}
-A data warehouse is a subject-oriented, integrated, time-variant, and non-volatile collection of data and decision support technologies, aimed at enabling the knowledge worker (executive, manager, analyst) to make better and faster decisions \cite{560407, 248616}.
-More than half of IT executives named Data warehousing as the highest-priority post-millennium project for them \cite{sen2005comparison}.
-The value of data warehousing for an organization depends on the organization's need for reliable, consolidated, unique and integrated reporting and analysis of its data, at different levels of aggregation.
+A data warehouse is a subject-oriented, integrated, time-variant, and non-volatile collection of data and decision support technologies, aimed at enabling the
+knowledge worker (executive, manager, analyst) to make better and faster decisions \cite{560407, chaudhuri1997overview}. More than half of IT executives named
+Data warehousing as the highest-priority post-millennium project for them \cite{sen2005comparison}. The value of data warehousing for an organization depends on
+the organization's need for reliable, consolidated, unique and integrated reporting and analysis of its data, at different levels of aggregation.
 
@@
 management), financial services (for claims analysis, risk analysis, credit card analysis, and fraud detection), transportation (for fleet management),
-telecommunications (for call analysis and fraud detection), utilities (for power usage analysis), and healthcare (for outcomes analysis) \cite{248616}.
+telecommunications (for call analysis and fraud detection), utilities (for power usage analysis), and healthcare (for outcomes analysis)
+\cite{chaudhuri1997overview}.
 
-Data warehouses tend to be extremely large, in fact it is quite possible for a data warehouse to store tens of petabytes of data while loading tens of terabytes of data everyday \cite{thusoo2010data}.
-Datta et al.
-\cite {628286} notes that the information in a data warehouse is usually multidimensional in nature, requiring the capability to view the data from a variety of perspectives.
-Aggregated and summarized data become more crucial than detailed records in such environment.
-Therefore, the workloads are query intensive with mostly ad hoc, complex queries, often requiring computationally expensive operations such as scans, joins, and aggregation.
-Performing such operations on large amounts of data, like in the case of data warehousing, complicates the situation further.
+Data warehouses tend to be extremely large, in fact it is quite possible for a data warehouse to store tens of petabytes of data while loading tens of terabytes
+of data everyday \cite{thusoo2010data}. Datta et al. \cite {628286} notes that the information in a data warehouse is usually multidimensional in nature,
+requiring the capability to view the data from a variety of perspectives. Aggregated and summarized data become more crucial than detailed records in such
+environment. Therefore, the workloads are query intensive with mostly ad hoc, complex queries, often requiring computationally expensive operations such as
+scans, joins, and aggregation. Performing such operations on large amounts of data, like in the case of data warehousing, complicates the situation further.
 Moreover, the results have to be delivered interactively to the business analyst using the system.
 
-Although data in a data warehouse is extracted and loaded from multiple on-line transaction processing (OLTP) data sources (including DB2, Oracle, IMS databases, and flat files) using Extract, Transfer, and Load (ETL) tools (see Figure \ref{fig:dw}), a data warehouse is usually maintained seperately from the organization's operational databases \cite{sen2005comparison, 248616}.
-This architecture can be justified owing to the fact that operational databases are finely tuned to satisfy known OLTP requirements and functionalities which are quite different from that of on-line analytical processing (OLAP) which is supported by data warehousing.
-Analyzing data for decision support usually requires consolidating data from many heterogeneous sources of varying quality, or use inconsistent representations, codes and formats.
-Moreover, understanding trends or making predictions requires historical data, whereas operational databases store only current data.
-In OLAP, there's a need to support multidimensional data models and operations which requires special data organization, access methods, and implementation methods, not generally provided by DBMSs targeted for OLTP \cite{248616}.
-Finally, in data warehousing, query throughput and response times are more important than transaction throughput which is the major performance metric for operational databases.
+Although data in a data warehouse is extracted and loaded from multiple on-line transaction processing (OLTP) data sources (including DB2, Oracle, IMS
+databases, and flat files) using Extract, Transfer, and Load (ETL) tools (see Figure \ref{fig:dw}), a data warehouse is usually maintained seperately from the
+organization's operational databases \cite{sen2005comparison, chaudhuri1997overview}. This architecture can be justified owing to the fact that operational
+databases are finely tuned to satisfy known OLTP requirements and functionalities which are quite different from that of on-line analytical processing (OLAP)
+which is supported by data warehousing. Analyzing data for decision support usually requires consolidating data from many heterogeneous sources of varying
+quality, or use inconsistent representations, codes and formats. Moreover, understanding trends or making predictions requires historical data, whereas
+operational databases store only current data. In OLAP, there's a need to support multidimensional data models and operations which requires special data
+organization, access methods, and implementation methods, not generally provided by DBMSs targeted for OLTP \cite{chaudhuri1997overview}. Finally, in data
+warehousing, query throughput and response times are more important than transaction throughput which is the major performance metric for operational databases.
 
@@
 \includegraphics[width=9cm]{figs/dw.png}
-\caption{Data Warehousing Architecture \cite{248616}}
+\caption{Data Warehousing Architecture \cite{chaudhuri1997overview}}
 \label{fig:dw}
@@
 
-\section{Automating Physical Database Design}
-\label{SEC-AUTO}
+\section{Open Research Problems in Physical Database Design}
+\label{SEC-OPEN}
 
@@
 as the query workload. The number of combinations of choices that can be made is very large making this task quite tedious and difficult for database
-administrators. Accordingly, researchers have been inclined to find ways to automate these decisions. There were many research efforts done in this
-area(E.g.~\cite{finkelstein1988physical, hammer76selec, frank1992selec}), but most of these efforts suffered from either examining too many configurations, not
-accounting for the query workload, or not considering if the query optimizer will benefit from the proposed configurations or not. Chaudhuri and
-Narasayya~\cite{chadhuri1997selec} were the first to propose an efficient automatic index selection tool that is dependent on the query workload and uses the
-query optimizer to evaluate the proposed configuration. They find the best configuration for the workload by examining each query individually to produce
-single-column indexes then combining these configurations. They then run a greedy algorithm to select the best configuration with the least cost (as determined
-by the query optimizer). They then try to further optimize the configuration by looking for possible multicolumn indexes as a combination of the single-column
-indexes already found. One of the main strengths of this technique was that all these indexes were virtually created such that the query optimizer could use
-them, but without having to incur the cost of physically creating them. This solution was part of the research at Microsoft and was tested in Microsoft SQL
-Server 7.0.
+administrators. Accordingly, researchers have been inclined to find ways to automate these decisions. There were many early research efforts done in this
+area(E.g.~\cite{finkelstein1988physical, hammer76selec, frank1992selec, chaudhuri1997overview, schiefer1999db2, valentin2000select}), but most of these efforts
+focused on having a static workload. Accordingly, research shifted to more dynamic ways to automate physical design. Initial attempts used an dynamic, on-line
+approaches(E.g.~\cite{bruno2007online, sattler2003, schnaitter2007}) where they observed the changes in the workload and adjusted the design accordingly while
+attempting to predict future behavior based on the observed behavior. To avoid the problems of on-line design, researchers are also currently looking at how to
+perform this dynamic physical design off-line (E.g. ~\cite{hannes2008,agrawal2006}).
 
-Similarly, IBM~\cite{schiefer1999db2, valentin2000select} suggested a similar technique for their database with some changes such as selecting the order of
-columns within an index based on the order in which the query optimizer evaluates the predicates.
-
-\section{Open Problems in Physical Database Design}
-\label{SEC-OPEN}
+Other problems which are currently being researched include improving I/O operations in database systems by using Solid State Drives (SSD) instead of of Hard
+Disk Drives (HDD)~\cite{lee2009, du2009, lee2008}. SSDs have faster read performance, but perform poorly in random writes. However, for certain applications in
+databases and for databases where random writes are not as frequent, SSDs might be suitable. Using SDDs also seems promising when used for
+transaction logs, and intermediate results of sorting and joining~\cite{lee2008}.
 

### 2010-07-24 cs848-db/Paper.tex

@@
 \begin{abstract}
-%\boldmath
-The abstract goes here.
+The physical design decisions taken during database design highly affect the database performance. These design decisions include the index structures used,
+which views can be materialized, how to partition the data etc. These decisions vary by the nature of the system being implemented. For example, the index
+structures used for disk-based database systems differ from those used in a memory-based database system. In this survey, we present the elements of
+physical design in different information systems. We start by describing the physical design of traditional disk-based relational database systems then compare
+it to other types of information systems. Specifically, we look at the physical design of Main Memory Database Systems (MMDB), XML Databases, and Data
+Warehouses. We examine how the specific nature of each of these systems affect its physical design. 
 \end{abstract}
@@
 
-Choosing which indexes to create, which views to materialize, and how to partition data are hard choices. They are highly dependent on the schema used as well
+Automating the physical database design choices has been a very attractive area of research that is still receiving lots of attention until now. Choosing which
+indexes to create, which views to materialize, and how to partition data are hard choices. They are highly dependent on the schema used as well
 as the query workload. The number of combinations of choices that can be made is very large making this task quite tedious and difficult for database
@@
 attempting to predict future behavior based on the observed behavior. To avoid the problems of on-line design, researchers are also currently looking at how to
-perform this dynamic physical design off-line (E.g. ~\cite{hannes2008,agrawal2006}).
+perform this dynamic physical design off-line (E.g. ~\cite{hannes2008,agrawal2006}). Additionally, finding better evaluation techniques to compare different
+automatic physical design decision is currently being studied~\cite{gebaly2008}.
 
@@
 
-Indexing in the cloud~\cite{wang2010indexing}.
+Just like all other research areas are moving to the cloud, database systems are moving there too. Accordingly, lots of current research focuses on adapting
+storage techniques and index structures to the cloud. For example, Wang et al. propose a new index structure for indexing mult-dimensional data in the
+cloud~\cite{wang2010indexing}. Lots of work is currently being done in this area. For example, Wu et al.~\cite{wu2010efficient} propose a modified B+tree index
+for datasets stored in each compute node. The exact requirements and solutions for data management in the cloud are still being
+researched~\cite{abadi2009data}, and so this is an area that is expected to receive more attention in the future as cloud computing becomes more prevalent.
 
-Two-Tier Storage DBMS (using disk based and main memory dbs)~\cite{eo2008two}.
+This is by no means an exhaustive list of the problems still being researched in physical database design, but they provide a flavor of the challenges still
+being faced. As shown, as new technologies such as SSDs or cloud computing environments emerge, database systems will need to adapt themselves to either make
+use of them or to be able to cope with the new requirements they pose.
 

### 2010-07-25 trunk/cs848-db/Paper.tex
integrated with Sarah's latest work

@@
 \usepackage{fullpage}
-\usepackage{multirow}
 
@@
 \usepackage{url}
-
+\usepackage{multirow}
 \begin{document}
@@
 
-Relational Database Management Systems (RDBMS) have been the most popular type of Information Systems for decades. When we think of designing a database
-system, we usually think of which data we need to keep track of and how will we properly organize it into meaningful tables. However, this is not the only phase
-in database design. All of the designed tables need to be physically stored on disk. How these tables will be stored and accessed greatly affects the database
+Relational Database Management Systems (RDBMS) have been the most popular type of Information Systems for decades. When we think of designing a database system,
+we usually think of which data we need to keep track of and how will we properly organize it into meaningful tables. However, this is not the only phase in
+database design. All of the designed tables need to be physically stored on disk. How these tables will be stored and accessed greatly affects the database
 performance. Physical database design is concerned with the actual storage of data on disk. This means how the files are organized, and how they are accessed.
-Data is stored in the
-form of records in files which reside on a physical disk. In this sense, we try to find the most efficient way possible to store and access the data such that
-queries are answered and executed efficiently. In the traditional relational database system, this involved selecting proper indexing techniques,
-partitioning the data for more efficient access, materializing the results of some queries, clustering and compressing the data, and many other
-techniques~\cite{fikelstein1988physical,lightstone2007physical}. The physical design aspects of relational databases have been studied in much detail. However,
+Data is stored in the form of records in files which reside on a physical disk. In this sense, we try to find the most efficient way possible to store and
+access the data such that queries are answered and executed efficiently. In the traditional relational database system, this involved selecting proper indexing
+techniques, partitioning the data for more efficient access, materializing the results of some queries, clustering and compressing the data, and many other
+techniques \cite{fikelstein1988physical,lightstone2007physical}. The physical design aspects of relational databases have been studied in much detail. However,
 over time, different types of Information Systems have emerged that have different characteristics. For example, most relational databases have been disk-based
@@
 systems. This includes explaining how alterations or additions are made to data structures to fit the specific needs of every system. Using the different
-physical design methods used in disk based
-RDBMSs, we explore how these methods apply to the other systems and where the differences and similarities lie.
+physical design methods used in disk based RDBMSs, we explore how these methods apply to the other systems and where the differences and similarities lie.
 
-The rest of the survey is organized as follows. Section~\ref{SEC-ELEMENTS} first explains different elements of physical database design used in traditional
-disk-based databases. Section~\ref{SEC-DIFFSYS} then discusses how each of the three other systems uses these elements, as
-applicable. Sections~\ref{SEC-MMDB},~\ref{SEC-XML},~\ref{SEC-WAREHOUSES}, discuss Main Memory Databases, XML Databases, and Data Warehouses
-respectively. Throughout these sections, comparisons are made as to how the different elements have been modified to fit the needs of each system.
-Section~\ref{SEC-OPEN} mentions some of the
-open problems of physical database design, and suggests possible solutions. Finally Section~\ref{SEC-CONCL} summarizes and concludes this survey.
+The rest of the survey is organized as follows. Section \ref{SEC-ELEMENTS} first explains different elements of physical database design used in traditional
+disk-based databases. Section \ref{SEC-DIFFSYS} then discusses how each of the three other systems uses these elements, as applicable. Sections \ref{SEC-MMDB},
+\ref{SEC-XML}, \ref{SEC-WAREHOUSES}, discuss Main Memory Databases, XML Databases, and Data Warehouses respectively. Throughout these sections, comparisons are
+made as to how the different elements have been modified to fit the needs of each system. Section \ref{SEC-OPEN} mentions some of the open problems of physical
+database design, and suggests possible solutions. Finally, Section \ref{SEC-CONCL} summarizes and concludes this survey.
 
@@
 
-Relational Databases were first introduced by Codd~\cite{codd1970relational}. Disk based Relational Database Management Systems (RDBMS) are the most common and
-popular
-type of Information Systems. Since they have been extensively researched and have been around for many decades, we use them as the basis of our comparison. In
-this section, we describe the different elements of physical design in RDBMS. These elements will also be examined in the other systems to see how they apply
-to them.
+Relational Databases were first introduced by Codd \cite{codd1970relational}. Disk based Relational Database Management Systems (RDBMS) are the most common and
+popular type of Information Systems. Since they have been extensively researched and have been around for many decades, we use them as the basis of our
+comparison. In this section, we describe the different elements of physical design in RDBMS. These elements will also be examined in the other systems to see
+how they apply to them.
 
@@
 Btrees (short for Balanced Trees) are the most used database index structure.  The Btree is essentially similar to the traditional Binary Search Tree, but
-instead of having one value per node, a B-tee can have many values per node~\cite{comer1979ubiquitous}. B-trees are suitable for relational databases since the
-cost of retrieval in a B-tree is at most proportional to: $log_{d}\frac{n+1}{2}$. The cost of insertion
-and deletion is at most proportional to $log_{d}n$ due to the possibility of progressing back up the tree to balance it after an insertion or a deletion.
-Although B-trees do well in retrieval, deletion and insertion, they do not perform well in sequential search.
+instead of having one value per node, a B-tee can have many values per node \cite{comer1979ubiquitous}. B-trees are suitable for relational databases since the
+cost of retrieval in a B-tree is at most proportional to: $log_{d}\frac{n+1}{2}$. The cost of insertion and deletion is at most proportional to $log_{d}n$ due
+to the possibility of progressing back up the tree to balance it after an insertion or a deletion. Although B-trees do well in retrieval, deletion and
+insertion, they do not perform well in sequential search.
 
-Different variations of B-trees have been used. For example, B+trees are now the main method of indexing in current disk-based Relational
-Databases~\cite{lightstone2007physical} such as DB2, Oracle, and SQL Server. The main difference between the
-B+tree and the Btree is that only leaf nodes contain data pointers in a B+tree, and leaf nodes contain pointers to each other as shown in
-Figure~\ref{fig:b+tree}. In Prefix B+trees, instead of using the
-actual key value of a record, a prefix from the key value is used to decrease the storage size needed which will in turn allow more records to be stored per
-node thus decreasing the height of the tree. B+trees have been successful in relational databases since data is ultimately recorded in files. Since B+trees have
-a high fanout, they allow less I/O operations to access a specific data record which is stored in a specific file.
+Different variations of B-trees have been used. For example, B+trees are now the main method of indexing in current disk-based Relational Databases
+\cite{lightstone2007physical} such as DB2, Oracle, and SQL Server. The main difference between the B+tree and the Btree is that only leaf nodes contain data
+pointers in a B+tree, and leaf nodes contain pointers to each other as shown in Figure \ref{fig:b+tree}. In Prefix B+trees, instead of using the actual key
+value of a record, a prefix from the key value is used to decrease the storage size needed which will in turn allow more records to be stored per node thus
+decreasing the height of the tree. B+trees have been successful in relational databases since data is ultimately recorded in files. Since B+trees have a high
+fanout, they allow less I/O operations to access a specific data record which is stored in a specific file.
 
-Of course, there are other indexing techniques used such as hash tables, and bitmap indices. EXPLAAAAAAAAIN
+Bitmap indices represent another type of index structures that is commonly used and cited in the literature. Using bitmaps as indexes first appeared in the
+literature as part of the architecture of the Model 204 commercial database of Computer Corporation of America for the IBM 370 computer \cite{patrick1987model}.
+A bitmap index structure is based on a bitmap (i.e. bitset or bit array) data structure where individual bits correspond to values in unique records. Given a
+table column $C$ where a record $x$ can have one of six unique values, a traditional bitmap indexe defined for $C$ generates one bitmap vector for each of the
+six different values ($B0-B5$). If the value of $C$ for record $x$ is set to 3, the corresponding bit in vector $B3$ is set to 1. Otherwise the bit is set to 0.
 
-briefly discuss clustered vs. unclustered ?
+Bitmap indices are best used when the set of possible key values in the index is small with a large number of rows, e.g. gender attribute \cite{212001}.
+Therefore, bitmap indices outperform B+trees for low cardinality attributes. The reason is that given an attribute with low cardinality, defining a B+tree index
+on it means that the tree will have very few leaf nodes, each one storing a very long list of record identifiers (RIDs). On the other hand, defining a
+bitmapindex on that same attribute means that such long lists or RIDs will not be stored, rather bit vectors that represent the distinct attribute values will
+be stored as previously explained. This leads to a direct benefit of consuming less storage space. Moreover, queries in that case can be answered using bit-wise
+logical operatios (e.g. and, or, not) which can improve query performance \cite{253268}. Another advantage of bitmap indices is the ability to perform more
+efficient joins by creating bitmap join indices and using those indices to calculate the result of the join \cite{212001}. Although there are many benefits
+gained from using bitmap indices, they poorly perform during updates. The reason is that when an updates is to be done to a bitmap index, the whole bit vector
+for the index is locked, blocking any access to the index for other operations, then the lock is released when the update is done. That is why bitmap indices
+are usually used for read-mostly databases, a famous example of which are data warehouses (Section \ref{SEC-WAREHOUSES}).
+
+Another variation of indexing, a hash table index, often simply known as a hash index is used to organize unordered physical database tables for efficient
+access \cite{lightstone2007physical}. This type of index is usually defined for keys that have unique values in the data records. Creating this index requires
+passing the attribute (key) value to a hash function that maps it to a starting block address (i.e. bucket address). Afterwards, table insertions will be done
+according to the hash function, as well as quering the records later.
+
+In general, indices are a way of organizing the logical structure of a database without altering its physical structure. However, sometimes it is desirable to
+store related records near each other on disk for better locality. This can be achieved through implementing a clustered index on the desired key values.
+Otherwise, the created index is known as non-clustered index. The primary advantage of a clustered index is therefore the ordering of the physical data rows in
+accordance with the index blocks that point to them which. Therefore, fewer data block reads are required to answer a given query. The only caveat with using a
+clustered index is that there can only be one clustered index per table because the physical organization of the table must be fixed
+\cite{lightstone2007physical}.
 
@@
 simply store the results of the queries on disk instead of recomputing them each time. Oracle 8i introduced materialized views to fit this purpose. By
-precalculating the
-results of a complex query, and storing them in a table on disk, the new table with the results is very likely to be much smaller than the original tables
-decreasing I/O access costs and thus increasing performance~\cite{lightstone2007physical, chaudhuri1998overview}. Once a materialized view exists, the user can
-either explicitly query the materialized view, or enter the original query, and have the query optimizer rewrite the query to use the materialized views instead
-of the original tables. Many query optimization techniques have relied on rewriting queries using views (E.g.~\cite{levy1995answering, gupta1995aggregate,
-goldstein2001optimizing,
-abiteboul1998complexity}), but discussing the details of the rewriting process is beyond the scope of this survey.
+precalculating the results of a complex query, and storing them in a table on disk, the new table with the results is very likely to be much smaller than the
+original tables decreasing I/O access costs and thus increasing performance \cite{lightstone2007physical, chaudhuri1998overview}. Once a materialized view
+exists, the user can either explicitly query the materialized view, or enter the original query, and have the query optimizer rewrite the query to use the
+materialized views instead of the original tables. Many query optimization techniques have relied on rewriting queries using views (e.g.
+\cite{levy1995answering, gupta1995aggregate, goldstein2001optimizing, abiteboul1998complexity}), but discussing the details of the rewriting process is beyond
+the scope of this survey.
 
 Despite the fact that materialized views provide a significant improvement in performance, we cannot simply create a materialized view for each frequent query.
-First, materialized views consume disk storage which might be limited. The second, and main, problem with materialized views is their maintenance.
-Ensuring
-that the views have the most up to date data is tricky, and may outset the benefit of using materialized views if it is not properly designed. Finally, having
+First, materialized views consume disk storage which might be limited. The second, and main, problem with materialized views is their maintenance. Ensuring that
+the views have the most up to date data is tricky, and may outset the benefit of using materialized views if it is not properly designed. Finally, having
 several materialized views may increase the cost of searching for the appropriate view to use during the query optimization stage.
@@
 
-Partitioning is an important aspect of physical design as it reduces table scanning time~\cite{lightstone2007physical}. There are two main categories of
-partitioning: horizontal partitioning
-and vertical partitioning. Horizontal partitioning divides the table into sets of table rows where each row or record still has all the attributes of the table.
-For example, dividing the data by date where all data dating less than year 2000 lies in one partition while all data dating more lies in another. On the other
-hand, vertical partitioning reduces the width of the dataset by storing some attributes in one table, and some in another. Usually, less frequently queried
-columns are partitioned together to reduce the size of the table.
+Partitioning is an important aspect of physical design as it reduces table scanning time \cite{lightstone2007physical}. There are two main categories of
+partitioning: horizontal partitioning and vertical partitioning \cite{agrawal2004integrating}. Horizontal partitioning divides the table into sets of table rows
+where each row or record still has all the attributes of the table \cite{ceri1982horizontal, shin1985partitioning}. For example, dividing the data by date where
+all data dating less than year 2000 lies in one partition while all data dating more lies in another. On the other hand, vertical partitioning reduces the width
+of the dataset by storing some attributes in one table, and some in another. Usually, less frequently queried columns are partitioned together to reduce the
+size of the table, especially in data warehouses where not all the columns of a table are analyzed for decision support.
 
-The main advantage of any type of partitioning is to reduce the amount of time it takes to scan a table which in turn improves performance. For example, if a
-table has 1000 records, and a query is only using the first 20 records then having these 20 records in a partition containing 50 records will save time as
-opposed to examining all the 1000 records. Horizontal partitioning is more widely used, and is implemented in all current DBMSes. Table~\ref{tab:horpartition}
-shows different categories of horizontal partitioning, and the situations they are suited to. Combining partitioning and indexing can often lead to better
-performance. For example, if List partitioning is used to divide a table by regions and then a bitmap index is used on that table, finding records in the same
-region or state can be much faster. Finally, partitioning the data into different physical devices to achieve parallelism when accessing it can greatly
-increase the performance of the database.
-
-NEED SOME REFERENCES HERE
+The main advantage of any type of partitioning is to reduce the amount of time it takes to scan a table which in turn improves performance
+\cite{lightstone2007physical}. For example, if a table has 1000 records, and a query is only using the first 20 records then having these 20 records in a
+partition containing 50 records will save time as opposed to examining all the 1000 records. Horizontal partitioning is more widely used, and is implemented in
+all current DBMSes \cite{mysqlpartitioning, sqlsrvpartitioning, sybasepartitioning, postgrepartitioning, oraclepartitioning, db2partitioning} which explains
+why there exists many flavors of horizontal paritioning techniques. There are three major types of horizontal partitioning: range partitioning
+\cite{dewitt1986gamma}, list partitioning \cite{lightstone2007physical} (which is a variation of range partitioning), and hash partitioning
+\cite{lightstone2007physical}. Of course, one can mix and match those partitioning schemes and come up with a composite partitioning that fits a particular
+situation \cite{lightstone2007physical}. Table \ref{tab:horpartition} shows different categories of horizontal partitioning, and the situations they are suited
+to. Combining partitioning and indexing can often lead to better performance. For example, if List partitioning is used to divide a table by regions and then a
+bitmap index is used on that table, finding records in the same region or state can be much faster. Finally, partitioning the data into different physical
+devices to achieve parallelism when accessing it can greatly increase the performance of the database.
 
 \begin{table}[!t]
-\scalebox{0.8}{
+\scalebox{0.7}{
  \begin{tabular}{p{2cm}||p{4cm}|p{4cm}|p{4cm}|p{4cm}}
@@
 \hline
-\textbf{Benefits} & \multicolumn{4}{c}{Partition elimination, Improved administration, Fast roll-in/out, and Implicit clustering }
-\\
+\textbf{Benefits} & \multicolumn{4}{c}{Partition elimination, Improved administration, Fast roll-in/out, and Implicit clustering }\\
 \hline
@@
 		    \item ProductId
-		    \end{itemize} & Hash then List \\
+		    \end{itemize} &
+		    Hash then List \\
 \hline
 \textbf{RDBMS} & MySQL, DB2, Oracle, PostgreSQL, SQL Server & MySQL, Oracle, PostgreSQL, SQL Server & MySQL, Oracle & MySQL, Oracle\\
-
 \end{tabular}
@@
 other stored data, but it is captured in real time and will not necessarily be stored on disk. MMDB have many implementation challenges that have been addressed
-throughout the research community. These challenges will be discussed throughout this section. Table~\ref{tab:mmdbsumm} summarizes
+throughout the research community. These challenges will be discussed throughout this section. %Table~\ref{tab:mmdbsumm} summarizes
 the physical design elements of MMDBs.
@@
 \centering
-\scalebox{0.7}{
+\scalebox{0.6}{
 \begin{tabular}{p{2cm}|p{6cm}|p{6cm}}
@@
 attempting to predict future behavior based on the observed behavior. To avoid the problems of on-line design, researchers are also currently looking at how to
-perform this dynamic physical design off-line (E.g. ~\cite{hannes2008,agrawal2006}).
+perform this dynamic physical design off-line (E.g. ~\cite{hannes2008,agrawal2006}). Additionally, finding better evaluation techniques to compare different
+automatic physical design decision is currently being studied~\cite{gebaly2008}.
 
-A second question is how to characterize
-scenarios or classes of workloads for which constrained dynamic
-physical designs will be beneficial.
+Other problems which are currently being researched include improving I/O operations in database systems by using Solid State Drives (SSD) instead of of Hard
+Disk Drives (HDD)~\cite{lee2009, du2009, lee2008}. SSDs have faster read performance, but perform poorly in random writes. However, for certain applications in
+databases and for databases where random writes are not as frequent, SSDs might be suitable. Using SDDs also seems promising when used for
+transaction logs, and intermediate results of sorting and joining~\cite{lee2008}.
 
-Similarly, IBM~\cite{} suggested a similar technique for their database with some changes such as selecting the order of
-columns within an index based on the order in which the query optimizer evaluates the predicates.
+Just like all other research areas are moving to the cloud, database systems are moving there too. Accordingly, lots of current research focuses on adapting
+storage techniques and index structures to the cloud. For example, Wang et al. propose a new index structure for indexing mult-dimensional data in the
+cloud~\cite{wang2010indexing}. Lots of work is currently being done in this area. For example, Wu et al.~\cite{wu2010efficient} propose a modified B+tree index
+for datasets stored in each compute node. The exact requirements and solutions for data management in the cloud are still being
+researched~\cite{abadi2009data}, and so this is an area that is expected to receive more attention in the future as cloud computing becomes more prevalent.
 
-Indexing in the cloud~\cite{wang2010indexing}.
-
-Two-Tier Storage DBMS (using disk based and main memory dbs)~\cite{eo2008two}.
+This is by no means an exhaustive list of the problems still being researched in physical database design, but they provide a flavor of the challenges still
+being faced. As shown, as new technologies such as SSDs or cloud computing environments emerge, database systems will need to adapt themselves to either make
+use of them or to be able to cope with the new requirements they pose.
 

### 2010-07-25 cs848-db/Paper.tex
finished clustering

@@
 \usepackage{multirow}
-
 \begin{document}
@@
 
-Relational Database Management Systems (RDBMS) have been the most popular type of Information Systems for decades. When we think of designing a database
-system, we usually think of which data we need to keep track of and how will we properly organize it into meaningful tables. However, this is not the only phase
-in database design. All of the designed tables need to be physically stored on disk. How these tables will be stored and accessed greatly affects the database
+Relational Database Management Systems (RDBMS) have been the most popular type of Information Systems for decades. When we think of designing a database system,
+we usually think of which data we need to keep track of and how will we properly organize it into meaningful tables. However, this is not the only phase in
+database design. All of the designed tables need to be physically stored on disk. How these tables will be stored and accessed greatly affects the database
 performance. Physical database design is concerned with the actual storage of data on disk. This means how the files are organized, and how they are accessed.
-Data is stored in the
-form of records in files which reside on a physical disk. In this sense, we try to find the most efficient way possible to store and access the data such that
-queries are answered and executed efficiently. In the traditional relational database system, this involved selecting proper indexing techniques,
-partitioning the data for more efficient access, materializing the results of some queries, clustering and compressing the data, and many other
-techniques~\cite{finkelstein1988physical,lightstone2007physical}. The physical design aspects of relational databases have been studied in much detail. However,
+Data is stored in the form of records in files which reside on a physical disk. In this sense, we try to find the most efficient way possible to store and
+access the data such that queries are answered and executed efficiently. In the traditional relational database system, this involved selecting proper indexing
+techniques, partitioning the data for more efficient access, materializing the results of some queries, clustering and compressing the data, and many other
+techniques \cite{fikelstein1988physical,lightstone2007physical}. The physical design aspects of relational databases have been studied in much detail. However,
 over time, different types of Information Systems have emerged that have different characteristics. For example, most relational databases have been disk-based
@@
 systems. This includes explaining how alterations or additions are made to data structures to fit the specific needs of every system. Using the different
-physical design methods used in disk based
-RDBMSs, we explore how these methods apply to the other systems and where the differences and similarities lie.
+physical design methods used in disk based RDBMSs, we explore how these methods apply to the other systems and where the differences and similarities lie.
 
-The rest of the survey is organized as follows. Section~\ref{SEC-ELEMENTS} first explains different elements of physical database design used in traditional
-disk-based databases. Section~\ref{SEC-DIFFSYS} then discusses how each of the three other systems uses these elements, as
-applicable. Sections~\ref{SEC-MMDB},~\ref{SEC-XML},~\ref{SEC-WAREHOUSES}, discuss Main Memory Databases, XML Databases, and Data Warehouses
-respectively. Throughout these sections, comparisons are made as to how the different elements have been modified to fit the needs of each system.
-Section~\ref{SEC-OPEN} mentions some of the current
-open problems of physical database design. Finally Section~\ref{SEC-CONCL} summarizes and concludes this survey.
+The rest of the survey is organized as follows. Section \ref{SEC-ELEMENTS} first explains different elements of physical database design used in traditional
+disk-based databases. Section \ref{SEC-DIFFSYS} then discusses how each of the three other systems uses these elements, as applicable. Sections \ref{SEC-MMDB},
+\ref{SEC-XML}, \ref{SEC-WAREHOUSES}, discuss Main Memory Databases, XML Databases, and Data Warehouses respectively. Throughout these sections, comparisons are
+made as to how the different elements have been modified to fit the needs of each system. Section \ref{SEC-OPEN} mentions some of the open problems of physical
+database design, and suggests possible solutions. Finally, Section \ref{SEC-CONCL} summarizes and concludes this survey.
 
@@
 
-Partitioning is an important aspect of physical design as it reduces table scanning time~\cite{lightstone2007physical}. There are two main categories of
-partitioning: horizontal partitioning
-and vertical partitioning. Horizontal partitioning divides the table into sets of table rows where each row or record still has all the attributes of the table.
-For example, dividing the data by date where all data dating less than year 2000 lies in one partition while all data dating more lies in another. On the other
-hand, vertical partitioning reduces the width of the dataset by storing some attributes in one table, and some in another. Usually, less frequently queried
-columns are partitioned together to reduce the size of the table.
+Partitioning in physical database design is a method for reducing the workload on any one hardware component, like an individual disk, by partitioning
+(dividing) the data over several disks \cite{lightstone2007physical}. There are two main categories of partitioning: horizontal partitioning and vertical
+partitioning \cite{agrawal2004integrating}. Horizontal partitioning divides the table into sets of table rows where each row or record still has all the
+attributes of the table \cite{ceri1982horizontal, shin1985partitioning}. For example, dividing the data by date where all data dating less than year 2000 lies
+in one partition while all data dating more lies in another. On the other hand, vertical partitioning reduces the width of the dataset by storing some
+attributes in one table, and some in another. Usually, less frequently queried columns are partitioned together to reduce the size of the table, especially in
+data warehouses where not all the columns of a table are analyzed for decision support.
 
-The main advantage of any type of partitioning is to reduce the amount of time it takes to scan a table which in turn improves performance. For example, if a
-table has 1000 records, and a query is only using the first 20 records then having these 20 records in a partition containing 50 records will save time as
-opposed to examining all the 1000 records. Horizontal partitioning is more widely used, and is implemented in all current DBMSes.
-
-%Table~\ref{tab:horpartition}
-shows different categories of horizontal partitioning, and the situations they are suited to. Combining partitioning and indexing can often lead to better
-performance. For example, if List partitioning is used to divide a table by regions and then a bitmap index is used on that table, finding records in the same
-region or state can be much faster. Finally, partitioning the data into different physical devices to achieve parallelism when accessing it can greatly
-increase the performance of the database.
-
-NEED SOME REFERENCES HERE
+The main advantage of any type of partitioning is to reduce the amount of time it takes to scan a table which in turn improves performance
+\cite{lightstone2007physical}. For example, if a table has 1000 records, and a query is only using the first 20 records then having these 20 records in a
+partition containing 50 records will save time as opposed to examining all the 1000 records. Horizontal partitioning is more widely used, and is implemented in
+all current DBMSes \cite{mysqlpartitioning, sqlsrvpartitioning, sybasepartitioning, postgrepartitioning, oraclepartitioning, db2partitioning} which explains
+why there exists many flavors of horizontal paritioning techniques. There are three major types of horizontal partitioning: range partitioning
+\cite{dewitt1986gamma}, list partitioning \cite{lightstone2007physical} (which is a variation of range partitioning), and hash partitioning
+\cite{lightstone2007physical}. Of course, one can mix and match those partitioning schemes and come up with a composite partitioning that fits a particular
+situation \cite{lightstone2007physical}. Table \ref{tab:horpartition} shows different categories of horizontal partitioning, and the situations they are suited
+to. Combining partitioning and indexing can often lead to better performance. For example, if List partitioning is used to divide a table by regions and then a
+bitmap index is used on that table, finding records in the same region or state can be much faster. Finally, partitioning the data into different physical
+devices to achieve parallelism when accessing it can greatly increase the performance of the database.
 
@@
 \subsection{Clustering}
-Clustering reflects how the records are physically located together on disk. This is very important since sequential I/O access is much cheaper than random
-access. Therefore, records that are likely to be queried together should be stored physically together to decrease the number of random I/O accesses.
-MOREEEEEEEEEEEEEEEE
+Clustering is the process of grouping related data together on disk for efficiency of access and maximizing resource utilization. Data clustering benefits from
+the fact that sequential I/O access in physical disks is much cheaper than random access. Therefore, records that are likely to be queried together are
+clustered (stored physically together) to decrease the number of random I/O accesses. Whenever the number of I/O random accesses is decreased, CPU blocks are
+also reduced which result in an overall reduction of CPU costs improving the query performance, especially for multidimensional queries (e.g. group by)
+\cite{lightstone2007physical}. The same concept of data clustering can similarly be applied to multiple dimensions (columns or column attributes) of a table.
+Kennedy et al. \cite{kennedy2005} summarizes the benefits of multidimensional clustering as follows:
+
+\begin{itemize}
+ \item Dramatic reduction in I/O requirements due to clustering.
+ \item Scans on any dimension index provide clustered data access.
+ \item Dimensions/slices can be accessed independently using their block indexes.
+ \item Block index scans can be combined (ANDed, ORed).
+ \item Access to the clustered data is much faster.
+\end{itemize}
+
+On the other hand, defining clustering keys, clustering scheme, and the granularity of clustering which makes the problem of designing an optimal clustering in
+this context has combinatorial complexity, having an extremely large search space \cite{jagadish1999snakes}.
 
@@
 other stored data, but it is captured in real time and will not necessarily be stored on disk. MMDB have many implementation challenges that have been addressed
-throughout the research community. These challenges will be discussed throughout this section. %Table~\ref{tab:mmdbsumm} summarizes
+throughout the research community. These challenges will be discussed throughout this section. Table~\ref{tab:mmdbsumm} summarizes
 the physical design elements of MMDBs.
@@
 structures developed for MMDB such as J+ Trees~\cite{luan2009prefetching}, but which have not gained much popularity, and so we do not discuss them in our
-survey. %Table~\ref{tab:mmdbindexsumm} summarizes the features of the different index structures discussed in this section with respect to MMDB. The T Tree is
+survey. Table~\ref{tab:mmdbindexsumm} summarizes the features of the different index structures discussed in this section with respect to MMDB. The T Tree is
 the most used index for MMDB (E.g. MySQL CLuster). However, there are other indexes used as well. For example, IBM's solidDB uses Trie (or prefix tree) for its
@@
 \end{table}
-
-%cite~\cite{dewitt1984implementation}.
 
@@
 
-%\bibliographystyle{IEEEtran}
-
 \bibliographystyle{IEEEtran}

### 2010-07-26 trunk/cs848-db/Paper.tex

@@
 \usepackage{multirow}
+\usepackage{array}
+
 \begin{document}
@@
 techniques, partitioning the data for more efficient access, materializing the results of some queries, clustering and compressing the data, and many other
-techniques \cite{fikelstein1988physical,lightstone2007physical}. The physical design aspects of relational databases have been studied in much detail. However,
+techniques \cite{finkelstein1988physical,lightstone2007physical}. The physical design aspects of relational databases have been studied in much detail. However,
 over time, different types of Information Systems have emerged that have different characteristics. For example, most relational databases have been disk-based
@@
 A bitmap index structure is based on a bitmap (i.e. bitset or bit array) data structure where individual bits correspond to values in unique records. Given a
-table column $C$ where a record $x$ can have one of six unique values, a traditional bitmap indexe defined for $C$ generates one bitmap vector for each of the
+table column $C$ where a record $x$ can have one of six unique values, a traditional bitmap index defined for $C$ generates one bitmap vector for each of the
 six different values ($B0-B5$). If the value of $C$ for record $x$ is set to 3, the corresponding bit in vector $B3$ is set to 1. Otherwise the bit is set to 0.
@@
 on it means that the tree will have very few leaf nodes, each one storing a very long list of record identifiers (RIDs). On the other hand, defining a
-bitmapindex on that same attribute means that such long lists or RIDs will not be stored, rather bit vectors that represent the distinct attribute values will
+bitmap index on that same attribute means that such long lists or RIDs will not be stored, rather bit vectors that represent the distinct attribute values will
 be stored as previously explained. This leads to a direct benefit of consuming less storage space. Moreover, queries in that case can be answered using bit-wise
-logical operatios (e.g. and, or, not) which can improve query performance \cite{253268}. Another advantage of bitmap indices is the ability to perform more
+logical operations (e.g. and, or, not) which can improve query performance \cite{253268}. Another advantage of bitmap indices is the ability to perform more
 efficient joins by creating bitmap join indices and using those indices to calculate the result of the join \cite{212001}. Although there are many benefits
@@
 passing the attribute (key) value to a hash function that maps it to a starting block address (i.e. bucket address). Afterwards, table insertions will be done
-according to the hash function, as well as quering the records later.
+according to the hash function, as well as querying the records later.
 
@@
 
-Partitioning is an important aspect of physical design as it reduces table scanning time \cite{lightstone2007physical}. There are two main categories of
-partitioning: horizontal partitioning and vertical partitioning \cite{agrawal2004integrating}. Horizontal partitioning divides the table into sets of table rows
-where each row or record still has all the attributes of the table \cite{ceri1982horizontal, shin1985partitioning}. For example, dividing the data by date where
-all data dating less than year 2000 lies in one partition while all data dating more lies in another. On the other hand, vertical partitioning reduces the width
-of the dataset by storing some attributes in one table, and some in another. Usually, less frequently queried columns are partitioned together to reduce the
-size of the table, especially in data warehouses where not all the columns of a table are analyzed for decision support.
+Partitioning in physical database design is a method for reducing the workload on any one hardware component, like an individual disk, by partitioning
+(dividing) the data over several disks \cite{lightstone2007physical}. There are two main categories of partitioning: horizontal partitioning and vertical
+partitioning \cite{agrawal2004integrating}. Horizontal partitioning divides the table into sets of table rows where each row or record still has all the
+attributes of the table \cite{ceri1982horizontal, shin1985partitioning}. For example, dividing the data by date where all data dating less than year 2000 lies
+in one partition while all data dating more lies in another. On the other hand, vertical partitioning reduces the width of the dataset by storing some
+attributes in one table, and some in another. Usually, less frequently queried columns are partitioned together to reduce the size of the table, especially in
+data warehouses where not all the columns of a table are analyzed for decision support.
 
@@
 all current DBMSes \cite{mysqlpartitioning, sqlsrvpartitioning, sybasepartitioning, postgrepartitioning, oraclepartitioning, db2partitioning} which explains
-why there exists many flavors of horizontal paritioning techniques. There are three major types of horizontal partitioning: range partitioning
+why there exists many flavors of horizontal partitioning techniques. There are three major types of horizontal partitioning: range partitioning
 \cite{dewitt1986gamma}, list partitioning \cite{lightstone2007physical} (which is a variation of range partitioning), and hash partitioning
@@
 \begin{table}[!t]
+\centering
 \scalebox{0.7}{
- \begin{tabular}{p{2cm}||p{4cm}|p{4cm}|p{4cm}|p{4cm}}
+ \begin{tabular}{p{1.8cm}||p{4cm}|p{3.6cm}|p{4cm}|p{5cm}}
 & \textbf{Range Partitioning} & \textbf{List Partitioning} & \textbf{Hash Partitioning} & \textbf{Composite Partitioning}\\
@@
 \hline
-\textbf{Benefits} & \multicolumn{4}{c}{Partition elimination, Improved administration, Fast roll-in/out, and Implicit clustering }\\
+\textbf{Benefits} & \multicolumn{4}{p{16cm}}{\centering Partition elimination \linebreak Improved administration \linebreak Fast roll-in/out \linebreak Implicit
+clustering }\\
 \hline
-\textbf{Example} & \begin{itemize}
-                    \item Zip code 
-		    \item Time range
-                   \end{itemize} &
-		    \begin{itemize}
-		    \item Regions
-		    \item States
-		    \end{itemize} &
-		    \begin{itemize}
-		    \item EmployeeId
-		    \item ProductId
-		    \end{itemize} &
+\textbf{Example} & %\begin{itemize} 
+                    \raggedright - Zip code \linebreak
+		    - Time range &
+                  \raggedright
+		    - Regions \linebreak
+		    - States
+		     & \raggedright
+		    - EmployeeId \linebreak
+		    - ProductId &
 		    Hash then List \\
@@
 other stored data, but it is captured in real time and will not necessarily be stored on disk. MMDB have many implementation challenges that have been addressed
-throughout the research community. These challenges will be discussed throughout this section. %Table~\ref{tab:mmdbsumm} summarizes
-the physical design elements of MMDBs.
-
-\begin{table}[t!]
-\centering
-\scalebox{0.9}{
-\begin{tabular}{p{3cm}|p{6cm}|p{6cm}}
-\hline
- \textbf{Physical Design Element} & \textbf{Desired Features} & \textbf{Methods/Structures Used}\\
-\hline\hline
-Index Structures & \begin{itemize}
-                    \item No need to store actual values in index \item Larger node size that is aligned with cache line size for cache consciousness
-                   \end{itemize} & \begin{itemize}
-                                    \item B Tree
-\item B+ Tree
-\item CSB+ Tree
-\item pB+ Tree
-\item T Tree
-\item CST Tree
-\item CSS Tree
-                                   \end{itemize}\\
-\hline
-Materialized Views & \begin{itemize} \item Not highly utilized since processing and memory access is cheap \end{itemize}& N/A\\
-\hline
-Partitioning & \begin{itemize} \item Only needed in secondary storage to speed up reloading in case of a crash \end{itemize} & \begin{itemize}
-                                                                                            \item Horizontal Partitioning
-\item Vertical Partitioning
-                                                                                           \end{itemize}\\
-\hline
-Clustering & \begin{itemize} \item Not needed since random or dispersed main memory access is not more expensive than sequential access \end{itemize}
-& N/A
-
-\end{tabular}
-}
-\caption{Summary of physical design of Main Memory Databases (MMDB)}
-\label{tab:mmdbsumm}
-\end{table}
+throughout the research community. These challenges will be discussed throughout this section.
 
@@
 \centering
-\scalebox{0.6}{
-\begin{tabular}{p{2cm}|p{6cm}|p{6cm}}
+\scalebox{0.8}{
+\begin{tabular}{p{2.2cm}|p{6cm}|p{6cm}}
 \hline
- \textbf{Index Structure} & \textbf{Features} & \textbf{Cache Consciousness}\\
+ \centering \textbf{Index Structure} & \textbf{Features} & \textbf{Cache Consciousness}\\
 \hline\hline
 B Trees & 
-\begin{itemize}\item Good storage utilization
-\item Reasonably quick searching
-\item Fast updating
-\end{itemize} 
-& \begin{itemize}\item Reasonable cache behavior if node fits in cache line\end{itemize}\\
+\raggedright -Good storage utilization \linebreak  \linebreak
+\raggedright -Reasonably quick searching \linebreak \linebreak
+-Fast updating 
+& 
+-Reasonable cache behavior if node fits in cache line\\
 \hline
 B+ Trees &
-\begin{itemize}
-\item Reasonably quick searching
-\item Fast updating
-\end{itemize} &
-\begin{itemize}
-\item Reasonable cache behavior if node fits in cache line\end{itemize}\\
+\raggedright -Reasonably quick searching \linebreak \linebreak
+\raggedright -Fast updating
+&
+-Reasonable cache behavior if node fits in cache line\\
 \hline
-CSB+ Trees & \begin{itemize}
-              \item Same features of B+ Trees
-\item Stores child nodes in an array. Stores pointer to first child only and uses index calculation to get the rest
-             \end{itemize} &
-\begin{itemize}
- \item Good cache performance due to improved locality \& lower tree height (because of larger node size)
-\end{itemize}\\
+CSB+ Trees & 
+-Same features of B+ Trees \linebreak \linebreak
+-Stores child nodes in an array. Stores pointer to first child only and uses index calculation to get the rest
+ &
+-Good cache performance due to improved locality \& lower tree height (because of larger node size)\\
 \hline
 
-pB+ Trees & \begin{itemize}
-              \item Same features of B+ Trees
-             \end{itemize} &
-\begin{itemize}
- \item Improves cache miss performance by prefetching more data in each cache miss
-\item Node size is also increased leading to the same benefits of CSB+ Trees
-\end{itemize}\\
+pB+ Trees & 
+-Same features of B+ Trees
+             &
+-Improves cache miss performance by prefetching more data in each cache miss \linebreak \linebreak
+-Node size is also increased leading to the same benefits of CSB+ Trees\\
 \hline
-T Tree &  \begin{itemize}
-              \item Contain pointers to data values instead of the values themselves leading to better space utilization
-\item Many sorted keys per node
-\item Only minimum and maximum keys used for comparisons
-\item Contains record pointers in every key
-\item Good update \& storage characteristics if Btrees.
-
-             \end{itemize} &
-\begin{itemize}
- \item Cache behavior was not considered at time of design
-\item Record pointers waste space in cache
-\end{itemize}\\
+T Tree &  
+-Contain pointers to data values instead of the values themselves leading to better space utilization \linebreak \linebreak
+-Many sorted keys per node \linebreak \linebreak
+-Only minimum and maximum keys used for comparisons \linebreak \linebreak
+\raggedright -Contains record pointers in every key \linebreak \linebreak
+-Good update \& storage characteristics if Btrees.
+             &
+-Cache behavior was not considered at time of design \linebreak \linebreak
+-Record pointers waste space in cache\\
 \hline
-CST & \begin{itemize}
-       \item Binary search tree for maximum value of each node leads to faster search
-\item More space efficient by removing pointers \& using array storage. Index calculation used to allocate values
-      \end{itemize} &
-\begin{itemize}
- \item Node size is aligned with cache line size to avoid misses
-\item Higher usage of cached data in binary search tree
-\end{itemize}\\
+CST & 
+-Binary search tree for maximum value of each node leads to faster search \linebreak \linebreak
+-More space efficient by removing pointers \& using array storage. Index calculation used to allocate values &
+-Node size is aligned with cache line size to avoid misses \linebreak \linebreak
+-Higher usage of cached data in binary search tree\\
 \hline
-CSS Trees &  \begin{itemize}
-       \item Fast traversal in $log_{m+1}n$ ($m$ is the number of keys per node)
-\item Mainly suitable for read environments
-      \end{itemize} &
-\begin{itemize}
- \item Good cache behavior since $m$ is chosen to fit in the cache line size
-\end{itemize}\\
+CSS Trees & 
+-Fast traversal in $log_{m+1}n$ ($m$ is the number of keys per node) \linebreak \linebreak
+-Mainly suitable for read environments &
+-Good cache behavior since $m$ is chosen to fit in the cache line size\\
 
@@
 management), financial services (for claims analysis, risk analysis, credit card analysis, and fraud detection), transportation (for fleet management),
-telecommunications (for call analysis and fraud detection), utilities (for power usage analysis), and healthcare (for outcomes analysis)
+telecommunications (for call analysis and fraud detection), utilities (for power usage analysis), and health care (for outcomes analysis)
 \cite{chaudhuri1997overview}.
@@
 Although data in a data warehouse is extracted and loaded from multiple on-line transaction processing (OLTP) data sources (including DB2, Oracle, IMS
-databases, and flat files) using Extract, Transfer, and Load (ETL) tools (see Figure \ref{fig:dw}), a data warehouse is usually maintained seperately from the
+databases, and flat files) using Extract, Transfer, and Load (ETL) tools (see Figure \ref{fig:dw}), a data warehouse is usually maintained separately from the
 organization's operational databases \cite{sen2005comparison, chaudhuri1997overview}. This architecture can be justified owing to the fact that operational
@@
 Just like all other research areas are moving to the cloud, database systems are moving there too. Accordingly, lots of current research focuses on adapting
-storage techniques and index structures to the cloud. For example, Wang et al. propose a new index structure for indexing mult-dimensional data in the
+storage techniques and index structures to the cloud. For example, Wang et al. propose a new index structure for indexing multi-dimensional data in the
 cloud~\cite{wang2010indexing}. Lots of work is currently being done in this area. For example, Wu et al.~\cite{wu2010efficient} propose a modified B+tree index
@@
 
+\begin{table}[!t]
+\centering
+\scalebox{0.7}{
+\begin{tabular}{p{2cm}||p{5cm}|p{4cm}|p{4cm}|p{4cm}}
+ & \textbf{Index Structures} & \textbf{Materialized Views} & \textbf{Partitioning} & \textbf{Clustering}\\
+\hline\hline
+\textbf{RDBMS}& E.g. B+trees, Bitmap Index, Hash Tree & Store results of common complicated queries on disk & Horizontal Partitioning (range, list, hash,
+composite) Vertical Partitioning& Group related items together on disk\\
+\hline
+\textbf{MMDB} & 
+-No need to store actual data values in index \linebreak \linebreak 
+-Larger node size that is aligned with cache line size \linebreak \linebreak
+-E.g. B+tree, CSB+tree, T tree, CST tree & Not highly utilized since processing and main memory access is cheap & Only needed in secondary storage to speed up
+reloading in case of a crash\linebreak \linebreak 
+-Horizontal \& Vertical partitioning& 
+Not needed since random access in main memory costs the same as sequential access\\
+\hline
+\textbf{XML DB} & Need a number scheme. E.g. B+tree and variations used. & XQuery and XPath results are materialized & Horizontal partitioning based on inlining
+or node
+type & Based on document order or structural similarity\\
+\hline
+\textbf{Data Warehouses} & E.g. Universal B+tree, Bitmap index, Projection Index & 
+-Essential to improve performance \linebreak \linebreak 
+-Not applicable to adhoc queries & 
+-Horizontal partitioning based on time dimension \linebreak \linebreak 
+-Vertical partitioning of seldom used columns. & Data is
+clustered by
+nature.\\
+\end{tabular}
+}
+\caption{Summary of the different elements of physical design in the studied systems}
+\label{tab:summary}
+\end{table}
+
 \section{Conclusions}
@@
 
-%\bibliographystyle{IEEEtran}
+In this survey, we present four elements of physical design in database: index structures, materialized views, partitioning, and clustering. We started out
+by explaining how these design factors are implemented in traditional Relational Database Management Systems (RDBMS), and then presented how they are used in
+other information systems. We focused on three types of information systems: Main Memory Database Systems (MMDB), XML Databases, and Data Warehouses. We
+presented the challenges involved with each type of system, and how each of the four physical design elements were adapted or utilized to overcome these
+challenges. Table~\ref{tab:summary} summarizes the results of our survey. From our literature review, we found that the B+tree is the most utilized index
+structure, and has been adapted and customized to fit the different needs of other systems. For example, it was made more cache conscious to fit MMDB needs,
+and was changed to apply to multidimensions to fit the Data Warehouses needs. Materialized views were found to be heavily utlitized in all the systems we
+presented, except in MMDBs. Partitioning was also used in all systems, but was only used for data reloading in MMDBs. Horizontal partitioning is much more
+common that vertical partitioning. Clustering is very important in all systems, except for MMDBs. Clustering is especially very important for Data Warehouses
+where huge amounts of data are stored.
+
+It was interesting to see how the different nature of these systems affects the physical design. Throughout the survey, we have showed some of the
+commonalities between the different systems. Moreover, understanding how to meet these needs would help us design mash-ups of these different systems. For
+example, just like relational database systems shifted into main memory, we might have an XML main memory database system. In that case, the index structures
+fitted for XML data indexing will have to be adapted for main memory by making sure they are aligned with the cache line size, for example. Similarly, if XML
+data warehouses are developed, more elaborate path indexing techniques and more clustering would be required.
 

### 2010-07-26 trunk/cs848-db/Paper.tex
integrate.

@@
 
-Bitmap indices are best used when the set of possible key values in the index is small with a large number of rows, e.g. gender attribute \cite{212001}.
+Bitmap indices are best used when the set of possible key values in the index is small with a large number of rows, e.g. gender attribute \cite{oneil1995multi}.
 Therefore, bitmap indices outperform B+trees for low cardinality attributes. The reason is that given an attribute with low cardinality, defining a B+tree index
@@
 A data warehouse is a subject-oriented, integrated, time-variant, and non-volatile collection of data and decision support technologies, aimed at enabling the
-knowledge worker (executive, manager, analyst) to make better and faster decisions \cite{560407, chaudhuri1997overview}. More than half of IT executives named
-Data warehousing as the highest-priority post-millennium project for them \cite{sen2005comparison}. The value of data warehousing for an organization depends on
-the organization's need for reliable, consolidated, unique and integrated reporting and analysis of its data, at different levels of aggregation.
+knowledge worker (executive, manager, analyst) to make better and faster decisions \cite{inmon2002building, chaudhuri1997overview}. More than half of IT
+executives named Data warehousing as the highest-priority post-millennium project for them \cite{sen2005comparison}. The value of data warehousing for an
+organization depends on the organization's need for reliable, consolidated, unique and integrated reporting and analysis of its data, at different levels of
+aggregation.
 
@@
 management), financial services (for claims analysis, risk analysis, credit card analysis, and fraud detection), transportation (for fleet management),
-telecommunications (for call analysis and fraud detection), utilities (for power usage analysis), and health care (for outcomes analysis)
+telecommunications (for call analysis and fraud detection), utilities (for power usage analysis), and healthcare (for outcomes analysis)
 \cite{chaudhuri1997overview}.
@@
 Data warehouses tend to be extremely large, in fact it is quite possible for a data warehouse to store tens of petabytes of data while loading tens of terabytes
-of data everyday \cite{thusoo2010data}. Datta et al. \cite {628286} notes that the information in a data warehouse is usually multidimensional in nature,
-requiring the capability to view the data from a variety of perspectives. Aggregated and summarized data become more crucial than detailed records in such
-environment. Therefore, the workloads are query intensive with mostly ad hoc, complex queries, often requiring computationally expensive operations such as
+of data everyday \cite{thusoo2010data}. Datta et al. \cite {datta2002parallel} notes that the information in a data warehouse is usually multidimensional in
+nature, requiring the capability to view the data from a variety of perspectives. Aggregated and summarized data become more crucial than detailed records in
+such environment. Therefore, the workloads are query intensive with mostly ad hoc, complex queries, often requiring computationally expensive operations such as
 scans, joins, and aggregation. Performing such operations on large amounts of data, like in the case of data warehousing, complicates the situation further.
@@
 Although data in a data warehouse is extracted and loaded from multiple on-line transaction processing (OLTP) data sources (including DB2, Oracle, IMS
-databases, and flat files) using Extract, Transfer, and Load (ETL) tools (see Figure \ref{fig:dw}), a data warehouse is usually maintained separately from the
+databases, and flat files) using Extract, Transfer, and Load (ETL) tools (Figure \ref{fig:dw}), a data warehouse is usually maintained seperately from the
 organization's operational databases \cite{sen2005comparison, chaudhuri1997overview}. This architecture can be justified owing to the fact that operational
@@
 
-\cite{Cheung20011}
-ROLAP
-          o Data is stored in relational databases.
-          o support for multi-dimensional view of data is achieved thru star scheme
-          o Support extensions to SQL.
-          o Efficiently implement the multidimensional data model and operations.
-          o Ex: Microstrategy.
-MOLAP
-          o Directly store multidimensional data in special data structures (arrays).
-          o Implement the OLAP operations over these special data structures.
-          o Snowflake Schema. \cite{kimball2009data}
-          o Fact Constellations.
-          o Ex: Essbase (Arbor), statistical DBMSes.
+The design of a data warehouse can follow one of two paradigms: relational-based OLAP (ROLAP) or multidimensional OLAP (MOLAP). In ROLAP, data is stored in
+relational databases and the support for multidimensional data is achieved through the use of the star schema \cite{cheung2001towards}. A star schema (Figure
+\ref{fig:starschema}) is a database schema where a fact table $T$ that holds the base data in the database. Other tables in the database are dimensions for
+table $T$, where each dimension table is linked to the fact table through foreign keys in $T$ pointing to the primary keys in the dimension tables. Thus, the
+fact table, which is always the largest table in such schema, participates in almost all join operations \cite{datta2002parallel}. This might enforce some
+design decisions for which indices to use, whether materialized views should be implemented to reduce the overhead of such expensive join operation or not. In
+MOLAP, multidimensional data is directly stored in special data structures (arrays) forming data cubes where each side of the cube reflect a dimension of the
+base data. OLAP operations are implemented on top of these special data structures. Usually a snowflake schema \cite{kimball2009data} or fact constellations are
+used to logically design the data warehouse.
 
-hierarchical data clusters \cite{karayannidis2008hierarchical}.
+The rest of this section will discuss how such challenges and requirements of data warehouse design affect the selection of physical design elements during the
+design process.
 
-star schema
-fact table
-
-Typical Operations
-    * Data Cleaning
-    * Load
-    * Refresh
-
-Joins in data warehouses are usually expensive since the fact table (the largest one) participates in every join \cite{628286}
+\begin{figure}[!t]
+\centering
+\includegraphics[width=9cm]{figs/starschema.png}
+\caption{An example of a star schema}
+\label{fig:starschema}
+\end{figure}
 
 \subsubsection{Index Structures}
-\paragraph{Tree-based Indexes}
-B-tree (B+ trees but commonly referred to as B-trees) \cite{253268}, R-tree \cite{602266, Cheung20011}, K-D-B-tree \cite{582321}, BV-tree \cite{223796}, UB-tree
-\cite{bayer1997universal}
+Despite the design challenges involved in data warehousing, some relational index structures are still used after adapting them to match those challenges.
 
-\paragraph{Bitmapped Join Indexes}
-\cite{212001}
+\paragraph{Tree-based Indices}
+The widely used B-trees have been modified to suit the requirements of a data warehouse design, such modification resulted in various tree-based indices:
+UB-trees, BV-trees, K-D-B-trees, and R-trees. A UB-tree index \cite{bayer1997universal} is a variation of B+trees where support for multidimensional data is
+provided by linearizing such data to be able to store it in one dimension. Usually the Z-order (Morton-order) \cite{morton1966computer} is used to perform this
+type of linearization. UB-trees inherit the advantages of B+trees, so they are well suited for high cardinality attributes. They are also excellent for point
+and interval queries (using the pointer in each leaf node to the next leaf node for faster index traversal). UB-trees also inherit the disadvantages of
+B+trees. Therefore, they poorly perform for low cardinality attributes.
 
-\paragraph{Projection Indexes}
-Involves positional indexing where tuples are accessed based on their ordinal position \cite{628286}
+A BV-tree \cite{freeston1995general} is an unbalanced tree that maintains the characteristics of the B-tree in $n$ dimensions, and it can be degenerated to a
+balanced tree in the one-dimensional case. A BV-tree index was an attempt to solve the problem of generalizing the characteristics of a B-tree index in $n$
+dimensions. However, this type of indices did not really catch up in the literature of data warehouses since it has a very grave disadvantage. The
+design of a BV-tree index suggests that the size of the index can span multiple-page for very large files (which is the case in data warehouses). Since
+searching a BV-tree index might involve backtracking \cite{freeston1995general}, such property is not desirable.
 
-\paragraph{Bit-sliced Indexes}
+K-D-B-trees \cite{robinson1981kdb} is another type of tree-based indices that tries to combine the multidimensional search efficiency of K-D-trees
+\cite{bentley1975multidimensional} and the I/O efficiency of B-trees \cite{bayer1970organization}.
+
+Another widely used tree-based index is R-tree \cite{guttman1984rtrees}. An R-tree is an extension of a B-tree with index records in its leaf nodes
+containing pointers to data objects. Nodes in an R-tree correspond to disk pages if the index is disk-resident. The structure of an R-tree is designed so that a
+spatial search requires visiting only a small number of nodes. The index is completely dynamic; inserts and deletes can be intermixed with searches and no
+periodic reorganization is required. R-trees support multidimensional spaces by storing an n-dimensional rectangle, which is the bounding box of the spatial
+object indexed, in the index record entries \cite{guttman1984rtrees}. R-trees were extended by Norbert et al \cite{norbert1990rstree} by designing the R*-tree
+to support both sparse and dense data at the same time with a slightly higher cost (during insertion) than R-trees. Cheung et al. \cite{cheung2001towards}
+describe how R-trees can be used to index dense data regions (MOLAP data cubes), while having pointers from the leaf nodes to sparse data points (ROLAP).
+
+\paragraph{Bitmap Join Index}
+Bitmap join indices are another widely used type of indices in data warehouses epecially when a star schema is used. A bitmap join index is simply a bitmap
+index that is used for calculating the product of the join of the dimension tables. The result is then joined with the fact table to answer the given query
+\cite{oneil1995multi}. In fact such process is less costly than joining the fact table with the dimension tables right away. The reason is that the fact table
+is usually larger than the product of the dimension tables. The disadvantage of bitmap join being not amenable to index updates is not really applicable to the
+case of data warehousing since a data warehouse can be considered as a read-mostly database.
+
+\paragraph{Projection Index}
+Not all columns in a data warehouse are of interest to the given query workloads. Therefore, it makes sense to have faster access to the columns that are used
+the most to improve the query performance. That is why Projection indices are often used in data warehousing. A projection index involves positional indexing
+of a column (form of vertical partitioning) where tuples are accessed based on their ordinal position \cite{datta2002parallel}. More results per I/O operation
+is a direct benefit of using projection indices, since only data of interest are accessed and retrieved. However, only raw data can be fetched (e.g. column
+list in a selection clause). A projection index can be further extended into a bit-sliced index \cite{oneil1997improved} where the projection index is sliced
+vertically into columns of bits. Each column is a bit-sliced index. Bit-sliced indices can help reduce storage costs by using a number of bits per value which
+is not a power of 2. For example, a SALES column contains all sales made in the last month, in pennies. Assuming no item cost more than $2^{20} - 1$ pennies (=
+\$10485.75), this is 20 bit-slice indices). Moreover, computing aggregates, checking predicates can be done using efficient base-2 arithmetic operation. This
+greatly enhances query performance since aggregates are crucial for data warehouses.
+
+\subsubsection{Materialized Views}
+not very useful since most queries are ad-hoc \cite{armstrong1997data, theodoratos2000general}
 
@@
 \paragraph{Horizontal Partitioning}
-
-\subsubsection{Materialized Views}
-not very useful since most queries are ad-hoc \cite{653447}
-\cite{355309}
 

### 2010-07-26 cs848-db/Paper.tex

@@
 \begin{abstract}
-The physical design decisions taken during database design highly affect the database performance. These design decisions include the index structures used,
+The physical design decisions taken during the database design lifecycle highly affect the database performance. These design decisions include the index
+structures used,
 which views can be materialized, how to partition the data etc. These decisions vary by the nature of the system being implemented. For example, the index
@@
 it to other types of information systems. Specifically, we look at the physical design of Main Memory Database Systems (MMDB), XML Databases, and Data
-Warehouses. We examine how the specific nature of each of these systems affect its physical design. 
+Warehouses. We examine how the specific nature of each of these systems affect its physical design. We focus on index structures, materialized views,
+partitioning, and clustering.
 \end{abstract}
@@
 This survey presents the different physical design patterns that have been used in different types of information systems. The systems discussed in this survey
-are disk based relational databases (traditional database systems), main memory relational databases, data warehouses, and XML databases. The aim of this survey
-is not to explain the details of every data structure used in physical design, but rather to compare how the different data structures have been used in these
+are disk-based Relational Database Management Systems (RDBMS - traditional database systems), Main Memory Databases (MMDB), Data Warehouses, and XML
+Databases. The aim of this survey
+is not to explain the details of every data structure used in physical design, but rather to compare how the different data structures and techniques have been
+used in these
 systems. This includes explaining how alterations or additions are made to data structures to fit the specific needs of every system. Using the different
@@
 disk-based databases. Section \ref{SEC-DIFFSYS} then discusses how each of the three other systems uses these elements, as applicable. Sections \ref{SEC-MMDB},
-\ref{SEC-XML}, \ref{SEC-WAREHOUSES}, discuss Main Memory Databases, XML Databases, and Data Warehouses respectively. Throughout these sections, comparisons are
-made as to how the different elements have been modified to fit the needs of each system. Section \ref{SEC-OPEN} mentions some of the open problems of physical
-database design, and suggests possible solutions. Finally, Section \ref{SEC-CONCL} summarizes and concludes this survey.
+\ref{SEC-XML}, \ref{SEC-WAREHOUSES} discuss MMDBs, XML Databases, and Data Warehouses respectively. Throughout these sections, comparisons are
+made as to how the different elements have been modified to fit the needs of each system. Section \ref{SEC-OPEN} mentions some of the currently researched open
+problems in physical database design. Section \ref{SEC-CONCL} then summarizes and concludes this survey.
 
@@
 Relational Databases were first introduced by Codd \cite{codd1970relational}. Disk based Relational Database Management Systems (RDBMS) are the most common and
-popular type of Information Systems. Since they have been extensively researched and have been around for many decades, we use them as the basis of our
+popular type of information systems. Since they have been extensively researched and have been around for many decades, we use them as the basis of our
 comparison. In this section, we describe the different elements of physical design in RDBMS. These elements will also be examined in the other systems to see
@@
 
-The first physical database design decision to be considered is the choice of indexes to be implemented. The concept of indices has been around for a long time.
-Just like and index at the end of the book is provided to help find certain content faster, indices in an information system help find content in tables faster.
+The first physical database design decision to be considered is the choice of indices to be implemented. The concept of indices has been around for a long time.
+Just like an index at the end of the book is provided to help find certain content faster, indices in an information system help find content in tables faster.
 The main idea of an index is to have each record in a table have a unique identifier (primary key) and then organize these identifiers in a certain way that
 allows for fast access of a specific record. Of course, indexing can be done on any column in the database and not necessarily the primary key. There are many
-type of indexes that have been used in information systems.
+type of indices that have been used in information systems.
 
-Btrees (short for Balanced Trees) are the most used database index structure.  The Btree is essentially similar to the traditional Binary Search Tree, but
-instead of having one value per node, a B-tee can have many values per node \cite{comer1979ubiquitous}. B-trees are suitable for relational databases since the
+B-trees (short for Balanced Trees) are the most used database index structure.  The B-tree is essentially similar to the traditional Binary Search Tree, but
+instead of having one value per node, a B-tree can have many values per node \cite{comer1979ubiquitous}. B-trees are suitable for relational databases since the
 cost of retrieval in a B-tree is at most proportional to: $log_{d}\frac{n+1}{2}$. The cost of insertion and deletion is at most proportional to $log_{d}n$ due
@@
 Different variations of B-trees have been used. For example, B+trees are now the main method of indexing in current disk-based Relational Databases
-\cite{lightstone2007physical} such as DB2, Oracle, and SQL Server. The main difference between the B+tree and the Btree is that only leaf nodes contain data
-pointers in a B+tree, and leaf nodes contain pointers to each other as shown in Figure \ref{fig:b+tree}. In Prefix B+trees, instead of using the actual key
+\cite{lightstone2007physical} such as DB2, Oracle, and SQL Server. The main difference between the B+tree and the B-tree is that only the leaf nodes contain
+data pointers in a B+tree, and leaf nodes contain pointers to each other as shown in Figure \ref{fig:b+tree}. In Prefix B+trees, instead of using the actual key
 value of a record, a prefix from the key value is used to decrease the storage size needed which will in turn allow more records to be stored per node thus
@@
 
-Bitmap indices represent another type of index structures that is commonly used and cited in the literature. Using bitmaps as indexes first appeared in the
-literature as part of the architecture of the Model 204 commercial database of Computer Corporation of America for the IBM 370 computer \cite{patrick1987model}.
+Bitmap indices represent another type of index structures that is commonly used and cited in the literature. Using bitmaps as an index structure was first
+implemented as part of the architecture of the Model 204 commercial database of Computer Corporation of America for the IBM 370 computer
+\cite{patrick1987model}.
 A bitmap index structure is based on a bitmap (i.e. bitset or bit array) data structure where individual bits correspond to values in unique records. Given a
-table column $C$ where a record $x$ can have one of six unique values, a traditional bitmap indexe defined for $C$ generates one bitmap vector for each of the
+table column $C$ where a record $x$ can have one of six unique values, a traditional bitmap index defined for $C$ generates one bitmap vector for each of the
 six different values ($B0-B5$). If the value of $C$ for record $x$ is set to 3, the corresponding bit in vector $B3$ is set to 1. Otherwise the bit is set to 0.
-
-Bitmap indices are best used when the set of possible key values in the index is small with a large number of rows, e.g. gender attribute \cite{212001}.
-Therefore, bitmap indices outperform B+trees for low cardinality attributes. The reason is that given an attribute with low cardinality, defining a B+tree index
-on it means that the tree will have very few leaf nodes, each one storing a very long list of record identifiers (RIDs). On the other hand, defining a
-bitmapindex on that same attribute means that such long lists or RIDs will not be stored, rather bit vectors that represent the distinct attribute values will
-be stored as previously explained. This leads to a direct benefit of consuming less storage space. Moreover, queries in that case can be answered using bit-wise
-logical operatios (e.g. and, or, not) which can improve query performance \cite{253268}. Another advantage of bitmap indices is the ability to perform more
-efficient joins by creating bitmap join indices and using those indices to calculate the result of the join \cite{212001}. Although there are many benefits
-gained from using bitmap indices, they poorly perform during updates. The reason is that when an updates is to be done to a bitmap index, the whole bit vector
-for the index is locked, blocking any access to the index for other operations, then the lock is released when the update is done. That is why bitmap indices
-are usually used for read-mostly databases, a famous example of which are data warehouses (Section \ref{SEC-WAREHOUSES}).
-
-Another variation of indexing, a hash table index, often simply known as a hash index is used to organize unordered physical database tables for efficient
-access \cite{lightstone2007physical}. This type of index is usually defined for keys that have unique values in the data records. Creating this index requires
-passing the attribute (key) value to a hash function that maps it to a starting block address (i.e. bucket address). Afterwards, table insertions will be done
-according to the hash function, as well as quering the records later.
-
-In general, indices are a way of organizing the logical structure of a database without altering its physical structure. However, sometimes it is desirable to
-store related records near each other on disk for better locality. This can be achieved through implementing a clustered index on the desired key values.
-Otherwise, the created index is known as non-clustered index. The primary advantage of a clustered index is therefore the ordering of the physical data rows in
-accordance with the index blocks that point to them which. Therefore, fewer data block reads are required to answer a given query. The only caveat with using a
-clustered index is that there can only be one clustered index per table because the physical organization of the table must be fixed
-\cite{lightstone2007physical}.
 
@@
 
+Bitmap indices are best used when the set of possible key values in the index is small with a large number of records, e.g. gender attribute
+\cite{oneil1995multi}. Bitmap indices outperform B+trees for low cardinality attributes. The reason is that given an attribute with low cardinality, defining a
+B+tree index
+on it means that the tree will have very few leaf nodes, each one storing a very long list of record identifiers (RIDs). On the other hand, defining a
+bitmap index on that same attribute means that such long lists or RIDs will not be stored, rather bit vectors that represent the distinct attribute values will
+be stored as previously explained. This leads to a direct benefit of consuming less storage space. Moreover, queries in that case can be answered using bit-wise
+logical operations (e.g. and, or, not) which can improve query performance \cite{oneil1997improved}. Another advantage of bitmap indices is the ability to
+perform more
+efficient joins by creating bitmap join indices and using those indices to calculate the result of the join \cite{oneil1995multi}. Although there are many
+benefits
+gained from using bitmap indices, they poorly perform during updates. The reason is that when an update is to be executed to a bitmap index, the whole bit
+vector
+for the index is locked, blocking any access to the index for other operations. The lock is then released when the update is done. That is why bitmap indices
+are usually used for read-mostly databases, a famous example of which are data warehouses (Section \ref{SEC-WAREHOUSES}).
+
+Another variation of indexing, a hash table index, often simply known as a hash index is used to organize unordered physical database tables for efficient
+access \cite{lightstone2007physical}. This type of index is usually defined for keys that have unique values in the data records. Creating this index requires
+passing the attribute (key) value to a hash function that maps it to a starting block address (i.e. bucket address). Afterwards, table insertions will be done
+according to the hash function, as well as querying the records later.
+
+In general, indices are a way of organizing the logical structure of a database without altering its physical structure. However, sometimes it is desirable to
+store related records near each other on disk for better locality. This can be achieved through implementing a clustered index on the desired key values.
+Otherwise, the created index is known as a non-clustered index. The primary advantage of a clustered index is therefore the ordering of the physical data rows
+in accordance with the index blocks that point to them. Therefore, fewer data block reads are required to answer a given query. The only caveat with using a
+clustered index is that there can only be one clustered index per table because the physical organization of the table must be fixed
+\cite{lightstone2007physical}.
+
 \subsection{Materialized Views}
@@
 simply store the results of the queries on disk instead of recomputing them each time. Oracle 8i introduced materialized views to fit this purpose. By
-precalculating the results of a complex query, and storing them in a table on disk, the new table with the results is very likely to be much smaller than the
-original tables decreasing I/O access costs and thus increasing performance \cite{lightstone2007physical, chaudhuri1998overview}. Once a materialized view
-exists, the user can either explicitly query the materialized view, or enter the original query, and have the query optimizer rewrite the query to use the
-materialized views instead of the original tables. Many query optimization techniques have relied on rewriting queries using views (e.g.
+precalculating the results of a complex query, and storing them in a table on disk, this new table containing the query results will be much
+smaller than the original tables. This decreases I/O access costs and thus increases performance \cite{lightstone2007physical, chaudhuri1998overview}. Once a
+materialized view exists, the user can either explicitly query the materialized view, or enter the original query, and have the query optimizer rewrite the
+query to use the materialized views instead of the original tables. Many query optimization techniques have relied on rewriting queries using views (e.g.
 \cite{levy1995answering, gupta1995aggregate, goldstein2001optimizing, abiteboul1998complexity}), but discussing the details of the rewriting process is beyond
@@
 
-Partitioning in physical database design is a method for reducing the workload on any one hardware component, like an individual disk, by partitioning
+Partitioning in physical database design aims to reduct the workload on any one hardware component (E.g. an individual disk). This is achieved by partitioning
 (dividing) the data over several disks \cite{lightstone2007physical}. There are two main categories of partitioning: horizontal partitioning and vertical
@@
 in one partition while all data dating more lies in another. On the other hand, vertical partitioning reduces the width of the dataset by storing some
-attributes in one table, and some in another. Usually, less frequently queried columns are partitioned together to reduce the size of the table, especially in
-data warehouses where not all the columns of a table are analyzed for decision support.
+attributes in one table, and some in another. Usually, less frequently queried columns are partitioned together to reduce the size of the table. We will see in
+Section~\ref{SEC-WAREHOUSES} that vertical partitioning is frequently used in Data Warehouses where not all the columns of a table are analyzed for decision
+support.
 
@@
 to. Combining partitioning and indexing can often lead to better performance. For example, if List partitioning is used to divide a table by regions and then a
-bitmap index is used on that table, finding records in the same region or state can be much faster. Finally, partitioning the data into different physical
+bitmap index is used on that table, finding records in the same region or state can be much faster. Finally, partitioning is one of the important
+methods used to achieve parallelism. Partitioning the data into different physical
 devices to achieve parallelism when accessing it can greatly increase the performance of the database.
@@
 
-On the other hand, defining clustering keys, clustering scheme, and the granularity of clustering which makes the problem of designing an optimal clustering in
-this context has combinatorial complexity, having an extremely large search space \cite{jagadish1999snakes}.
+On the other hand, defining clustering keys, clustering scheme, and the granularity of clustering is very difficult. Accordingly, the problem of
+designing an optimal clustering scheme has combinatorial complexity, having an extremely large search space \cite{jagadish1999snakes}.
 
@@
 
-There are other physical design methods to improve the performance of the information systems~\cite{lightstone2007physical}. For example, data compression can
+There are other physical design methods used to improve the performance of the information systems~\cite{lightstone2007physical}. For example, data compression
+can
 be used to make more data fit into a fixed amount of disk space such that it can be access faster. However, compression is only beneficial if the cost of
-executing the compression algorithm is lower than the I/O cost of accessing the data. Otherwise, compressing the data must just add an overhead without
+executing the compression algorithm is lower than the I/O cost of accessing the data. Otherwise, compressing the data might just add an overhead without
 improving performance. Data striping is used to distribute data that is accessed together across multiple disks. This allows the data
@@
 which involves duplicating the data on multiple disks. Finally, refining the global schema to reflect query and transaction requirements is also sometimes used.
-This is called denormalization which can be thought of as manual clustering of the data to improve performance.
+This is called denormalization which can be thought of as manual clustering of the data by hand to improve performance.
 
@@
 A Main Memory Database (MMDB) (sometimes referred to as in-memory database) is one where the primary copy of the data resides in the main memory of the system
-rather than on a
-disk~\cite{garcia1992main}. It should be noted that this is different from caching. Disk based database systems use the main memory to cache query results.
-However, MMDB's primary copy of the data resides in main memory. Accessing data in main memory is faster than disk, and so MMDBs usually have better
-performance that disk based systems. However, main memory is more expensive and the size is limited in comparison to disk. However, MMDBs are essential to use
-where some of the collected data cannot, or will not, be stored on disk in the first place. For example, phone switch data may be needed for reasoning with
-other stored data, but it is captured in real time and will not necessarily be stored on disk. MMDB have many implementation challenges that have been addressed
-throughout the research community. These challenges will be discussed throughout this section. Table~\ref{tab:mmdbsumm} summarizes
-the physical design elements of MMDBs.
-
-\begin{table}[t!]
-\centering
-\scalebox{0.9}{
-\begin{tabular}{p{3cm}|p{6cm}|p{6cm}}
-\hline
- \textbf{Physical Design Element} & \textbf{Desired Features} & \textbf{Methods/Structures Used}\\
-\hline\hline
-Index Structures & \begin{itemize}
-                    \item No need to store actual values in index \item Larger node size that is aligned with cache line size for cache consciousness
-                   \end{itemize} & \begin{itemize}
-                                    \item B Tree
-\item B+ Tree
-\item CSB+ Tree
-\item pB+ Tree
-\item T Tree
-\item CST Tree
-\item CSS Tree
-                                   \end{itemize}\\
-\hline
-Materialized Views & \begin{itemize} \item Not highly utilized since processing and memory access is cheap \end{itemize}& N/A\\
-\hline
-Partitioning & \begin{itemize} \item Only needed in secondary storage to speed up reloading in case of a crash \end{itemize} & \begin{itemize}
-                                                                                            \item Horizontal Partitioning
-\item Vertical Partitioning
-                                                                                           \end{itemize}\\
-\hline
-Clustering & \begin{itemize} \item Not needed since random or dispersed main memory access is not more expensive than sequential access \end{itemize}
-& N/A
-
-\end{tabular}
-}
-\caption{Summary of physical design of Main Memory Databases (MMDB)}
-\label{tab:mmdbsumm}
-\end{table}
+rather than on a disk~\cite{garcia1992main}. It should be noted that this is different from caching. Disk-based database systems use the main memory to cache
+query results. However, MMDB's primary copy of the data resides in main memory. Accessing data in main memory is faster than disk, and so MMDBs usually have
+better performance that disk based systems. On the other hand, main memory is more expensive and the size is limited in comparison to disk. MMDBs are
+essential to use when some of the collected data cannot, or will not, be stored on disk in the first place. For example, phone switch data may be needed for
+reasoning with other stored data, but it is captured in real time and will not necessarily be stored on disk. MMDBs have many implementation challenges that
+have been addressed throughout the research community. These challenges will be discussed throughout this section.
 
@@
 designing index structures for MMDB. The first is that the data resides in main memory and not on disk, and so many of the
-considerations taken for I/O operations is no longer there. Instead of focusing on disk access and disk storage, the data structures for main memory
+considerations taken for I/O operations are no longer there. Instead of focusing on disk access and disk storage, the data structures for main memory
 databases should focus on the efficient use of CPU cycles and memory~\cite{lehman1986study}. The second is having index structures cache conscious. Cache
@@
 at processor speed. Otherwise, the data has to be fetched from main memory. With the big difference between processor speed and main memory speed, cache misses
-are still relatively expensive. Lots of research work has shown that cache performance is very critical in MMDBs~\cite{boncz1999database,rao1999cache}. This is
-because the big difference between processor speed and main memory access speed. Thus, a cache miss is still relatively expensive in main memory databases.
-Accordingly, we will see that many of the index data structures proposed for main memory databases focused on being cache conscious or cache sensitive. In
+are still relatively expensive. Lots of research work has shown that cache performance is very critical in MMDBs~\cite{boncz1999database,rao1999cache}. Thus, a
+cache miss is still relatively expensive in main memory databases.
+Accordingly, we will see that many of the index data structures proposed for MMDBs focused on being cache conscious or cache sensitive. In
 summary, a main memory index should be able to reduce overall computation time while using as little memory as possible~\cite{lehman1986study}. It should also
@@
 
-In this section, we talk about three families of index structures used in MMDB: B Trees, T Trees, and Binary Search Trees. There are more recent index
-structures developed for MMDB such as J+ Trees~\cite{luan2009prefetching}, but which have not gained much popularity, and so we do not discuss them in our
-survey. Table~\ref{tab:mmdbindexsumm} summarizes the features of the different index structures discussed in this section with respect to MMDB. The T Tree is
-the most used index for MMDB (E.g. MySQL CLuster). However, there are other indexes used as well. For example, IBM's solidDB uses Trie (or prefix tree) for its
+In this section, we talk about three families of index structures used in MMDB: B-trees, T-trees, and Binary Search Trees. There are more recent index
+structures developed for MMDB such as J+trees~\cite{luan2009prefetching}, but which have still not gained much popularity, and so we do not discuss them in our
+survey. Table~\ref{tab:mmdbindexsumm} summarizes the features of the different index structures discussed in this section with respect to MMDBs. The T-tree is
+the most used index for MMDBs (E.g. MySQL CLuster). However, there are other indexes used as well. For example, IBM's solidDB uses Trie (or prefix tree) for its
 indexing. For their purposes, it served as a good index for main memory since it eliminates the need for many key comparisons~\cite{ibmsoliddb}.
@@
 \centering
-\scalebox{0.6}{
-\begin{tabular}{p{2cm}|p{6cm}|p{6cm}}
+\scalebox{0.8}{
+\begin{tabular}{p{2.2cm}|p{6cm}|p{6cm}}
 \hline
- \textbf{Index Structure} & \textbf{Features} & \textbf{Cache Consciousness}\\
+ \centering \textbf{Index Structure} & \textbf{Features} & \textbf{Cache Consciousness}\\
 \hline\hline
-B Trees &
-\begin{itemize}\item Good storage utilization
-\item Reasonably quick searching
-\item Fast updating
-\end{itemize} 
-& \begin{itemize}\item Reasonable cache behavior if node fits in cache line\end{itemize}\\
+B-tree & 
+\raggedright -Good storage utilization \linebreak  \linebreak
+\raggedright -Reasonably quick searching \linebreak \linebreak
+-Fast updating 
+& 
+-Reasonable cache behavior if node fits in cache line\\
 \hline
-B+ Trees & 
-\begin{itemize}
-\item Reasonably quick searching
-\item Fast updating
-\end{itemize} &
-\begin{itemize}
-\item Reasonable cache behavior if node fits in cache line\end{itemize}\\
+B+tree &
+\raggedright -Reasonably quick searching \linebreak \linebreak
+\raggedright -Fast updating
+&
+-Reasonable cache behavior if node fits in cache line\\
 \hline
-CSB+ Trees & \begin{itemize}
-              \item Same features of B+ Trees
-\item Stores child nodes in an array. Stores pointer to first child only and uses index calculation to get the rest
-             \end{itemize} &
-\begin{itemize}
- \item Good cache performance due to improved locality \& lower tree height (because of larger node size)
-\end{itemize}\\
+CSB+tree & 
+-Same features of B+ Trees \linebreak \linebreak
+-Stores child nodes in an array. Stores pointer to first child only and uses index calculation to get the rest
+ &
+-Good cache performance due to improved locality \& lower tree height (because of larger node size)\\
 \hline
 
-pB+ Trees & \begin{itemize}
-              \item Same features of B+ Trees
-             \end{itemize} &
-\begin{itemize}
- \item Improves cache miss performance by prefetching more data in each cache miss
-\item Node size is also increased leading to the same benefits of CSB+ Trees
-\end{itemize}\\
+pB+tree & 
+-Same features of B+ Trees
+             &
+-Improves cache miss performance by prefetching more data in each cache miss \linebreak \linebreak
+-Node size is also increased leading to the same benefits of CSB+ Trees\\
 \hline
-T Tree &  \begin{itemize}
-              \item Contain pointers to data values instead of the values themselves leading to better space utilization
-\item Many sorted keys per node
-\item Only minimum and maximum keys used for comparisons
-\item Contains record pointers in every key
-\item Good update \& storage characteristics if Btrees.
-
-             \end{itemize} &
-\begin{itemize}
- \item Cache behavior was not considered at time of design
-\item Record pointers waste space in cache
-\end{itemize}\\
+T-tree &  
+-Contains pointers to data values instead of the values themselves leading to better space utilization \linebreak \linebreak
+-Many sorted keys per node \linebreak \linebreak
+-Only minimum and maximum keys used for comparisons \linebreak \linebreak
+\raggedright -Contains record pointers in every key \linebreak \linebreak
+-Good update \& storage characteristics if Btrees.
+             &
+-Cache behavior was not considered at time of design \linebreak \linebreak
+-Record pointers waste space in cache\\
 \hline
-CST & \begin{itemize}
-       \item Binary search tree for maximum value of each node leads to faster search
-\item More space efficient by removing pointers \& using array storage. Index calculation used to allocate values
-      \end{itemize} &
-\begin{itemize}
- \item Node size is aligned with cache line size to avoid misses
-\item Higher usage of cached data in binary search tree
-\end{itemize}\\
+CST & 
+-Binary search tree for maximum value of each node leads to faster search \linebreak \linebreak
+-More space efficient by removing pointers \& using array storage. Index calculation used to allocate values &
+-Node size is aligned with cache line size to avoid misses \linebreak \linebreak
+-Higher usage of cached data in binary search tree\\
 \hline
-CSS Trees &  \begin{itemize}
-       \item Fast traversal in $log_{m+1}n$ ($m$ is the number of keys per node)
-\item Mainly suitable for read environments
-      \end{itemize} &
-\begin{itemize}
- \item Good cache behavior since $m$ is chosen to fit in the cache line size
-\end{itemize}\\
+CSS-trees & 
+-Fast traversal in $log_{m+1}n$ ($m$ is the number of keys per node) \linebreak \linebreak
+-Mainly suitable for read environments &
+-Good cache behavior since $m$ is chosen to fit in the cache line size\\
 
@@
 \subsubsection*{B Trees}
-Although B Trees, explained in the previous section, are originally designed for disk based database systems, they were found to also be useful in
-MMDB~\cite{lehman1986study}. In disk based systems, the B+ Tree is more used than the regular B Tree. However, in main memory databases, the B Tree would be
+Although B-trees, explained in the previous section, are originally designed for disk based database systems, they were also found to be useful in
+MMDB~\cite{lehman1986study}. In disk based systems, the B+ Tree is more used than the regular B Tree. However, in MMDBs, the B-tree would be
 more suitable from a storage perspective since there is no gain from keeping all the data in the leaves. In a main memory system, this
 would only waste space without improving performance. On the other hand, the B+ Tree uses multiple keys to search within a node. This means that
-if the node fits in a cache line, this cache load can satisfy more than more comparison leading to better cache utilization~\cite{rao1999cache}. Therefore, B+
-Trees in general have reasonable cache behavior, and variations of them have been adapted in MMDBs.
+if the node fits in a cache line, this cache load can satisfy more than more comparison leading to better cache utilization~\cite{rao1999cache}. Additionally,
+less data per node is stored in a B+tree which means there is better cache utilization. Therefore, B+trees in general have reasonable cache behavior, and
+variations of them have been adapted in MMDBs.
 
-However, the cache behavior of B+ Trees could still be improved. Therefore, Rao and Ross~\cite{rao2000making} propose the Cache Sensitive B+-Trees
-($CSB^{+}-Trees$)~\cite{rao2000making} as an index structure for main memory. To do so, they eliminated most of the child pointers by storing the child nodes
+However, the cache behavior of B+trees could still be improved. Therefore, Rao and Ross~\cite{rao2000making} propose the Cache Sensitive B+trees
+(CSB+trees)~\cite{rao2000making} as an index structure for main memory. To do so, they eliminated most of the child pointers by storing the child nodes
 in an array, and only keeping a pointed to the first child. Additionally, they used more keys in each
-node to improve locality and reduce tree height~\cite{luan2009prefetching}. Since the number of cache misses in search operations is proportional to the height
-of the tree, $CSB^{+}-trees$ have fewer cache misses, and thus better performance.
+node to improve locality and reduce tree height~\cite{luan2009prefetching}. This also aligns the nodes size with the cache line size. Since the number of cache
+misses in search operations is proportional to the height
+of the tree, CSB+trees have fewer cache misses, and thus better performance.
 
-Chen et al.~\cite{chen2001improving} propose using prefetching to improve the performance of B+ Trees. They call the new index structure prefetching B+ Tree
-(pB+ Tree). Since current computer systems can prefetch different data simultaneously, the authors take advantage of this. The idea here is to increase the node
-size of B+ Trees and employ prefectching at the same time. In a B+ Tree, every time we move down a level in the tree, a cache miss occurs. Accordingly,
+Chen et al.~\cite{chen2001improving} propose using prefetching to improve the performance of B+trees. They call the new index structure prefetching B+tree
+(pB+tree). Since current computer systems can prefetch different data simultaneously, the authors take advantage of this. The idea here is to increase the node
+size of B+ Trees and employ prefectching at the same time. In a B+tree, every time we move down a level in the tree, a cache miss occurs. Accordingly,
 prefetching larger sizes nodes means that we can overlap multiple cache misses before they occur. This is because in one cache miss, extra data is retrieved in
@@
 \subsubsection*{T Trees}
-Lehmen and Carey~\cite{lehman1986study} introduce a new index structure, the T Tree, which combines the features from AVL Trees and B Trees that are suited to
-main memory. Figure~\ref{fig:ttree} shows the proposed structure of the T Tree. They run some experiments on data residing in main memory to compare the
-performance of the T
-Tree to existing structures. All structures were modified to contain pointers to data values rather than the data values themselves. The structures
-compared included  AVL Trees, simple arrays, B Trees, Chained Bucket Hashing, Extendible Hashing, Linear Hashing and Modified Linear Hashing. Their
-experiments showed that for unordered data, Modified Linear Hashing gave the best performance, and for ordered data, T Trees gave the best performance for a mix
-of searches, inserts and deletes. This is mainly because a T node contains many elements which results in good update and storage characteristics. The T Tree
-was the first index structure specifically designed for main memory databases. The T Tree has many sorted keys per node, and only uses the minimum and maximum
+Lehmen and Carey~\cite{lehman1986study} introduce a new index structure, the T-tree, which combines the features from AVL Trees and B-trees that are suited to
+main memory. Figure~\ref{fig:ttree} shows the proposed structure of the T-tree. They ran some experiments on data residing in main memory to compare the
+performance of the T-tree to existing structures. All structures were modified to contain pointers to data values rather than the data values themselves. The
+structures compared included  AVL Trees, simple arrays, B-trees, Chained Bucket Hashing, Extendible Hashing, Linear Hashing and Modified Linear Hashing. Their
+experiments showed that for unordered data, Modified Linear Hashing gave the best performance, and for ordered data, T-trees gave the best performance for a mix
+of searches, inserts and deletes. This is mainly because a T node contains many elements which results in good update and storage characteristics. The T-tree
+was the first index structure specifically designed for main memory databases. The T-tree has many sorted keys per node, and only uses the minimum and maximum
 keys for comparison. Additionally, every key contains a record pointer.
@@
 
-Although T Trees have been specifically designed for MMDB, they did not consider the cache behavior at that time since the gap between processor speed and main
+Although T-trees have been specifically designed for MMDB, they did not consider the cache behavior at that time since the gap between processor speed and main
 memory speed was not big then~\cite{rao1999cache}. Although many keys are
 fit into one node, only the two end keys are used for comparison leading to low node utilization. Similarly, array binary search trees have the same problem. 
-It was discovered that B+-trees outperform
-T-Trees on modern processors because of the growth of cache miss latency relative to processor speed~\cite{rao1999cache,lee2007cst}. This is because the height
+It was discovered that B+trees outperform
+T-trees on modern processors because of the growth of cache miss latency relative to processor speed~\cite{rao1999cache,lee2007cst}. This is because the height
 of the tree is high which makes the total number of memory accesses from the root to the leaf node higher. The other reason is that the node size is not aligned
-with the cache line size. Another problem with T-Trees is that there is a big waste of space in the cache because unnecessary data which is not used (the
+with the cache line size. Another problem with T-trees is that there is a big waste of space in the cache because unnecessary data which is not used (the
 record pointers) brought into the cache.
 
-To resolve these issues, Lee et al.~\cite{lee2007cst} propose the Cache Sensitive T-Trees (CST-Trees). First, a binary
+To resolve these issues, Lee et al.~\cite{lee2007cst} propose the Cache Sensitive T-trees (CST-trees). First, a binary
 search tree whose node values are the maximum key of each node in the T Tree is constructed. That way, for a given value, the node containing this value could
@@
 Binary Search Trees have poor cache behavior when the array is much bigger than the cache. Rao and Ross~\cite{rao1999cache} propose the Cache Sensitive Search
-Tree (CSS-Tree) to improve this cache behavior. To do that, the search tree is stored as a directory in an array. The number of keys per node is chosen such
+Tree (CSS-tree) to improve this cache behavior. To do that, the search tree is stored as a directory in an array. The number of keys per node is chosen such
 that the whole node fits in a cache line. This improves local searching within a node to just one cache miss. They also hard code the traversal within a node
 such that finding the next node happens in $log_{m+1}n$ times rather than $log_{2}n$ times in a binary search tree where $m$ is the number of keys per node.
-However, CSS Trees are only suitable for read environments since they eliminate nearly all child pointers, thus removing the support for incremental update.
+However, CSS-trees are only suitable for read environments since they eliminate nearly all child pointers, thus removing the support for incremental update.
 
 \subsubsection{Materialized Views}
-From our literature survey, it seems that materialized views are not used in MMDB. This makes sense since the problems materialized views try to address are
-not present in MMDB. That is, the cost of computing complicated join queries is much less in MMDB due to the low cost of accessing the data in memory.
-Accordingly, the maintenance cost of materialized views in MMDB will outweigh their minimal benefit. However, this does not mean that they cannot be used at
-all in MMDBs. As with all other physical design decisions, it rather depend on the benefits of implementing the technique. Accordingly, if the data is very
-large and may benefit from using materialized views, there is no reason not to use them.
+From our literature survey, it seems that materialized views are not frequently used in MMDB. This makes sense since the problems materialized views try to
+address problems are not highly present in MMDB. That is, the cost of computing complicated join queries is much less in MMDB due to the low cost of accessing
+the data in memory. Accordingly, the maintenance cost of materialized views in MMDB will outweigh their minimal benefit. However, this does not mean that they
+cannot be used at all in MMDBs. As with all other physical design decisions, it rather depend on the query workload and the relative benefits of implementing
+the technique. Accordingly, if the data is very large and may benefit from using materialized views, there is no reason not to use them.
 
@@
 reload and the number of main memory references during transaction processing. Gruenwald and Eich~\cite{gruenwald1990database,gruenwald1990choosing} show that
-horizontal partitioning is the most suitable when the database performs more modifications than tuple deletions and more selections that projections and joins.
+horizontal partitioning is most suitable when the database performs more modifications than tuple deletions and more selections that projections and joins.
 Vertical partitioning is chosen in the other cases. They also show that if we just concentrate on the reload performance (i.e the number of I/Os for
@@
 Although data in a data warehouse is extracted and loaded from multiple on-line transaction processing (OLTP) data sources (including DB2, Oracle, IMS
-databases, and flat files) using Extract, Transfer, and Load (ETL) tools (see Figure \ref{fig:dw}), a data warehouse is usually maintained seperately from the
+databases, and flat files) using Extract, Transfer, and Load (ETL) tools (Figure \ref{fig:dw}), a data warehouse is usually maintained seperately from the
 organization's operational databases \cite{sen2005comparison, chaudhuri1997overview}. This architecture can be justified owing to the fact that operational
@@
 
+\begin{table}[!t]
+\centering
+\scalebox{0.7}{
+\begin{tabular}{p{2cm}||p{5cm}|p{4cm}|p{4cm}|p{4cm}}
+ & \textbf{Index Structures} & \textbf{Materialized Views} & \textbf{Partitioning} & \textbf{Clustering}\\
+\hline\hline
+\textbf{RDBMS}& E.g. B+trees, Bitmap Index, Hash Tree & Store results of common complicated queries on disk & Horizontal Partitioning (range, list, hash,
+composite) Vertical Partitioning& Group related items together on disk\\
+\hline
+\textbf{MMDB} & 
+-No need to store actual data values in index \linebreak \linebreak 
+-Larger node size that is aligned with cache line size \linebreak \linebreak
+-E.g. B+tree, CSB+tree, T tree, CST tree & Not highly utilized since processing and main memory access is cheap & Only needed in secondary storage to speed up
+reloading in case of a crash\linebreak \linebreak 
+-Horizontal \& Vertical partitioning& 
+Not needed since random access in main memory costs the same as sequential access\\
+\hline
+\textbf{XML DB} & Need a number scheme. E.g. B+tree and variations used. & XQuery and XPath results are materialized & Horizontal partitioning based on inlining
+or node
+type & Based on document order or structural similarity\\
+\hline
+\textbf{Data Warehouses} & E.g. Universal B+tree, Bitmap index, Projection Index & 
+-Essential to improve performance \linebreak \linebreak 
+-Not applicable to adhoc queries & 
+-Horizontal partitioning based on time dimension \linebreak \linebreak 
+-Vertical partitioning of seldom used columns. & Data is
+clustered by
+nature.\\
+\end{tabular}
+}
+\caption{Summary of the different elements of physical design in the studied systems}
+\label{tab:summary}
+\end{table}
+
 \section{Conclusions}
 \label{SEC-CONCL}
+
+In this survey, we present four elements of physical design in database: index structures, materialized views, partitioning, and clustering. We started out
+by explaining how these design factors are implemented in traditional Relational Database Management Systems (RDBMS), and then presented how they are used in
+other information systems. We focused on three types of information systems: Main Memory Database Systems (MMDB), XML Databases, and Data Warehouses. We
+presented the challenges involved with each type of system, and how each of the four physical design elements were adapted or utilized to overcome these
+challenges. Table~\ref{tab:summary} summarizes the results of our survey. From our literature review, we found that the B+tree is the most utilized index
+structure, and has been adapted and customized to fit the different needs of other systems. For example, it was made more cache conscious to fit MMDB needs,
+and was changed to apply to multidimensions to fit the Data Warehouses needs. Materialized views were found to be heavily utlitized in all the systems we
+presented, except in MMDBs. Partitioning was also used in all systems, but was only used for data reloading in MMDBs. Horizontal partitioning is much more
+common that vertical partitioning. Clustering is very important in all systems, except for MMDBs. Clustering is especially very important for Data Warehouses
+where huge amounts of data are stored.
+
+It was interesting to see how the different nature of these systems affects the physical design. Throughout the survey, we have showed some of the
+commonalities between the different systems. Moreover, understanding how to meet these needs would help us design mash-ups of these different systems. For
+example, just like relational database systems shifted into main memory, we might have an XML main memory database system. In that case, the index structures
+fitted for XML data indexing will have to be adapted for main memory by making sure they are aligned with the cache line size, for example. Similarly, if XML
+data warehouses are developed, more elaborate path indexing techniques and more clustering would be required.
 

### 2010-07-26 cs848-db/Paper.tex
after integrating everything

@@
 
-\subsubsection*{B Trees}
+\subsubsection*{B-trees}
 Although B-trees, explained in the previous section, are originally designed for disk based database systems, they were also found to be useful in
@@
 \subsubsection{Materialized Views}
-not very useful since most queries are ad-hoc \cite{armstrong1997data, theodoratos2000general}
+A data warehouse must be built with a solid foundation that is flexible enough and responsive to business changes over time \cite{armstrong1997data}. Meanwhile,
+a data warehouse can be seen as a set of materialized views defined over the remote data sources (Figure \ref{fig:dw}). That is why selection of materialized
+views is one of the most challenging problems that faces data warehouse architects \cite{theodoratos2000general}. Most of the solutions in the literature
+defined the problem of selecting views to materialize as \textquotedblleft selecting views for materialization in order to satisfy a number of design goals
+given a set of constraints \textquotedblright. The problem can be reduced to a cost minimization problem, whether it is the query evaluation cost
+\cite{harinarayan1996implementing}, total view maintenance cost \cite{kenneth1996materialized}, both \cite{yang1997algorithms}, or the operational cost
+\cite{elena1997materialized, gupta2005selection, theodoratos1999designing, theodoratos1997data, theodoratos1999designingdw, theodoratos2000incremental,
+yang1997algorithms}. Although materialized views greatly improves the performance of a data warehouse, it might not be as useful when the query workload is
+formed of ad-hoc complex queries.
 
 \subsubsection{Partitioning}
-\cite{thusoo2010data}
-\paragraph{Vertical Partitioning}
+Both horizontal and vertical partitioning are heavily used in data warehouse design. Partitioning a data warehouse has many advantages. Horizontal partitioning
+is usually used in data warehouses to make the job of the data warehouse administrators easier by facilitating the process of manages such large data
+sources from which a data warehouse gets its data. In other words, it breaks down the management of large databases to managing smaller sets (i.e. partitions)
+of data \cite{thusoo2010data}. More importantly, partitioning provides parallelism in various situations: loading data sources \cite{barclay1994loading},
+performing parallel joins on huge data \cite{datta2002parallel}, faster drill-down and roll-up operations \cite{chaudhuri1997overview}. Partitioning also
+facilitates compressing data that is not of interest to users of the data warehouse. Vertical paritioning is also used to keep seldom used columns away from
+the process of answering queries for better performance.
 
-\paragraph{Horizontal Partitioning}
+On the other hand, if partitioning is over-used in a data warehouse, managing them will be a complex process due to the fact of having a huge amount of data.
+Finally, there is an inherent cost of managing the meta data (e.g. data cube definition) when the number of partitions increases.
+
+\subsubsection{Clustering}
+A data warehouse is clustered by nature as Figure \ref{fig:dw} suggests. The data loaded into a data warehouse comes from various operational data sources,
+each source has the data that is the most interesting/related to it (i.e. clustering). The data warehouse can be further divided into a collection of
+conforming data marts \cite{sen2005comparison}. Each data mart stores data that is related to each other (e.g. based on the organization's business processes).
+For example, one data mart can store sales data that is of particular interest to the Sales Department in an organization. Along with sales data,
+product-related information can also be stored to provide better locality for queries that usually ask question about sales and products, and so on.
 
@@
 administrators. Accordingly, researchers have been inclined to find ways to automate these decisions. There were many early research efforts done in this
-area(E.g.~\cite{finkelstein1988physical, hammer76selec, frank1992selec, chaudhuri1997overview, schiefer1999db2, valentin2000select}), but most of these efforts
+area(E.g. \cite{finkelstein1988physical, hammer76selec, frank1992selec, chaudhuri1997overview, schiefer1999db2, valentin2000select}), but most of these efforts
 focused on having a static workload. Accordingly, research shifted to more dynamic ways to automate physical design. Initial attempts used an dynamic, on-line
-approaches(E.g.~\cite{bruno2007online, sattler2003, schnaitter2007}) where they observed the changes in the workload and adjusted the design accordingly while
+approaches(E.g. \cite{bruno2007online, sattler2003, schnaitter2007}) where they observed the changes in the workload and adjusted the design accordingly while
 attempting to predict future behavior based on the observed behavior. To avoid the problems of on-line design, researchers are also currently looking at how to
-perform this dynamic physical design off-line (E.g. ~\cite{hannes2008,agrawal2006}). Additionally, finding better evaluation techniques to compare different
-automatic physical design decision is currently being studied~\cite{gebaly2008}.
+perform this dynamic physical design off-line (E.g. \cite{hannes2008,agrawal2006}). Additionally, finding better evaluation techniques to compare different
+automatic physical design decision is currently being studied \cite{gebaly2008}.
 
 Other problems which are currently being researched include improving I/O operations in database systems by using Solid State Drives (SSD) instead of of Hard
-Disk Drives (HDD)~\cite{lee2009, du2009, lee2008}. SSDs have faster read performance, but perform poorly in random writes. However, for certain applications in
+Disk Drives (HDD) \cite{lee2009, du2009, lee2008}. SSDs have faster read performance, but perform poorly in random writes. However, for certain applications in
 databases and for databases where random writes are not as frequent, SSDs might be suitable. Using SDDs also seems promising when used for
-transaction logs, and intermediate results of sorting and joining~\cite{lee2008}.
+transaction logs, and intermediate results of sorting and joining \cite{lee2008}.
 
@@
 storage techniques and index structures to the cloud. For example, Wang et al. propose a new index structure for indexing multi-dimensional data in the
-cloud~\cite{wang2010indexing}. Lots of work is currently being done in this area. For example, Wu et al.~\cite{wu2010efficient} propose a modified B+tree index
+cloud \cite{wang2010indexing}. Lots of work is currently being done in this area. For example, Wu et al. \cite{wu2010efficient} propose a modified B+tree index
 for datasets stored in each compute node. The exact requirements and solutions for data management in the cloud are still being
-researched~\cite{abadi2009data}, and so this is an area that is expected to receive more attention in the future as cloud computing becomes more prevalent.
+researched \cite{abadi2009data}, and so this is an area that is expected to receive more attention in the future as cloud computing becomes more prevalent.
 
@@
 use of them or to be able to cope with the new requirements they pose.
+
+\section{Conclusions}
+\label{SEC-CONCL}
+
+In this survey, we present four elements of physical design in database: index structures, materialized views, partitioning, and clustering. We started out
+by explaining how these design factors are implemented in traditional Relational Database Management Systems (RDBMS), and then presented how they are used in
+other information systems. We focused on three types of information systems: Main Memory Database Systems (MMDB), XML Databases, and Data Warehouses. We
+presented the challenges involved with each type of system, and how each of the four physical design elements were adapted or utilized to overcome these
+challenges. Table \ref{tab:summary} summarizes the results of our survey. From our literature review, we found that the B+tree is the most utilized index
+structure, and has been adapted and customized to fit the different needs of other systems. For example, it was made more cache conscious to fit MMDB needs,
+and was changed to apply to multidimensions to fit the Data Warehouses needs. Materialized views were found to be heavily utlitized in all the systems we
+presented, except in MMDBs. Partitioning was also used in all systems, but was only used for data reloading in MMDBs. Horizontal partitioning is much more
+common that vertical partitioning. Clustering is very important in all systems, except for MMDBs. Clustering is especially very important for Data Warehouses
+where huge amounts of data are stored.
+
+It was interesting to see how the different nature of these systems affects the physical design. Throughout the survey, we have showed some of the
+commonalities between the different systems. Moreover, understanding how to meet these needs would help us design mash-ups of these different systems. For
+example, just like relational database systems shifted into main memory, we might have an XML main memory database system. In that case, the index structures
+fitted for XML data indexing will have to be adapted for main memory by making sure they are aligned with the cache line size, for example. Similarly, if XML
+data warehouses are developed, more elaborate path indexing techniques and more clustering would be required.
 
@@
 \hline
-\textbf{Data Warehouses} & E.g. Universal B+tree, Bitmap index, Projection Index & 
--Essential to improve performance \linebreak \linebreak 
--Not applicable to adhoc queries & 
+\textbf{Data Warehouses} & E.g. Universal B+tree, Bitmap index, Projection Index, R*-tree & 
+-Reduce query evaluation cost \linebreak
+-Reduce view maintenance cost \linebreak 
+-Reduce operational costs & 
 -Horizontal partitioning based on time dimension \linebreak \linebreak 
@@
 
-\section{Conclusions}
-\label{SEC-CONCL}
-
-In this survey, we present four elements of physical design in database: index structures, materialized views, partitioning, and clustering. We started out
-by explaining how these design factors are implemented in traditional Relational Database Management Systems (RDBMS), and then presented how they are used in
-other information systems. We focused on three types of information systems: Main Memory Database Systems (MMDB), XML Databases, and Data Warehouses. We
-presented the challenges involved with each type of system, and how each of the four physical design elements were adapted or utilized to overcome these
-challenges. Table~\ref{tab:summary} summarizes the results of our survey. From our literature review, we found that the B+tree is the most utilized index
-structure, and has been adapted and customized to fit the different needs of other systems. For example, it was made more cache conscious to fit MMDB needs,
-and was changed to apply to multidimensions to fit the Data Warehouses needs. Materialized views were found to be heavily utlitized in all the systems we
-presented, except in MMDBs. Partitioning was also used in all systems, but was only used for data reloading in MMDBs. Horizontal partitioning is much more
-common that vertical partitioning. Clustering is very important in all systems, except for MMDBs. Clustering is especially very important for Data Warehouses
-where huge amounts of data are stored.
-
-It was interesting to see how the different nature of these systems affects the physical design. Throughout the survey, we have showed some of the
-commonalities between the different systems. Moreover, understanding how to meet these needs would help us design mash-ups of these different systems. For
-example, just like relational database systems shifted into main memory, we might have an XML main memory database system. In that case, the index structures
-fitted for XML data indexing will have to be adapted for main memory by making sure they are aligned with the cache line size, for example. Similarly, if XML
-data warehouses are developed, more elaborate path indexing techniques and more clustering would be required.
-
 \bibliographystyle{IEEEtran}

### 2010-07-26 cs848-db/Paper.tex

@@
 the XML data are labeled with the ORDPATH~\cite{o2004ordpaths} mechanism. The XML data is then shredded into a relational table with five main columns: ORDPATH,
-TAG, NODE\_TYPE, VALUE, and PATH\_ID. This information is then stored in a $B^{+}-tree$ as the primary index of the XML type. Additionally, a secondary index
+TAG, NODE\_TYPE, VALUE, and PATH\_ID. This information is then stored in a B+tree as the primary index of the XML type. Additionally, a secondary index
 can be created on any of the columns of the primary index to speed things up.
@@
 identifiers to each node in the XML document. These unique number identifiers allow the determination of any node's parent, sibling or possible child nodes.
-Four index files are created for XML data where all the indexes are based on $B^{+}-trees$. One index manages the collection hierarchy, one index collects nodes
+Four index files are created for XML data where all the indexes are based on B+-trees. One index manages the collection hierarchy, one index collects nodes
 in a paged file and associates unique node identifiers to the actual nodes, one index indexes elements and attributes, and the last index keeps track of word
@@
 The main idea behind using materialized views is to rewrite queries using these views so that they run more efficiently. This is very important in XML
-Databases since XQuery and XPath queries are complicated~\cite{arion2007structured}. Accordingly, materialized views are used in XML databases to improve
+databases since XQuery and XPath queries are complicated~\cite{arion2007structured}. Accordingly, materialized views are used in XML databases to improve
 performance. The semi-structured nature of
@@
 Clustering in XML databases follows the heuristic that sub-elements are likely to be queries with an element~\cite{jagadish2002timber}. Accordingly, elements
-and their subelements are clustered together. In general, storing the XML data in document order it thought to be the best technique for efficient querying.
+and their subelements are clustered together. In general, storing the XML data in document order is thought to be the best technique for efficient querying.
 Additionally, Lian et al.~\cite{lian2004efficient} suggest clustering XML documents by structural similarity to produce more efficient access. This is done by
@@
 executives named Data warehousing as the highest-priority post-millennium project for them \cite{sen2005comparison}. The value of data warehousing for an
-organization depends on the organization's need for reliable, consolidated, unique and integrated reporting and analysis of its data, at different levels of
+organization depends on the organization's need for reliable, consolidated, unique and integrated reporting and analysis of its data at different levels of
 aggregation.
@@
 
-Data warehouses tend to be extremely large, in fact it is quite possible for a data warehouse to store tens of petabytes of data while loading tens of terabytes
-of data everyday \cite{thusoo2010data}. Datta et al. \cite {datta2002parallel} notes that the information in a data warehouse is usually multidimensional in
+Data warehouses tend to be extremely large. In fact, it is quite possible for a data warehouse to store tens of petabytes of data while loading tens of
+terabytes
+of data everyday \cite{thusoo2010data}. Datta et al. \cite {datta2002parallel} note that the information in a data warehouse is usually multidimensional in
 nature, requiring the capability to view the data from a variety of perspectives. Aggregated and summarized data become more crucial than detailed records in
-such environment. Therefore, the workloads are query intensive with mostly ad hoc, complex queries, often requiring computationally expensive operations such as
+such environment. Therefore, the workloads are query intensive with mostly ad hoc, complex queries often requiring computationally expensive operations such as
 scans, joins, and aggregation. Performing such operations on large amounts of data, like in the case of data warehousing, complicates the situation further.
@@
 Although data in a data warehouse is extracted and loaded from multiple on-line transaction processing (OLTP) data sources (including DB2, Oracle, IMS
-databases, and flat files) using Extract, Transfer, and Load (ETL) tools (Figure \ref{fig:dw}), a data warehouse is usually maintained seperately from the
+databases, and flat files) using Extract, Transfer, and Load (ETL) tools (Figure \ref{fig:dw}), a data warehouse is usually maintained separately from the
 organization's operational databases \cite{sen2005comparison, chaudhuri1997overview}. This architecture can be justified owing to the fact that operational
@@
 quality, or use inconsistent representations, codes and formats. Moreover, understanding trends or making predictions requires historical data, whereas
-operational databases store only current data. In OLAP, there's a need to support multidimensional data models and operations which requires special data
-organization, access methods, and implementation methods, not generally provided by DBMSs targeted for OLTP \cite{chaudhuri1997overview}. Finally, in data
+operational databases store only current data. In OLAP, there is a need to support multidimensional data models and operations which requires special data
+organization, access methods, and implementation methods not generally provided by DBMSs targeted for OLTP \cite{chaudhuri1997overview}. Finally, in data
 warehousing, query throughput and response times are more important than transaction throughput which is the major performance metric for operational databases.
@@
 The design of a data warehouse can follow one of two paradigms: relational-based OLAP (ROLAP) or multidimensional OLAP (MOLAP). In ROLAP, data is stored in
-relational databases and the support for multidimensional data is achieved through the use of the star schema \cite{cheung2001towards}. A star schema (Figure
-\ref{fig:starschema}) is a database schema where a fact table $T$ that holds the base data in the database. Other tables in the database are dimensions for
-table $T$, where each dimension table is linked to the fact table through foreign keys in $T$ pointing to the primary keys in the dimension tables. Thus, the
+relational databases and the support for multidimensional data is achieved through the use of the star schema \cite{cheung2001towards}. A star schema,
+shown in Figurepearson
+\ref{fig:starschema}, is a database schema where a fact table $T$ holds the base data in the database. Other tables in the database are dimensions for
+table $T$ where each dimension table is linked to the fact table through foreign keys in $T$ pointing to the primary keys in the dimension tables. Thus, the
 fact table, which is always the largest table in such schema, participates in almost all join operations \cite{datta2002parallel}. This might enforce some
@@
 \paragraph{Tree-based Indices}
-The widely used B-trees have been modified to suit the requirements of a data warehouse design, such modification resulted in various tree-based indices:
+The widely used B-trees have been modified to suit the requirements of a data warehouse design. Such modifications resulted in various tree-based indices:
 UB-trees, BV-trees, K-D-B-trees, and R-trees. A UB-tree index \cite{bayer1997universal} is a variation of B+trees where support for multidimensional data is
@@
 type of linearization. UB-trees inherit the advantages of B+trees, so they are well suited for high cardinality attributes. They are also excellent for point
-and interval queries (using the pointer in each leaf node to the next leaf node for faster index traversal). UB-trees also inherit the disadvantages of
+and interval queries (using the pointer in each leaf node to the next leaf node for faster index traversal). However, UB-trees also inherit the disadvantages of
 B+trees. Therefore, they poorly perform for low cardinality attributes.
@@
 \paragraph{Projection Index}
-Not all columns in a data warehouse are of interest to the given query workloads. Therefore, it makes sense to have faster access to the columns that are used
-the most to improve the query performance. That is why Projection indices are often used in data warehousing. A projection index involves positional indexing
+Not all columns in a data warehouse are of interest to the given query workloads. Therefore, it makes sense to have faster access to the columns that are
+most used to improve the query performance. That is why Projection indices are often used in data warehousing. A projection index involves positional indexing
 of a column (form of vertical partitioning) where tuples are accessed based on their ordinal position \cite{datta2002parallel}. More results per I/O operation
@@
 \subsubsection{Materialized Views}
-A data warehouse must be built with a solid foundation that is flexible enough and responsive to business changes over time \cite{armstrong1997data}. Meanwhile,
-a data warehouse can be seen as a set of materialized views defined over the remote data sources (Figure \ref{fig:dw}). That is why selection of materialized
-views is one of the most challenging problems that faces data warehouse architects \cite{theodoratos2000general}. Most of the solutions in the literature
+A data warehouse must be built with a solid foundation that is flexible enough and responsive to business changes over time \cite{armstrong1997data}.
+Meanwhile, as shown in Figure~\ref{fig:dw}, a data warehouse can be seen as a set of materialized views defined over the remote data sources. That is why
+selection of materialized
+views is one of the most challenging problems that data warehouse architects face \cite{theodoratos2000general}. Most of the solutions in the literature
 defined the problem of selecting views to materialize as \textquotedblleft selecting views for materialization in order to satisfy a number of design goals
@@
 Both horizontal and vertical partitioning are heavily used in data warehouse design. Partitioning a data warehouse has many advantages. Horizontal partitioning
-is usually used in data warehouses to make the job of the data warehouse administrators easier by facilitating the process of manages such large data
+is usually used in data warehouses to make the job of the data warehouse administrators easier by facilitating the process of managing such large data
 sources from which a data warehouse gets its data. In other words, it breaks down the management of large databases to managing smaller sets (i.e. partitions)
@@
 performing parallel joins on huge data \cite{datta2002parallel}, faster drill-down and roll-up operations \cite{chaudhuri1997overview}. Partitioning also
-facilitates compressing data that is not of interest to users of the data warehouse. Vertical paritioning is also used to keep seldom used columns away from
+facilitates compressing data that is not of interest to users of the data warehouse. Vertical partitioning is also used to keep seldom used columns away from
 the process of answering queries for better performance.
 
-On the other hand, if partitioning is over-used in a data warehouse, managing them will be a complex process due to the fact of having a huge amount of data.
+On the other hand, if partitioning is over-used in a data warehouse, managing it will be a complex process due to the fact of having a huge amount of
+data.
 Finally, there is an inherent cost of managing the meta data (e.g. data cube definition) when the number of partitions increases.
@@
 \subsubsection{Clustering}
-A data warehouse is clustered by nature as Figure \ref{fig:dw} suggests. The data loaded into a data warehouse comes from various operational data sources,
+A data warehouse is clustered by nature as Figure \ref{fig:dw} suggests. The data loaded into a data warehouse comes from various operational data sources where
 each source has the data that is the most interesting/related to it (i.e. clustering). The data warehouse can be further divided into a collection of
@@
 administrators. Accordingly, researchers have been inclined to find ways to automate these decisions. There were many early research efforts done in this
-area(E.g. \cite{finkelstein1988physical, hammer76selec, frank1992selec, chaudhuri1997overview, schiefer1999db2, valentin2000select}), but most of these efforts
-focused on having a static workload. Accordingly, research shifted to more dynamic ways to automate physical design. Initial attempts used an dynamic, on-line
+area(E.g. \cite{finkelstein1988physical, chaudhuri1997overview, hammer76selec, frank1992selec, schiefer1999db2, valentin2000select}), but most of these efforts
+focused on having a static workload. Accordingly, research shifted to more dynamic ways to automate physical design. Initial attempts used dynamic, on-line
 approaches(E.g. \cite{bruno2007online, sattler2003, schnaitter2007}) where they observed the changes in the workload and adjusted the design accordingly while
@@
 
-In this survey, we present four elements of physical design in database: index structures, materialized views, partitioning, and clustering. We started out
+In this survey, we presented four elements of physical design in database: index structures, materialized views, partitioning, and clustering. We started out
 by explaining how these design factors are implemented in traditional Relational Database Management Systems (RDBMS), and then presented how they are used in
@@
 \hline\hline
-\textbf{RDBMS}& E.g. B+trees, Bitmap Index, Hash Tree & Store results of common complicated queries on disk & Horizontal Partitioning (range, list, hash,
-composite) Vertical Partitioning& Group related items together on disk\\
+\textbf{RDBMS}& E.g. B+trees, Bitmap Index, Hash Tree & Store results of common complicated queries on disk & 
+\raggedright  -Horizontal Partitioning (range, list, hash,
+composite) \linebreak \linebreak
+-Vertical Partitioning& Group related items together on disk\\
 \hline
 \textbf{MMDB} & 
--No need to store actual data values in index \linebreak \linebreak 
--Larger node size that is aligned with cache line size \linebreak \linebreak
--E.g. B+tree, CSB+tree, T tree, CST tree & Not highly utilized since processing and main memory access is cheap & Only needed in secondary storage to speed up
+\raggedright  -No need to store actual data values in index \linebreak \linebreak 
+\raggedright  -Larger node size that is aligned with cache line size \linebreak \linebreak
+-E.g. B+tree, CSB+tree, T tree, CST tree & Not highly utilized since processing and main memory access is cheap & 
+\raggedright  -Only needed in secondary storage to speed up
 reloading in case of a crash\linebreak \linebreak 
@@
 \hline
-\textbf{XML DB} & Need a number scheme. E.g. B+tree and variations used. & XQuery and XPath results are materialized & Horizontal partitioning based on inlining
-or node
-type & Based on document order or structural similarity\\
+\textbf{XML DB} &
+\raggedright -Need a numbering scheme for nodes. \linebreak \linebreak
+-E.g. B+tree and variations used. & XQuery and XPath results are materialized & Horizontal partitioning
+based on inlining or node type & Based on document order or structural similarity\\
 \hline
 \textbf{Data Warehouses} & E.g. Universal B+tree, Bitmap index, Projection Index, R*-tree & 
--Reduce query evaluation cost \linebreak
--Reduce view maintenance cost \linebreak 
--Reduce operational costs & 
--Horizontal partitioning based on time dimension \linebreak \linebreak 
--Vertical partitioning of seldom used columns. & Data is
+\raggedright-Reduce query evaluation cost \linebreak \linebreak
+\raggedright -Reduce view maintenance cost \linebreak \linebreak
+\raggedright -Reduce operational costs & 
+\raggedright  -Horizontal partitioning based on time dimension \linebreak \linebreak 
+\raggedright -Vertical partitioning of seldom used columns. & Data is
 clustered by

### 2010-07-27 cs848-db/Paper.tex

@@
 Relational Database Management Systems (RDBMS) have been the most popular type of Information Systems for decades. When we think of designing a database system,
-we usually think of which data we need to keep track of and how will we properly organize it into meaningful tables. However, this is not the only phase in
+we usually think of which data we need to keep track of and how we will properly organize it into meaningful tables. However, this is not the only phase in
 database design. All of the designed tables need to be physically stored on disk. How these tables will be stored and accessed greatly affects the database
@@
 Data is stored in the form of records in files which reside on a physical disk. In this sense, we try to find the most efficient way possible to store and
-access the data such that queries are answered and executed efficiently. In the traditional relational database system, this involved selecting proper indexing
+access the data such that queries are answered and executed efficiently. In the traditional relational database system, this involves selecting proper indexing
 techniques, partitioning the data for more efficient access, materializing the results of some queries, clustering and compressing the data, and many other
@@
 Databases. The aim of this survey
-is not to explain the details of every data structure used in physical design, but rather to compare how the different data structures and techniques have been
+is not to explain the details of every data structure used in physical design, but rather to compare how different data structures and techniques have been
 used in these
@@
 Different variations of B-trees have been used. For example, B+trees are now the main method of indexing in current disk-based Relational Databases
-\cite{lightstone2007physical} such as DB2, Oracle, and SQL Server. The main difference between the B+tree and the B-tree is that only the leaf nodes contain
+\cite{lightstone2007physical} such as DB2, Oracle, and SQL Server. The main difference between the B+tree and the B-tree is that only leaf nodes contain
 data pointers in a B+tree, and leaf nodes contain pointers to each other as shown in Figure \ref{fig:b+tree}. In Prefix B+trees, instead of using the actual key
@@
 table column $C$ where a record $x$ can have one of six unique values, a traditional bitmap index defined for $C$ generates one bitmap vector for each of the
-six different values ($B0-B5$). If the value of $C$ for record $x$ is set to 3, the corresponding bit in vector $B3$ is set to 1. Otherwise the bit is set to 0.
+six different values ($B0-B5$). If the value of $C$ for record $x$ is set to value 3, the corresponding bit in vector $B3$ is set to 1. Otherwise the bit is set
+to 0.
 
@@
 benefits
-gained from using bitmap indices, they poorly perform during updates. The reason is that when an update is to be executed to a bitmap index, the whole bit
+gained from using bitmap indices, they poorly perform during updates. The reason is that when an update is to be executed on a bitmap index, the whole bit
 vector
@@
 
-In general, indices are a way of organizing the logical structure of a database without altering its physical structure. However, sometimes it is desirable to
-store related records near each other on disk for better locality. This can be achieved through implementing a clustered index on the desired key values.
+Sometimes it is desirable to store related records near each other on disk for better locality. This can be achieved through implementing a clustered index on
+the desired key values.
 Otherwise, the created index is known as a non-clustered index. The primary advantage of a clustered index is therefore the ordering of the physical data rows
-in accordance with the index blocks that point to them. Therefore, fewer data block reads are required to answer a given query. The only caveat with using a
+in accordance with the index blocks that point to them. Therefore, fewer data block reads are required to answer a given query. The only caveat of using a
 clustered index is that there can only be one clustered index per table because the physical organization of the table must be fixed
@@
 
-Partitioning in physical database design aims to reduct the workload on any one hardware component (E.g. an individual disk). This is achieved by partitioning
+Partitioning in physical database design aims to reduce the workload on any one hardware component (E.g. an individual disk). This is achieved by partitioning
 (dividing) the data over several disks \cite{lightstone2007physical}. There are two main categories of partitioning: horizontal partitioning and vertical
@@
 On the other hand, defining clustering keys, clustering scheme, and the granularity of clustering is very difficult. Accordingly, the problem of
-designing an optimal clustering scheme has combinatorial complexity, having an extremely large search space \cite{jagadish1999snakes}.
+designing an optimal clustering scheme has combinatorial complexity leading to an extremely large search space \cite{jagadish1999snakes}.
 
@@
 designing index structures for MMDB. The first is that the data resides in main memory and not on disk, and so many of the
-considerations taken for I/O operations are no longer there. Instead of focusing on disk access and disk storage, the data structures for main memory
-databases should focus on the efficient use of CPU cycles and memory~\cite{lehman1986study}. The second is having index structures cache conscious. Cache
+considerations taken for I/O operations are no longer applicable. Instead of focusing on disk access and disk storage, the data structures for main memory
+databases should focus on the efficient use of CPU cycles and memory~\cite{lehman1986study}. The second is having cache conscious index structures. Cache
 memories improve performance by holding recently referenced data~\cite{smith1982cache}. If the memory reference is found in the cache, then execution proceeds
@@
 summary, a main memory index should be able to reduce overall computation time while using as little memory as possible~\cite{lehman1986study}. It should also
-be noted that unlike disk based systems where it is advantageous to store actual attribute values in the index, main memory systems do not require that.
+be noted that unlike disk-based systems where it is advantageous to store actual attribute values in the index, main memory systems do not require that.
 Instead, pointers to the actual attribute values can be used. This, in turn, will reduce the size of the index which is preferable in this case.
+
+\begin{figure}[!t]
+\centering
+\includegraphics[width=7cm]{figs/prefetching.png}
+\caption{Performance of various B+tree searches where a cache miss to memory takes 150 cycles and a subsequent access can begin 10 cycles
+later~\cite{chen2001improving}}
+\label{fig:prefetching}
+\end{figure}
 
@@
 survey. Table~\ref{tab:mmdbindexsumm} summarizes the features of the different index structures discussed in this section with respect to MMDBs. The T-tree is
-the most used index for MMDBs (E.g. MySQL CLuster). However, there are other indexes used as well. For example, IBM's solidDB uses Trie (or prefix tree) for its
+the most used index for MMDBs (E.g. MySQL CLuster). However, there are other indices used as well. For example, IBM's solidDB uses Trie (or prefix tree) for its
 indexing. For their purposes, it served as a good index for main memory since it eliminates the need for many key comparisons~\cite{ibmsoliddb}.
@@
 \scalebox{0.8}{
-\begin{tabular}{p{2.2cm}|p{6cm}|p{6cm}}
-\hline
+\begin{tabular}{p{2.2cm}||p{6cm}|p{6cm}}
  \centering \textbf{Index Structure} & \textbf{Features} & \textbf{Cache Consciousness}\\
@@
 -Improves cache miss performance by prefetching more data in each cache miss \linebreak \linebreak
--Node size is also increased leading to the same benefits of CSB+ Trees\\
+-Node size is also increased leading to the same benefits of CSB+trees\\
 \hline
@@
 \raggedright -Contains record pointers in every key \linebreak \linebreak
--Good update \& storage characteristics if Btrees.
+-Good update \& storage characteristics if B-trees
              &
@@
 \hline
-CST & 
+CST-tree & 
 -Binary search tree for maximum value of each node leads to faster search \linebreak \linebreak
@@
 \hline
-CSS-trees & 
+CSS-tree & 
 -Fast traversal in $log_{m+1}n$ ($m$ is the number of keys per node) \linebreak \linebreak
@@
 Although B-trees, explained in the previous section, are originally designed for disk based database systems, they were also found to be useful in
-MMDB~\cite{lehman1986study}. In disk based systems, the B+ Tree is more used than the regular B Tree. However, in MMDBs, the B-tree would be
+MMDB~\cite{lehman1986study}. In disk based systems, the B+tree is more used than the regular B-tree. However, in MMDBs, the B-tree would be
 more suitable from a storage perspective since there is no gain from keeping all the data in the leaves. In a main memory system, this
-would only waste space without improving performance. On the other hand, the B+ Tree uses multiple keys to search within a node. This means that
+would only waste space without improving performance. On the other hand, the B+tree uses multiple keys to search within a node. This means that
 if the node fits in a cache line, this cache load can satisfy more than more comparison leading to better cache utilization~\cite{rao1999cache}. Additionally,
@@
 (pB+tree). Since current computer systems can prefetch different data simultaneously, the authors take advantage of this. The idea here is to increase the node
-size of B+ Trees and employ prefectching at the same time. In a B+tree, every time we move down a level in the tree, a cache miss occurs. Accordingly,
+size of B+trees and employ prefetching at the same time. In a B+tree, every time we move down a level in the tree, a cache miss occurs. Accordingly,
 prefetching larger sizes nodes means that we can overlap multiple cache misses before they occur. This is because in one cache miss, extra data is retrieved in
@@
 
-\begin{figure}[!t]
-\centering
-\includegraphics[width=7cm]{figs/prefetching.png}
-\caption{Performance of various B+ Tree searches where a cache miss to memory takes 150 cycles and a subsequent access can begin 10 cycles
-later~\cite{chen2001improving}}
-\label{fig:prefetching}
-\end{figure}
-
-\subsubsection*{T Trees}
+\subsubsection*{T-trees}
 Lehmen and Carey~\cite{lehman1986study} introduce a new index structure, the T-tree, which combines the features from AVL Trees and B-trees that are suited to
@@
 structures compared included  AVL Trees, simple arrays, B-trees, Chained Bucket Hashing, Extendible Hashing, Linear Hashing and Modified Linear Hashing. Their
-experiments showed that for unordered data, Modified Linear Hashing gave the best performance, and for ordered data, T-trees gave the best performance for a mix
-of searches, inserts and deletes. This is mainly because a T node contains many elements which results in good update and storage characteristics. The T-tree
+experiments showed that for unordered data, Modified Linear Hashing gives the best performance, and for ordered data, T-trees give the best performance for a
+mix
+of searches, inserts and deletes. This is mainly because a T-tree node contains many elements which results in good update and storage characteristics. The
+T-tree
 was the first index structure specifically designed for main memory databases. The T-tree has many sorted keys per node, and only uses the minimum and maximum
@@
 such that finding the next node happens in $log_{m+1}n$ times rather than $log_{2}n$ times in a binary search tree where $m$ is the number of keys per node.
-However, CSS-trees are only suitable for read environments since they eliminate nearly all child pointers, thus removing the support for incremental update.
+However, CSS-trees are only suitable for read environments since they eliminate nearly all child pointers, thus removing the support for incremental updates.
 
@@
 From our literature survey, it seems that materialized views are not frequently used in MMDB. This makes sense since the problems materialized views try to
-address problems are not highly present in MMDB. That is, the cost of computing complicated join queries is much less in MMDB due to the low cost of accessing
+address are not highly present in MMDB. That is, the cost of computing complicated join queries is much less in MMDB due to the low cost of accessing
 the data in memory. Accordingly, the maintenance cost of materialized views in MMDB will outweigh their minimal benefit. However, this does not mean that they
@@
 
-Partitioning is used in disk based systems to divide the data into the physical pages in a way that will yield the best performance. This could be through
-vertical partitioning or horizontal partitioning, as previously explained. In main memory databases, we do not have the issue of expensive disk access which
+Partitioning is used in disk-based systems to divide the data into the physical pages in a way that will yield the best performance. This could be through
+vertical partitioning or horizontal partitioning, as previously explained. MMDBs do not have the issue of expensive disk access which
 makes partitioning in memory unnecessary as it will not save any costs. However, partitioning is needed in the secondary storage of the data on
@@
 reload and the number of main memory references during transaction processing. Gruenwald and Eich~\cite{gruenwald1990database,gruenwald1990choosing} show that
-horizontal partitioning is most suitable when the database performs more modifications than tuple deletions and more selections that projections and joins.
-Vertical partitioning is chosen in the other cases. They also show that if we just concentrate on the reload performance (i.e the number of I/Os for
+horizontal partitioning is most suitable when the database performs more modifications than tuple deletions and more selections than projections and joins.
+Vertical partitioning is suitable for the other cases. They also show that if we just concentrate on the reload performance (i.e the number of I/Os for
 reload), then vertical partitioning is always the best choice.
@@
 
-XML (eXtensible Markup Language) is increasingly being used as a data exchange format. Accordingly, the ability to store and manage XML documents is needed.
+XML (eXtensible Markup Language) is being increasingly used as a data exchange format. Accordingly, the ability to store and manage XML documents is
+needed.
 Salminen and Tompa~\cite{salminen2001requirements} define an XML database as ``a collection of XML documents and their parts, maintained by a system
@@
 
-In this section, we will discuss how the database's physical design needs to be adapted in both cases. Other challenges such as expressing and executing queries
+In this section, we will discuss how the physical design elements need to be adapted in both cases. Other challenges such as expressing and executing queries
 and updates are faced in XML database design, but these are beyond the scope of this survey.
@@
 identifiers to each node in the XML document. These unique number identifiers allow the determination of any node's parent, sibling or possible child nodes.
-Four index files are created for XML data where all the indexes are based on B+-trees. One index manages the collection hierarchy, one index collects nodes
+Four index files are created for XML data where all the indexes are based on B+trees. One index manages the collection hierarchy, one index collects nodes
 in a paged file and associates unique node identifiers to the actual nodes, one index indexes elements and attributes, and the last index keeps track of word
@@
 
-TIMBER also uses a similar numbering schema. There are minor differences between the two numbering techniques, but the main idea is the same. The numbering is
+TIMBER~\cite{jagadish2002timber} also uses a similar numbering schema. There are minor differences between the two numbering techniques, but the main idea is
+the same. The numbering is
 based on three values of a node: its start label, its end label, and its level (i.e. nested depth). On the other hand, Sedna~\cite{taranov2010sedna} uses a
@@
 performance. The semi-structured nature of
-XML data, and the expressiveness of most XML querying languages, view selection for optimization becomes a more complicated
+XML data, and the expressiveness of most XML querying languages makes view selection for optimization a more complicated
 problem~\cite{tang2009materialized}. Therefore, the choice of which views to materialize becomes more complicated in XML database. Much work has been
@@
 representing the documents as a structure graph (s-graph), and then identifying clusters based on similarity using a clustering algorithm.
-Nayak~\cite{nayak2008fast} also proposes a similar clustering mechanism for XML data. Her technique is also based on structural similarity, but it is more
+Nayak~\cite{nayak2008fast} also proposes a similar clustering mechanism for XML data. Nayak's technique is also based on structural similarity, but it is more
 focused on the level structure of the XML documents.
@@
 knowledge worker (executive, manager, analyst) to make better and faster decisions \cite{inmon2002building, chaudhuri1997overview}. More than half of IT
-executives named Data warehousing as the highest-priority post-millennium project for them \cite{sen2005comparison}. The value of data warehousing for an
+executives named data warehousing as the highest-priority post-millennium project for them \cite{sen2005comparison}. The value of data warehousing for an
 organization depends on the organization's need for reliable, consolidated, unique and integrated reporting and analysis of its data at different levels of
@@
 nature, requiring the capability to view the data from a variety of perspectives. Aggregated and summarized data become more crucial than detailed records in
-such environment. Therefore, the workloads are query intensive with mostly ad hoc, complex queries often requiring computationally expensive operations such as
+such environment. Therefore, the workloads are query intensive with mostly ad-hoc, complex queries often requiring computationally expensive operations such as
 scans, joins, and aggregation. Performing such operations on large amounts of data, like in the case of data warehousing, complicates the situation further.
@@
 
-Although data in a data warehouse is extracted and loaded from multiple on-line transaction processing (OLTP) data sources (including DB2, Oracle, IMS
-databases, and flat files) using Extract, Transfer, and Load (ETL) tools (Figure \ref{fig:dw}), a data warehouse is usually maintained separately from the
+Data in a data warehouse is extracted and loaded from multiple on-line transaction processing (OLTP) data sources (including DB2, Oracle, IMS
+databases, and flat files) using Extract, Transfer, and Load (ETL) tools as shown in Figure \ref{fig:dw}. However, a data warehouse is usually maintained
+separately from the
 organization's operational databases \cite{sen2005comparison, chaudhuri1997overview}. This architecture can be justified owing to the fact that operational
@@
 relational databases and the support for multidimensional data is achieved through the use of the star schema \cite{cheung2001towards}. A star schema,
-shown in Figurepearson
-\ref{fig:starschema}, is a database schema where a fact table $T$ holds the base data in the database. Other tables in the database are dimensions for
+shown in Figure~\ref{fig:starschema}, is a database schema where a fact table $T$ holds the base data in the database. Other tables in the database are
+dimensions for
 table $T$ where each dimension table is linked to the fact table through foreign keys in $T$ pointing to the primary keys in the dimension tables. Thus, the
@@
 base data. OLAP operations are implemented on top of these special data structures. Usually a snowflake schema \cite{kimball2009data} or fact constellations are
-used to logically design the data warehouse.
+used to logically design the data warehouse in that case.
 
 The rest of this section will discuss how such challenges and requirements of data warehouse design affect the selection of physical design elements during the
-design process.
+data warehouse design process.
 
@@
 \paragraph{Bitmap Join Index}
-Bitmap join indices are another widely used type of indices in data warehouses epecially when a star schema is used. A bitmap join index is simply a bitmap
+Bitmap join indices are another widely used type of indices in data warehouses especially when a star schema is used. A bitmap join index is simply a bitmap
 index that is used for calculating the product of the join of the dimension tables. The result is then joined with the fact table to answer the given query
@@
 Not all columns in a data warehouse are of interest to the given query workloads. Therefore, it makes sense to have faster access to the columns that are
-most used to improve the query performance. That is why Projection indices are often used in data warehousing. A projection index involves positional indexing
+most used to improve the query performance. That is why projection indices are often used in data warehousing. A projection index involves positional indexing
 of a column (form of vertical partitioning) where tuples are accessed based on their ordinal position \cite{datta2002parallel}. More results per I/O operation
 is a direct benefit of using projection indices, since only data of interest are accessed and retrieved. However, only raw data can be fetched (e.g. column
-list in a selection clause). A projection index can be further extended into a bit-sliced index \cite{oneil1997improved} where the projection index is sliced
+list in a selection clause). A projection index can be further extended to a bit-sliced index \cite{oneil1997improved} where the projection index is sliced
 vertically into columns of bits. Each column is a bit-sliced index. Bit-sliced indices can help reduce storage costs by using a number of bits per value which
@@
 A data warehouse must be built with a solid foundation that is flexible enough and responsive to business changes over time \cite{armstrong1997data}.
-Meanwhile, as shown in Figure~\ref{fig:dw}, a data warehouse can be seen as a set of materialized views defined over the remote data sources. That is why
+Meanwhile, a data warehouse can be seen as a set of materialized views defined over the remote data sources. That is why
 selection of materialized
@@
 \subsubsection{Partitioning}
-Both horizontal and vertical partitioning are heavily used in data warehouse design. Partitioning a data warehouse has many advantages. Horizontal partitioning
+Both horizontal and vertical partitioning are heavily used in data warehouse design. Partitioning a data warehouse has many advantages. In specific, horizontal
+partitioning
 is usually used in data warehouses to make the job of the data warehouse administrators easier by facilitating the process of managing such large data
@@
 performing parallel joins on huge data \cite{datta2002parallel}, faster drill-down and roll-up operations \cite{chaudhuri1997overview}. Partitioning also
-facilitates compressing data that is not of interest to users of the data warehouse. Vertical partitioning is also used to keep seldom used columns away from
+facilitates compressing data that is not of interest to users of the data warehouse. Vertical partitioning helps keeping seldom used columns away from
 the process of answering queries for better performance.
@@
 Automating the physical database design choices has been a very attractive area of research that is still receiving lots of attention until now. Choosing which
-indexes to create, which views to materialize, and how to partition data are hard choices. They are highly dependent on the schema used as well
+indices to create, which views to materialize, and how to partition data are hard choices. They are highly dependent on the schema used as well
 as the query workload. The number of combinations of choices that can be made is very large making this task quite tedious and difficult for database
 administrators. Accordingly, researchers have been inclined to find ways to automate these decisions. There were many early research efforts done in this
-area(E.g. \cite{finkelstein1988physical, chaudhuri1997overview, hammer76selec, frank1992selec, schiefer1999db2, valentin2000select}), but most of these efforts
+area~\cite{finkelstein1988physical, chaudhuri1997overview, hammer76selec, frank1992selec, schiefer1999db2, valentin2000select}, but most of these efforts
 focused on having a static workload. Accordingly, research shifted to more dynamic ways to automate physical design. Initial attempts used dynamic, on-line
-approaches(E.g. \cite{bruno2007online, sattler2003, schnaitter2007}) where they observed the changes in the workload and adjusted the design accordingly while
+approaches~\cite{bruno2007online, sattler2003, schnaitter2007} where they observed the changes in the workload and adjusted the design accordingly while
 attempting to predict future behavior based on the observed behavior. To avoid the problems of on-line design, researchers are also currently looking at how to
-perform this dynamic physical design off-line (E.g. \cite{hannes2008,agrawal2006}). Additionally, finding better evaluation techniques to compare different
+perform this dynamic physical design off-line~\cite{hannes2008,agrawal2006}. Additionally, finding better evaluation techniques to compare different
 automatic physical design decision is currently being studied \cite{gebaly2008}.
@@
 Just like all other research areas are moving to the cloud, database systems are moving there too. Accordingly, lots of current research focuses on adapting
-storage techniques and index structures to the cloud. For example, Wang et al. propose a new index structure for indexing multi-dimensional data in the
+storage techniques and index structures to the cloud. For example, Wang et al. propose a new index structure for indexing multidimensional data in the
 cloud \cite{wang2010indexing}. Lots of work is currently being done in this area. For example, Wu et al. \cite{wu2010efficient} propose a modified B+tree index
@@
 other information systems. We focused on three types of information systems: Main Memory Database Systems (MMDB), XML Databases, and Data Warehouses. We
-presented the challenges involved with each type of system, and how each of the four physical design elements were adapted or utilized to overcome these
+presented the challenges involved with each system, and how each of the four physical design elements were adapted or utilized to overcome these
 challenges. Table \ref{tab:summary} summarizes the results of our survey. From our literature review, we found that the B+tree is the most utilized index
 structure, and has been adapted and customized to fit the different needs of other systems. For example, it was made more cache conscious to fit MMDB needs,
-and was changed to apply to multidimensions to fit the Data Warehouses needs. Materialized views were found to be heavily utlitized in all the systems we
+and was changed to apply to multidimensions to fit the data warehouses' needs. Materialized views were found to be heavily utlitized in all the systems we
 presented, except in MMDBs. Partitioning was also used in all systems, but was only used for data reloading in MMDBs. Horizontal partitioning is much more
-common that vertical partitioning. Clustering is very important in all systems, except for MMDBs. Clustering is especially very important for Data Warehouses
+common that vertical partitioning. Clustering is very important in all systems, except for MMDBs. Clustering is especially important for data warehouses
 where huge amounts of data are stored.
@@
 \textbf{XML DB} &
-\raggedright -Need a numbering scheme for nodes. \linebreak \linebreak
--E.g. B+tree and variations used. & XQuery and XPath results are materialized & Horizontal partitioning
+\raggedright -Need a numbering scheme for nodes \linebreak \linebreak
+-E.g. B+tree and variations used & XQuery and XPath results are materialized & Horizontal partitioning
 based on inlining or node type & Based on document order or structural similarity\\
@@
 clustered by
-nature.\\
+nature\\
 \end{tabular}
